### 2019-08-04 draft.Rmd
Initial commit

---
title: "Formal approaches to survivial child 'experienced' by mothers"
author: "DA-IW"
date: "August 2019"
output: rmarkdown::html_vignette
bibliography: bib.bibtex
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>"
)
```

## Abstract

In kinship theory, the survival (or deaths) descendents has a several impact in life reproductive of mothers. The formal demographic tools developed before allows to characterize the mortality from the mother perspective, comparing populations in different demographic stages. For that, some additional expressions are proposed, and also an approach to how mortality changes can affect the survival child experience.

## Relationship

Consider a stable female population with fertility rates $m_{x}$ and force of mortality $\mu_{x}$, where the expected number of surviving children to a woman aged \textit{a},

\begin{equation}
\label{eq:CS}
CS_a = \int_{\alpha}^a{m_{x}l_{a-x}dx}
\end{equation}

This was developed in @goodman1974, in the "counting method" approach, being the net reproduction rate $R_0$, or the ratio between generations (multiplying and dividing by the mother generation $B$).

can be affected by a change in of $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk  (@goodman1974). Following Wrycza & Baudisch (@wrycza2012), this change can be generalized as a function of $\delta$, being an absolute change  $m_{x,\delta}=m_{x}+\delta$ and a relative one $m_{x,\delta}=m_{x}(1+\delta)$. In the first case, a discrete approximate effect is:

\begin{equation}
\label{eq:abs_change}
\frac{\Delta CS_{a}}{CS_{a}} \approx -(a - \overline{x}_{(CS)}) \Delta\delta.
\end{equation}

Being proportional and inverse to the difference between maternal age \textit{a} and the mean age at birth for those living daughters $\overline{x}_{(CS)}$. In other words, the magnitude of the change depends negatively of how young is the successful reproductive pattern by age, as newborns will have larger risk periods.

In the second case, the ralationship is not so clear because of the interaction of birth and mortality rates, but some things can be said (@KeyfitzMath):

\begin{equation}
\label{eq:relat_change}
\frac{\Delta CS_{a}}{CS_{a}} \approx -H_{CS_a} \Delta\delta.
\end{equation}

Where $H_{CS_a}$ can be interpreted as a temporary entropy constant until age $a-\alpha$ but weighted by the fertility pattern between. In this sense, depending the way in that the survival childs are accumulated by age $C_{CS_a}(x)=\frac{m_{x}l_{a-x}}{CS_{a}}$ is the way that is taken each age contribution to the disparity measure $H_{CS_a}=\int_\alpha^a{C_{CS_a}(x)H(x)}$.
The effect in the child survival experienced by the mother will be bigger the mortality is the same in all ages, no matter when are concentrated the newborns.

**here comes a nice graph to explain this**

Some additional measures are proposed to measure the survival experience by a mother aged $a$. One is an absolute measure of the *Mean Time Spent with Lost*, which can be expressed in terms of a temporary expected lost years measure $e^\dagger$:

\begin{equation}
\label{eq:MTSL}
MTSL_a = \int_\alpha^a{m_x\int_0^{a-x}{d_t e_{t,a-x}dt}dx} = \int_\alpha^a{m_x e_{o,a-x}^\dagger dx}.
\end{equation}

But maybe a mother would relate this lost years with the expected ones, feeling the *intensity* of her losts (*Intensity Time Lost*), and make possible the comparison between fertility regimes:

\begin{equation}
\label{eq:ITL}
  ITL_a = \frac{\int_\alpha^a{m_x e_{o,a-x}^\dagger dx}}{\int_\alpha^a{m_x e_{o,a-x} dx}}
\end{equation}

Finally, different populations could have different mean age at lost of mothers for those death childs:

\begin{equation}
\label{eq:MAL}
\begin{split}  
  MAL_a = \frac{\int_\alpha^a{m_x {\int_0^{a-x}\frac{d_t (x+t) dt}{\int_0^{a-x}{d_tdt}}}}}{\int_\alpha^a{m_x}} \\

  MAL_a = \frac{\int_\alpha^a{m_x x}}{TFR_a} + \frac{\int_\alpha^a{m_x MAD_{a-x}}}{TFR_a} \\

  MAL_a = K + \frac{\int_\alpha^a{m_x MAD_{a-x}}}{TFR_a} \\

  MAL_a = K + \int_\alpha^a{f_{x,a} MAD_{a-x}}
\end{split}
\end{equation}

$K$ is the mean age at birth, $MAD_{a-x}$ refers to the mean age at death for those newborns who dies before $a-x$, $TFR_a$ is the cumulated fertility for a women aged a, and $f_{x,a}$ is the cumulative distribution of fertility until age $a$.

## Proofs

Here we show the proofs for the effects of changing mortality:

### Absolute change

Considering that $m_{x,\delta}=m_{x}+\delta$, $l_{a-x} = e^{-\int_{0}^{a-x}{(\mu_t+\delta})}$, and:

\begin{equation}
CS_{(a)}^\delta = \int_{\alpha}^{a}m_{x} l_{a-x} e^{-\delta (a-x)} dx.  
\end{equation}

To find the effects on daughter survival of adding $\delta$ to the age-specific death rates of daughters, we get the derivative of $dCS_{(a)}^\delta / d\delta$, evaluated near zero (Keyfitz \& Caswell, 2005, section 4.3):

\begin{equation}
\label{eq2}
= -a\int_{\alpha}^{a}m_{x} l_{a-x} dx + \int_{\alpha}^{a}m_{x} l_{a-x}x dx.
\end{equation}

Since $CS_{a} = \int_{\alpha}^{a}m_{x} l_{a-x} dx$, we can rewrite equation (\ref{eq2}) as

\begin{equation}
\frac{dCS^\delta}{d\delta} = -a CS_{a}  + \int_{\alpha}^{a} x m_{x} l_{a-x} dx.
\end{equation}

Dividing both sides by $CS_{a}$ and multiplying by $d\delta$, we get

\begin{equation}
\label{mort_cont}
\frac{d CS_{a}}{CS_{a}} = -(a - \overline{x}_{(CS)}) d\delta
\end{equation}

where $\overline{x}_{(CS)}$ is the mean age of women at the birth of their surviving daughters in a stationary population.

### Relative change

\begin{equation}
\label{mort_change_relative}
\frac{d CS_{(a)}}{CS_{(a)}} = \frac{ \int_{\alpha}^{a}m_{(x)} l_{(a-x)} \log(l_{(a-x)}) dx}{\int_{\alpha}^{a}m_{(x)} l_{(a-x)}} d\delta
\end{equation}

## History

Lotka (@lotka1931orphanhood) started to model the orphanhood in a theorical population, comparing the effect in different regimes.  The expected number of living daughters as a stable population concept was mentioned in Keyfitz & Pullum (@goodman1974), extending this idea to the rest of the kinship, in various genealogical directions. The formal effect of different kind of changes in mortality by age was well described in @wrycza2012.

## Applications

The proposed indexes will be applied to two very different regimes.
The changes of mortality will be done compationallyto evaluate which component is more important in

## Bibliography

### 2019-08-05 draft.Rmd
Extend the relative change proof

@@
 
-Consider a stable female population with fertility rates $m_{x}$ and force of mortality $\mu_{x}$, where the expected number of surviving children to a woman aged \textit{a},
+Consider a stable female population with fertility rates $m_{x}$ and force of mortality $\mu_{x}$, where the expected number of surviving children to a woman aged $a$,
 
@@
 
-This was developed in @goodman1974, in the "counting method" approach, being the net reproduction rate $R_0$, or the ratio between generations (multiplying and dividing by the mother generation $B$).
+This was developed in @goodman1974, in the "counting method" approach, being the net reproduction rate $R_0$ limited to $a$, or the (not complete) ratio between generations (multiplying and dividing by the mother generation $B$).
 
-can be affected by a change in of $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk  (@goodman1974).
-Following Wrycza & Baudisch (@wrycza2012), this change can be generalized as a function of $\delta$, being an absolute change  $m_{x,\delta}=m_{x}+\delta$ and a relative one $m_{x,\delta}=m_{x}(1+\delta)$.
+This can be affected by a change of $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk  (@goodman1974).
+Following @wrycza2012, this change can be generalized as a function of $\delta$, being an absolute change  $m_{x,\delta}=m_{x}+\delta$ and a relative one $m_{x,\delta}=m_{x}(1+\delta)$.
 In the first case, a discrete approximate effect is:
@@
 In this sense, depending the way in that the survival childs are accumulated by age $C_{CS_a}(x)=\frac{m_{x}l_{a-x}}{CS_{a}}$ is the way that is taken each age contribution to the disparity measure $H_{CS_a}=\int_\alpha^a{C_{CS_a}(x)H(x)}$.
-The effect in the child survival experienced by the mother will be bigger the mortality is the same in all ages, no matter when are concentrated the newborns.
+The effect in the child survival experienced by the mother will be bigger when the mortality is the same in all ages, no matter where are concentrated the newborns.
 
-**here comes a nice graph to explain this**
+```{r, echo=F}
+"graph"
+```
 
@@
 
-Finally, different populations could have different mean age at lost of mothers for those death childs:
+Finally, different populations could have different mean age at lost of mothers with death childs:
 
@@
 \label{eq:MAL}
-\begin{split}  
-  MAL_a = \frac{\int_\alpha^a{m_x {\int_0^{a-x}\frac{d_t (x+t) dt}{\int_0^{a-x}{d_tdt}}}}}{\int_\alpha^a{m_x}} \\
-
-  MAL_a = \frac{\int_\alpha^a{m_x x}}{TFR_a} + \frac{\int_\alpha^a{m_x MAD_{a-x}}}{TFR_a} \\
-
-  MAL_a = K + \frac{\int_\alpha^a{m_x MAD_{a-x}}}{TFR_a} \\
-
-  MAL_a = K + \int_\alpha^a{f_{x,a} MAD_{a-x}}
-\end{split}
+  MAL_a = K_a + \int_\alpha^a{f_{x,a} MAD_{a-x}}
 \end{equation}
 
-$K$ is the mean age at birth, $MAD_{a-x}$ refers to the mean age at death for those newborns who dies before $a-x$, $TFR_a$ is the cumulated fertility for a women aged a, and $f_{x,a}$ is the cumulative distribution of fertility until age $a$.
+$K_a$ is the mean age at birth for womens aged $a$, $MAD_{a-x}$ refers to the mean age at death for those newborns who dies before $a-x$, $TFR_a$ is the cumulated fertility for a women aged a, and $f_{x,a}$ is the cumulative distribution of fertility until age $a$.
 
 ## Proofs
-
-Here we show the proofs for the effects of changing mortality:
 
@@
 
+Considering that $m_{x,\delta}=m_{x}(1+\delta)$, $l_{a-x,\delta} = e^{-\int_{0}^{a-x}{\mu_t(1+\delta})dt}=l_{a-x}^{(1+\delta)}$, and:
+
+\begin{equation}
+CS_{a}^\delta = \int_{\alpha}^{a} {m_{x} l_{a-x}^{(1+\delta)}} dx.  
+\end{equation}
+
+Using that the derivative of $dCS_{(a)}^\delta / d\delta = log(l_{a-x}) l_{a-x}^{(1+\delta)}$, evaluating $\delta$ near zero we get the final expression:
+
 \begin{equation}
 \label{mort_change_relative}
-\frac{d CS_{(a)}}{CS_{(a)}} = \frac{ \int_{\alpha}^{a}m_{(x)} l_{(a-x)} \log(l_{(a-x)}) dx}{\int_{\alpha}^{a}m_{(x)} l_{(a-x)}} d\delta
+\frac{d CS_{a}}{CS_{a}} = \frac{ \int_{\alpha}^{a}m_{x} l_{a-x} \log(l_{a-x}) dx}{\int_{\alpha}^{a}m_{x} l_{a-x}} d\delta
+\end{equation}
+
+Using that $C_{CS_a}(x)=\frac{m_{x}l_{a-x}}{CS_{a}}$ is a weighting function and making an analogy with the classical entropy fucntion $H$ where the weights are just the survival function, we can define this inforamtion function $H_{CS_a}=\int_\alpha^a{C_{CS_a}H(x)}$, and get:
+
+\begin{equation}
+\label{eq:relat_change2}
+\frac{dCS_{a}}{CS_{a}} = -H_{CS_a} d\delta.
+\end{equation}
+
+### Mean Age at Lost
+
+Start with the integral of the age $x+t$ at each death child weighted by fertility and survival function
+
+\begin{equation}
+\label{eq:MALproof}
+\begin{split}  
+  MAL_a = \frac{\int_\alpha^a{m_x {\int_0^{a-x}\frac{d_t (x+t) dt}{\int_0^{a-x}{d_tdt}}}}}{\int_\alpha^a{m_x}} \\
+
+  MAL_a = \frac{\int_\alpha^a{m_x x}}{TFR_a} + \frac{\int_\alpha^a{m_x MAD_{a-x}}}{TFR_a} \\
+
+  MAL_a = K_a + \frac{\int_\alpha^a{m_x MAD_{a-x}}}{TFR_a} \\
+
+  MAL_a = K_a + \int_\alpha^a{f_{x,a} MAD_{a-x}}
+\end{split}
 \end{equation}
@@
 
-Lotka (@lotka1931orphanhood) started to model the orphanhood in a theorical population, comparing the effect in different regimes.
-The expected number of living daughters as a stable population concept was mentioned in Keyfitz & Pullum (@goodman1974), extending this idea to the rest of the kinship, in various genealogical directions.
+@lotka1931orphanhood started to model the orphanhood in a theorical population, comparing the effect in different regimes.
+The expected number of living daughters as a stable population concept was mentioned in @goodman1974, extending this idea to the rest of the kinship, in various genealogical directions.
 The formal effect of different kind of changes in mortality by age was well described in @wrycza2012.
@@
 The proposed indexes will be applied to two very different regimes.
-The changes of mortality will be done compationallyto evaluate which component is more important in
+The changes of mortality will be done computaionally to evaluate which component is more important in the raltive case
 

### 2019-08-07 draft.Rmd
strucutred abstract and language changes

@@
 ---
-title: "Formal approaches to survivial child 'experienced' by mothers"
+title: "A formal approach to child survival from the perspective of mothers"
 author: "DA-IW"
 date: "August 2019"
-output: rmarkdown::html_vignette
+output: pdf_document
 bibliography: bib.bibtex
@@
 ```{r setup, include = FALSE}
-knitr::opts_chunk$set(collapse = TRUE, comment = "#>"
-)
+knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
 ```
@@
 
-In kinship theory, the survival (or deaths) descendents has a several impact in life reproductive of mothers.
-The formal demographic tools developed before allows to characterize the mortality from the mother perspective, comparing populations in different demographic stages.
-For that, some additional expressions are proposed, and also an approach to how mortality changes can affect the survival child experience.
+### Background
+
+An important relation in kinship theory is the expected descendants for mothers at certain age.
+Child survival and child death affects important dimensions of a mothers' life, such as their reproduction and quality of life.
+
+### Objective
+
+Existing formal demographic tools are used within a stable population framework to characterize child mortality from the perspective of mothers considering both an absolute and a relative change in the force of mortality.
+
+### Results
+
+The expected change in child survival given an absolute change in mortality depends on the difference between maternal age and the age distribution o her surviving children.
+For absolute changes in mortality, it is necessary to consider a measure of entropy weighted by fertility.
+
+### Contribution
+
+The paper proposes related expressions to describe how mortality changes affects child survival for mothers across demographic regimes.
+The paper also proposes alternative ways of conceptualizing child survival.
 
@@
 
-Consider a stable female population with fertility rates $m_{x}$ and force of mortality $\mu_{x}$, where the expected number of surviving children to a woman aged $a$,
+$CS_a$ is the expected number of surviving children to a woman aged $a$ in a stable female population with fertility rates $m_{x}$ and force of mortality $\mu_{x}$:
 
@@
 
-This was developed in @goodman1974, in the "counting method" approach, being the net reproduction rate $R_0$ limited to $a$, or the (not complete) ratio between generations (multiplying and dividing by the mother generation $B$).
+We now consider the consequences of a change $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk (@goodman1974) and $\delta$ can be absolute or relative (@wrycza2012).
 
-This can be affected by a change of $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk  (@goodman1974).
-Following @wrycza2012, this change can be generalized as a function of $\delta$, being an absolute change  $m_{x,\delta}=m_{x}+\delta$ and a relative one $m_{x,\delta}=m_{x}(1+\delta)$.
-In the first case, a discrete approximate effect is:
+A discrete approximate of the effect on $CS_a$ of an absolute change in mortality $m_{x,\delta}=m_{x}+\delta$ is:
 
@@
 
-Being proportional and inverse to the difference between maternal age \textit{a} and the mean age at birth for those living daughters $\overline{x}_{(CS)}$.
-In other words, the magnitude of the change depends negatively of how young is the successful reproductive pattern by age, as newborns will have larger risk periods.
+In equation \ref{eq:abs_change}, the expected change in $CS_a$ is inversely proportional to the difference between maternal age \textit{a} and the mean age of the mother at the birth of her surviving daughters $\overline{x}_{(CS)}$.
+In other words, the magnitude of the change depends negatively on the age distribution of the surviving offspring (younger children experience longer periods of exposure to risk).
 
-In the second case, the ralationship is not so clear because of the interaction of birth and mortality rates, but some things can be said (@KeyfitzMath):
+<!-- No entiendo la oracion de abajo. Le reescribi en el parrafo de arriba, pero no se si esto es lo que querias decir? -->
+
+<!-- In other words, the magnitude of the change depends negatively on how young is the successful reproductive pattern by age, as newborns will have larger risk periods. -->
+
+The relationship is less clear for proportional changes in mortality $m_{x,\delta}=m_{x}(1+\delta)$ given the interaction of birth and mortality rates (@KeyfitzMath).
+Neverthless we can say that, in discrete terms:
 
@@
   MAL_a = \frac{\int_\alpha^a{m_x {\int_0^{a-x}\frac{d_t (x+t) dt}{\int_0^{a-x}{d_tdt}}}}}{\int_\alpha^a{m_x}} \\
-
   MAL_a = \frac{\int_\alpha^a{m_x x}}{TFR_a} + \frac{\int_\alpha^a{m_x MAD_{a-x}}}{TFR_a} \\
-
   MAL_a = K_a + \frac{\int_\alpha^a{m_x MAD_{a-x}}}{TFR_a} \\
-
   MAL_a = K_a + \int_\alpha^a{f_{x,a} MAD_{a-x}}
@@
 ## History
+
+Equation  \ref{eq:CS} was developed in @goodman1974, in the "counting method" approach, being the net reproduction rate $R_0$ limited to $a$, or the (not complete) ratio between generations (multiplying and dividing by the mother generation $B$).
 

### 2019-08-07 draft.Rmd
first draft of dag revisions 1

@@
 title: "A formal approach to child survival from the perspective of mothers"
-author: "DA-IW"
+author: "Ivan Williams and Diego Alburez-Gutierrez"
 date: "August 2019"
@@
 
-An important relation in kinship theory is the expected descendants for mothers at certain age.
-Child survival and child death affects important dimensions of a mothers' life, such as their reproduction and quality of life.
+The expected number of descendants for mothers that reach a certain age is an important relation in kinship theory.
+Child survival (and child death) affect mothers in a variety of ways, yet it is not clear how changes in mortality affect the number of surviving children for mothers.
 
@@
 
-Existing formal demographic tools are used within a stable population framework to characterize child mortality from the perspective of mothers considering both an absolute and a relative change in the force of mortality.
+Existing formal demographic tools are used within a stable population framework to characterize child survival from the point of view of a mother, considering both absolute and relative changes in mortality patterns.
 
@@
 
-The expected change in child survival given an absolute change in mortality depends on the difference between maternal age and the age distribution o her surviving children.
-For absolute changes in mortality, it is necessary to consider a measure of entropy weighted by fertility.
+The expected change in child survival given an absolute change in mortality depends on the difference between maternal age and the age structure of her surviving children.
+The effects of absolute changes in mortality are explained using a fertility-weighted measure of entropy.
 
@@
 
-The paper proposes related expressions to describe how mortality changes affects child survival for mothers across demographic regimes.
-The paper also proposes alternative ways of conceptualizing child survival.
+The paper proposes new expresions to characterize the effects of changes in mortality on child survival.
+The different meassures it proposes can be applied to populations at different stages of the demographic transition.
 
-## Relationship
+## 1. Relationship
 
@@
 \label{eq:CS}
-CS_a = \int_{\alpha}^a{m_{x}l_{a-x}dx}
+CS_a = \int_{\alpha}^a{m_{x}l_{a-x}dx}.
 \end{equation}
@@
 
-In equation \ref{eq:abs_change}, the expected change in $CS_a$ is inversely proportional to the difference between maternal age \textit{a} and the mean age of the mother at the birth of her surviving daughters $\overline{x}_{(CS)}$.
-In other words, the magnitude of the change depends negatively on the age distribution of the surviving offspring (younger children experience longer periods of exposure to risk).
+In this equation, the expected change in child survival is inversely proportional to the difference between maternal age \textit{a} and the mean age of the mother at the birth of her surviving daughters $\overline{x}_{(CS)}$.
+In other words, the magnitude of the change depends negatively on the age distribution of the surviving offspring (as younger children experience longer periods of exposure to risk).
 
@@
 
+<!-- Aqui no cambie nada porque no entiendo muy bien lo que hiciste. Talvez hablamos para discutir? -->
+
 Where $H_{CS_a}$ can be interpreted as a temporary entropy constant until age $a-\alpha$ but weighted by the fertility pattern between.
@@
 
-Some additional measures are proposed to measure the survival experience by a mother aged $a$.
-One is an absolute measure of the *Mean Time Spent with Lost*, which can be expressed in terms of a temporary expected lost years measure $e^\dagger$:
+## 2. Proofs
 
-\begin{equation}
-\label{eq:MTSL}
-MTSL_a = \int_\alpha^a{m_x\int_0^{a-x}{d_t e_{t,a-x}dt}dx} = \int_\alpha^a{m_x e_{o,a-x}^\dagger dx}.
-\end{equation}
+### 2.1. Absolute change
 
-But maybe a mother would relate this lost years with the expected ones, feeling the *intensity* of her losts (*Intensity Time Lost*), and make possible the comparison between fertility regimes:
-
-\begin{equation}
-\label{eq:ITL}
-  ITL_a = \frac{\int_\alpha^a{m_x e_{o,a-x}^\dagger dx}}{\int_\alpha^a{m_x e_{o,a-x} dx}}
-\end{equation}
-
-Finally, different populations could have different mean age at lost of mothers with death childs:
-
-\begin{equation}
-\label{eq:MAL}
-  MAL_a = K_a + \int_\alpha^a{f_{x,a} MAD_{a-x}}
-\end{equation}
-
-$K_a$ is the mean age at birth for womens aged $a$, $MAD_{a-x}$ refers to the mean age at death for those newborns who dies before $a-x$, $TFR_a$ is the cumulated fertility for a women aged a, and $f_{x,a}$ is the cumulative distribution of fertility until age $a$.
-
-## Proofs
-
-### Absolute change
-
-Considering that $m_{x,\delta}=m_{x}+\delta$, $l_{a-x} = e^{-\int_{0}^{a-x}{(\mu_t+\delta})}$, and:
+Considering that $m_{x,\delta}=m_{x}+\delta$ and $l_{a-x} = e^{-\int_{0}^{a-x}{(\mu_t+\delta})}$:
 
@@
 
-### Relative change
+### 2.2. Relative change
 
@@
 
-Using that the derivative of $dCS_{(a)}^\delta / d\delta = log(l_{a-x}) l_{a-x}^{(1+\delta)}$, evaluating $\delta$ near zero we get the final expression:
+Using the derivative of $dCS_{(a)}^\delta / d\delta = log(l_{a-x}) l_{a-x}^{(1+\delta)}$ and evaluating $\delta$ near zero we get the final expression:
 
@@
 \end{equation}
+
+<!-- I think this needs re-weriting for clarity but I didn't want to change it to avoid messing it up! -->
+
+<!-- Podrias dar alguna referencia para lo de entropia?  -->
 
@@
 
-### Mean Age at Lost
+## 3. History
 
-Start with the integral of the age $x+t$ at each death child weighted by fertility and survival function
+Kinship theory has a long pedigree in mathematical demography.
+@lotka1931orphanhood started to model the orphanhood in a theorical population, comparing the effect in different demographic regimes.
+Equation  \ref{eq:CS} was originally proposed by @brass_derivation_1953 to estimate child mortality.
+The notion of estimating the expected number of living daughters in a stable population was extended by @goodman1974 to other kin relations (granddaughters, cousins, etc.).
+The "counting method" approach, being the net reproduction rate $R_0$ limited to $a$, or the (not complete) ratio between generations (multiplying and dividing by the mother generation $B$), was further popularized in @KeyfitzMath.
+@bongaarts_projection_1987 used a similar approach to estimate descendants in his 'Family Status Model'.
+More recently, @wrycza2012 has looked at the formal effect of different kind of changes in mortality by age in a paper published in this series.
+
+## 4. Related meassures
+
+### 4.1. Time with children lost
+
+Some additional measures are proposed to measure the survival experience by a mother aged $a$.
+One is an absolute measure of the *Mean Time Spent with Lost*, which can be expressed in terms of a temporary expected lost years measure $e^\dagger$:
+
+\begin{equation}
+\label{eq:MTSL}
+MTSL_a = \int_\alpha^a{m_x\int_0^{a-x}{d_t e_{t,a-x}dt}dx} = \int_\alpha^a{m_x e_{o,a-x}^\dagger dx}.
+\end{equation}
+
+But maybe a mother would relate these lost years with the expected ones, feeling the *intensity* of her losts (*Intensity Time Lost*), and make possible the comparison between fertility regimes:
+
+\begin{equation}
+\label{eq:ITL}
+  ITL_a = \frac{\int_\alpha^a{m_x e_{o,a-x}^\dagger dx}}{\int_\alpha^a{m_x e_{o,a-x} dx}}
+\end{equation}
+
+### 4.2. Mean Age at Child Loss
+
+Finally, different populations could have different mean age at lost of mothers with death childs:
+
+\begin{equation}
+\label{eq:MAL}
+  MAL_a = K_a + \int_\alpha^a{f_{x,a} MAD_{a-x}}
+\end{equation}
+
+$K_a$ is the mean age at birth for womens aged $a$, $MAD_{a-x}$ refers to the mean age at death for those newborns who dies before $a-x$, $TFR_a$ is the cumulated fertility for a women aged a, and $f_{x,a}$ is the cumulative distribution of fertility until age $a$.
+
+Equation \ref{eq:MAL} can be derived by starting with the integral of the age $x+t$ at each death child weighted by fertility and survival function
 
@@
 
-## History
+## 5. Applications
 
-Equation  \ref{eq:CS} was developed in @goodman1974, in the "counting method" approach, being the net reproduction rate $R_0$ limited to $a$, or the (not complete) ratio between generations (multiplying and dividing by the mother generation $B$).
+The proposed meassures are applied to two different demographic regimes.
 
-@lotka1931orphanhood started to model the orphanhood in a theorical population, comparing the effect in different regimes.
-The expected number of living daughters as a stable population concept was mentioned in @goodman1974, extending this idea to the rest of the kinship, in various genealogical directions.
-The formal effect of different kind of changes in mortality by age was well described in @wrycza2012.
+[PENDING]
 
-## Applications
+A computation approach is used to evaluate the relative contribution of each component given a relative change in $\delta$.
 
-The proposed indexes will be applied to two very different regimes.
-The changes of mortality will be done computaionally to evaluate which component is more important in the raltive case
+[PENDING]
 
-## Bibliography
+## 5. Bibliography

### 2019-08-12 draft.Rmd

@@
 ---
-title: "A formal approach to child survival from the perspective of mothers"
+title: "Changes in mortality and child survival for mothers"
 author: "Ivan Williams and Diego Alburez-Gutierrez"
-date: "August 2019"
+date: "07 August 2019"
 output: pdf_document
@@
 
-The expected number of descendants for mothers that reach a certain age is an important relation in kinship theory.
-Child survival (and child death) affect mothers in a variety of ways, yet it is not clear how changes in mortality affect the number of surviving children for mothers.
+The expected number of descendants for mothers is an important relation in kinship theory.
+Child survival (and child death) affect mothers in a variety of ways, yet it is not clear how changes in mortality schedules affect the number of surviving children from the standpoint of mothers.
 
@@
 
-The expected change in child survival given an absolute change in mortality depends on the difference between maternal age and the age structure of her surviving children.
+The expected change in child survival given an absolute change in mortality depends on the difference between the age of a mother and the age structure of her surviving children.
 The effects of absolute changes in mortality are explained using a fertility-weighted measure of entropy.
@@
 The paper proposes new expresions to characterize the effects of changes in mortality on child survival.
-The different meassures it proposes can be applied to populations at different stages of the demographic transition.
+The different meassures it introduces can be applied to populations at different stages of the demographic transition.
 
@@
 
-We now consider the consequences of a change $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk (@goodman1974) and $\delta$ can be absolute or relative (@wrycza2012).
+We now consider the consequences of a change $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk, as proposed by @goodman1974, and $\delta$ can be absolute or relative.
 
-A discrete approximate of the effect on $CS_a$ of an absolute change in mortality $m_{x,\delta}=m_{x}+\delta$ is:
+A discrete approximate of the effect on $CS_a$ given an absolute change in mortality $m_{x,\delta}=m_{x}+\delta$ is:
 
@@
 
-Where $H_{CS_a}$ can be interpreted as a temporary entropy constant until age $a-\alpha$ but weighted by the fertility pattern between.
+Where $H_{CS_a}$ can be interpreted as a temporary entropy constant until age $a-\alpha$ but weighted by the fertility pattern between [falta algo aqui?].
 In this sense, depending the way in that the survival childs are accumulated by age $C_{CS_a}(x)=\frac{m_{x}l_{a-x}}{CS_{a}}$ is the way that is taken each age contribution to the disparity measure $H_{CS_a}=\int_\alpha^a{C_{CS_a}(x)H(x)}$.
-The effect in the child survival experienced by the mother will be bigger when the mortality is the same in all ages, no matter where are concentrated the newborns.
+The effect in the child survival experienced by the mother will be bigger when the mortality is the same in all ages, no matter where the newborns are concentrated.
 
@@
 
-Considering that $m_{x,\delta}=m_{x}(1+\delta)$, $l_{a-x,\delta} = e^{-\int_{0}^{a-x}{\mu_t(1+\delta})dt}=l_{a-x}^{(1+\delta)}$, and:
+Considering that $m_{x,\delta}=m_{x}(1+\delta)$ and $l_{a-x,\delta} = e^{-\int_{0}^{a-x}{\mu_t(1+\delta})dt}=l_{a-x}^{(1+\delta)}$:
 
@@
 
-Using that $C_{CS_a}(x)=\frac{m_{x}l_{a-x}}{CS_{a}}$ is a weighting function and making an analogy with the classical entropy fucntion $H$ where the weights are just the survival function, we can define this inforamtion function $H_{CS_a}=\int_\alpha^a{C_{CS_a}H(x)}$, and get:
+Using that $C_{CS_a}(x)=\frac{m_{x}l_{a-x}}{CS_{a}}$ is a weighting function and making an analogy with the classical entropy function $H$ where the weights are just the survival function, we can define this information function $H_{CS_a}=\int_\alpha^a{C_{CS_a}H(x)}$, and get:
 
@@
 
-Kinship theory has a long pedigree in mathematical demography.
-@lotka1931orphanhood started to model the orphanhood in a theorical population, comparing the effect in different demographic regimes.
+Kin count estimation has a long pedigree in mathematical demography.
+@lotka1931orphanhood proposed to model the orphanhood in a theorical population, comparing the effect across demographic regimes.
 Equation  \ref{eq:CS} was originally proposed by @brass_derivation_1953 to estimate child mortality.
-The notion of estimating the expected number of living daughters in a stable population was extended by @goodman1974 to other kin relations (granddaughters, cousins, etc.).
-The "counting method" approach, being the net reproduction rate $R_0$ limited to $a$, or the (not complete) ratio between generations (multiplying and dividing by the mother generation $B$), was further popularized in @KeyfitzMath.
+The notion of estimating the expected number of living daughters in a stable population was generalized by @goodman1974 for other kin relations (granddaughters, cousins, etc.).
+The "counting method" approach, being the net reproduction rate $R_0$ limited to $a$, or the (not complete) ratio between generations (multiplying and dividing by the mother generation $B$), was further popularized by @KeyfitzMath.
 @bongaarts_projection_1987 used a similar approach to estimate descendants in his 'Family Status Model'.
@@
 
-Some additional measures are proposed to measure the survival experience by a mother aged $a$.
-One is an absolute measure of the *Mean Time Spent with Lost*, which can be expressed in terms of a temporary expected lost years measure $e^\dagger$:
+<!-- Siento que esta parte necesita un poco mas de explicacion -->
+
+This section proposes two additional measures of child survival for a mother aged $a$.
+$MTSL$ is an absolute measure of the *Mean Time Spent with Lost*, which can be expressed in terms of a temporary expected lost years measure $e^\dagger$:
 
@@
 
-But maybe a mother would relate these lost years with the expected ones, feeling the *intensity* of her losts (*Intensity Time Lost*), and make possible the comparison between fertility regimes:
+It is reasonable to expect a mother to relate these lost years with the expected ones, feeling the *intensity* of her losts (*Intensity Time Lost*).
+This makes it possible to compare between fertility regimes:
 
@@
 
-Finally, different populations could have different mean age at lost of mothers with death childs:
+Different populations could have different mean age at lost of mothers with death child:
 
@@
 
-Equation \ref{eq:MAL} can be derived by starting with the integral of the age $x+t$ at each death child weighted by fertility and survival function
+Equation \ref{eq:MAL} can be derived by starting with the integral of the age $x+t$ at each death child weighted by fertility and survival function:
 

### 2019-08-12 draft.Rmd
Aclarando cambio relativo. Aun por finalizar

@@
 date: "07 August 2019"
-output: pdf_document
+output: html_document
 bibliography: bib.bibtex
@@
 The expected number of descendants for mothers is an important relation in kinship theory.
-Child survival (and child death) affect mothers in a variety of ways, yet it is not clear how changes in mortality schedules affect the number of surviving children from the standpoint of mothers.
+Child survival (and child death) affect mothers in a variety of ways, yet it is not clear how changes in mortality schedules, given a fertility pattern, affect the number of surviving children from the standpoint of mothers.
 
@@
 The expected change in child survival given an absolute change in mortality depends on the difference between the age of a mother and the age structure of her surviving children.
-The effects of absolute changes in mortality are explained using a fertility-weighted measure of entropy.
+The effects of relative changes in mortality are explained using a fertility-weighted measure of entropy.
 
@@
 
-$CS_a$ is the expected number of surviving children to a woman aged $a$ in a stable female population with fertility rates $m_{x}$ and force of mortality $\mu_{x}$:
+$CS_a$ is the expected number of surviving children to a woman aged $a$ in a stable female population with fertility rates $m_{x}$ and force of mortality $\mu_{x}$, as proposed by @goodman1974:
 
@@
 
-We now consider the consequences of a change $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk, as proposed by @goodman1974, and $\delta$ can be absolute or relative.
+We now consider the consequences of a change $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk.
 
@@
 
-<!-- No entiendo la oracion de abajo. Le reescribi en el parrafo de arriba, pero no se si esto es lo que querias decir? -->
-
-<!-- In other words, the magnitude of the change depends negatively on how young is the successful reproductive pattern by age, as newborns will have larger risk periods. -->
-
 The relationship is less clear for proportional changes in mortality $m_{x,\delta}=m_{x}(1+\delta)$ given the interaction of birth and mortality rates (@KeyfitzMath).
@@
 
-Where $H_{CS_a}$ can be interpreted as a temporary entropy constant until age $a-\alpha$ but weighted by the fertility pattern between [falta algo aqui?].
-In this sense, depending the way in that the survival childs are accumulated by age $C_{CS_a}(x)=\frac{m_{x}l_{a-x}}{CS_{a}}$ is the way that is taken each age contribution to the disparity measure $H_{CS_a}=\int_\alpha^a{C_{CS_a}(x)H(x)}$.
-The effect in the child survival experienced by the mother will be bigger when the mortality is the same in all ages, no matter where the newborns are concentrated.
+Where $H_{CS_a}$ can be interpreted as a temporary entropy constant until age $a-\alpha$ . 
+Depending the age structure of the survival daughters $C_{CS_a}(x)=\frac{m_{x}l_{a-x}}{CS_{a}}$,  is the way that is taken each age contribution to the disparity measure $H_{CS_a}=\int_\alpha^a{C_{CS_a}(x)H(x)}$ (@Aburto2019).
 
-```{r, echo=F}
-"graph"
+The effect of a relative change $\delta$ will be bigger when the mortality is more disperse in ages where the fertility is very concentrated.
+For example in \ref{fig:relative_change}, the youngest schedule would have a bigger change in $C_{CS_a}$...
+
+```{r relative_change, echo=F, warning=F, fig.cap="Two different fertility schedules (black lines) and a big hump mortality case (right axis, line in red).
+In dashed grey lines are the mean age at birth"}
+
+# genero m_x (old and young pattern) y l_x (with big hump) 
+f_young = data.frame(x=c(15,20,25,30,35,40), f=c(0,.3,.3,.05,.01,.01))
+f_old = data.frame(  x=c(15,20,25,30,35,40), f=c(0,.04,.06,.25,.1,.01))
+f_young_p = predict(loess(f~x, data = f_young), 15:40)
+f_old_p = predict(loess(f~x, data = f_old), 15:40)
+f_young_mean = sum(f_young_p*15:40/sum(f_young_p))
+f_old_mean = sum(f_old_p*15:40/sum(f_old_p))
+m_si_hump = data.frame(x= c(0,1,5,10,15,20,25), l=c(1,.98,.97,.97,.96,.9,.85))
+m_si_hump_sp = splinefun(m_si_hump, method = "monoH.FC")
+
+#graph
+library(latex2exp)
+par(mfrow=c(1,1),mar = c(5,5,2,5))
+plot(15:40, f_young_p, t="l",xlab="x (mother age)", ylab= TeX(sprintf("$m_x$"))) 
+lines(15:40, f_old_p, lty=2) 
+abline(v=f_young_mean, lty=2, col="grey")
+abline(v=f_old_mean, lty=2, col="grey")
+par(new = T)
+plot(0:25, rev(m_si_hump_sp(0:25)), axes=F, xlab=NA, ylab=NA, t="l", col=2)
+axis(side = 4)
+mtext(side = 4, line = 3, TeX(sprintf("$l_{40-x}$")))
+
+#calculate entropy CS
+H_young = -sum(f_young_p*rev(m_si_hump_sp(0:25))*log(rev(m_si_hump_sp(0:25))))/
+  sum(f_young_p*rev(m_si_hump_sp(0:25)))
+H_old = -sum(f_old_p*rev(m_si_hump_sp(0:25))*log(rev(m_si_hump_sp(0:25))))/
+  sum(f_old_p*rev(m_si_hump_sp(0:25)))
 ```
@@
 
-Considering that $m_{x,\delta}=m_{x}+\delta$ and $l_{a-x} = e^{-\int_{0}^{a-x}{(\mu_t+\delta})}$:
+Considering that $m_{x,\delta}=m_{x}+\delta$ ((@wrycza2012)), and $l_{a-x} = e^{-\int_{0}^{a-x}{(\mu_t+\delta})}$ :
 
@@
 
-To find the effects on daughter survival of adding $\delta$ to the age-specific death rates of daughters, we get the derivative of $dCS_{(a)}^\delta / d\delta$, evaluated near zero (Keyfitz \& Caswell, 2005, section 4.3):
+To find the effects on daughter survival of adding $\delta$ to the age-specific death rates of daughters, we get the derivative of $dCS_{(a)}^\delta / d\delta$, evaluated near zero (@KeyfitzMath, section 4.3):
 
@@
 \label{mort_change_relative}
-\frac{d CS_{a}}{CS_{a}} = \frac{ \int_{\alpha}^{a}m_{x} l_{a-x} \log(l_{a-x}) dx}{\int_{\alpha}^{a}m_{x} l_{a-x}} d\delta
+\frac{d CS_{a}}{CS_{a}} = \frac{ \int_{\alpha}^{a}m_{x} l_{a-x} \log(l_{a-x}) dx}{CS_{a}} d\delta
 \end{equation}
@@
 
-Using that $C_{CS_a}(x)=\frac{m_{x}l_{a-x}}{CS_{a}}$ is a weighting function and making an analogy with the classical entropy function $H$ where the weights are just the survival function, we can define this information function $H_{CS_a}=\int_\alpha^a{C_{CS_a}H(x)}$, and get:
+Calling $C_{CS_a}(x)=\frac{m_{x}l_{a-x}}{CS_{a}}$ the weighting function of the succesful fertility pattern (onset distribution of survival daughters), and making an analogy with the classical entropy function $H$ where the weights are just the survival function (@KeyfitzMath), we can define this information function $H_{CS_a}=\int_\alpha^a{C_{CS_a}H(x)}$, and get:
 
@@
 
-<!-- Siento que esta parte necesita un poco mas de explicacion -->
+This section proposes three additional measures related to the child survival measure for a mother aged $a$.
 
-This section proposes two additional measures of child survival for a mother aged $a$.
-$MTSL$ is an absolute measure of the *Mean Time Spent with Lost*, which can be expressed in terms of a temporary expected lost years measure $e^\dagger$:
+*Mean Time Spent with Lost* ($MTSL$) is an absolute measure of the expected time with a death daughter that lives a mother, which can be expressed in terms of a temporary expected lost years measure $e^\dagger$:
 
@@
 \label{eq:MTSL}
-MTSL_a = \int_\alpha^a{m_x\int_0^{a-x}{d_t e_{t,a-x}dt}dx} = \int_\alpha^a{m_x e_{o,a-x}^\dagger dx}.
+MTSL_a = \int_\alpha^a{m_x\int_0^{a-x}{d_t e_{0,a-x-t} dt}dx} = \int_\alpha^a{m_x e_{o,a-x}^\dagger dx}.
 \end{equation}
 
-It is reasonable to expect a mother to relate these lost years with the expected ones, feeling the *intensity* of her losts (*Intensity Time Lost*).
-This makes it possible to compare between fertility regimes:
+Where $d_t$ is the death distribution from birth, and $e_{0,a-x-t}$ is the life expectancy at birth until age $a-x-t$.
+
+It is reasonable to think that mothers relates these lost years with the expected time with live daughters, getting the *intensity* of her losts (*Intensity Time Lost* ($ITL$)).
+This makes possible also to compare between fertility regimes:
 
@@
 
-Different populations could have different mean age at lost of mothers with death child:
+Different populations could have different mean age at lost ($MAL$) of mothers with death child:
 
@@
 
-The proposed meassures are applied to two different demographic regimes.
+<!-- Aquí tenemos dos opciones de aplicación  -->
+
+* The proposed meassures are applied to two different demographic regimes.
 
@@
 
-A computation approach is used to evaluate the relative contribution of each component given a relative change in $\delta$.
+* A computation approach is used to evaluate the relative contribution of each component given a relative change in $\delta$.
 

### 2019-08-13 draft.Rmd
Cambios luego de sugerencias de Diego. Incorporé gráfico de cambio relativo y extendí un poco la explicación en las medidas relativas.

@@
 ---
-title: "Changes in mortality and child survival for mothers"
+title: "Child survival for mothers: mortality change and related measures"
 author: "Ivan Williams and Diego Alburez-Gutierrez"
 date: "07 August 2019"
-output: html_document
+output: pdf_document
 bibliography: bib.bibtex
@@
 
-A discrete approximate of the effect on $CS_a$ given an absolute change in mortality $m_{x,\delta}=m_{x}+\delta$ is:
+A discrete approximate of the effect on $CS_a$ given an absolute change in mortality $m_{x,\delta}=m_{x}+\delta$ is (@wrycza2012):
 
@@
 
-The effect of a relative change $\delta$ will be bigger when the mortality is more disperse in ages where the fertility is very concentrated.
-For example in \ref{fig:relative_change}, the youngest schedule would have a bigger change in $C_{CS_a}$...
+The effect of a relative change $\delta$ will be bigger when the mortality is more disperse in ages where the fertility is more concentrated.
+For example in \ref{fig:relative_change}, with a big hump in first adult ages, the youngest schedule would have a bigger change in than the older one in $C_{CS_a}$.
 
-```{r relative_change, echo=F, warning=F, fig.cap="Two different fertility schedules (black lines) and a big hump mortality case (right axis, line in red).
-In dashed grey lines are the mean age at birth"}
+```{r relative_change, echo=F, fig.align='center', warning=F, out.width = '70%', fig.cap="Two different fertility schedules (black lines) and a big hump mortality case (right axis, line in red).
+In dashed grey lines are the mean age at birth for each pattern."}
 
@@
 
-Considering that $m_{x,\delta}=m_{x}+\delta$ ((@wrycza2012)), and $l_{a-x} = e^{-\int_{0}^{a-x}{(\mu_t+\delta})}$ :
+Considering that $m_{x,\delta}=m_{x}+\delta$ (@wrycza2012), and $l_{a-x} = e^{-\int_{0}^{a-x}{(\mu_t+\delta})}$ :
 
@@
 
+This section proposes two additional measures related to the child survival measure for a mother aged $a$.
+
 ### 4.1. Time with children lost
 
-This section proposes three additional measures related to the child survival measure for a mother aged $a$.
-
-*Mean Time Spent with Lost* ($MTSL$) is an absolute measure of the expected time with a death daughter that lives a mother, which can be expressed in terms of a temporary expected lost years measure $e^\dagger$:
+*Mean Time Spent with Lost* ($MTSL$) is an absolute measure of the expected total time with a death daughter that lives a mother, which can be expressed in terms of a temporary expected lost years measure $e^\dagger$:
 
@@
 
-It is reasonable to think that mothers relates these lost years with the expected time with live daughters, getting the *intensity* of her losts (*Intensity Time Lost* ($ITL$)).
+It is reasonable to think that mothers relates these lost years with the expected time with live daughters, getting the *intensity* of her losts (*Intensity Time Lost* ($ITL$)): ratio between expected time with "lost"" and expected time with "life".
 This makes possible also to compare between fertility regimes:
@@
 
-Different populations could have different mean age at lost ($MAL$) of mothers with death child:
+Different populations could have different mean age at lost ($MAL_a$) for mothers aged $a$ with at least a death child:
 
@@
 
-$K_a$ is the mean age at birth for womens aged $a$, $MAD_{a-x}$ refers to the mean age at death for those newborns who dies before $a-x$, $TFR_a$ is the cumulated fertility for a women aged a, and $f_{x,a}$ is the cumulative distribution of fertility until age $a$.
+$K_a$ is the mean age at birth for womens aged $a$, $MAD_{a-x}$ refers to the mean age at death for those newborns who dies before $a-x$, $TFR_a$ is the cumulated fertility for a women aged a, and $f_{x,a}$ is the cumulative age distribution of fertility until age $a$.
 
-Equation \ref{eq:MAL} can be derived by starting with the integral of the age $x+t$ at each death child weighted by fertility and survival function:
+Equation \ref{eq:MAL} can be derived by starting with the mother age $x+t$ at each death child age at death $t$, weighted by the fertility and survival function:
 
@@
   MAL_a = \frac{\int_\alpha^a{m_x x}}{TFR_a} + \frac{\int_\alpha^a{m_x MAD_{a-x}}}{TFR_a} \\
-  MAL_a = K_a + \frac{\int_\alpha^a{m_x MAD_{a-x}}}{TFR_a} \\
-  MAL_a = K_a + \int_\alpha^a{f_{x,a} MAD_{a-x}}
+  MAL_a = K_a + \frac{\int_\alpha^a{m_x MAD_{a-x}}}{TFR_a}
 \end{split}
 \end{equation}
+
+Arriving to the equation \ref{eq:MAL}.
 

### 2019-08-16 draft.Rmd
diego revision 2

@@
 
-The paper proposes new expresions to characterize the effects of changes in mortality on child survival.
-The different meassures it introduces can be applied to populations at different stages of the demographic transition.
+The paper proposes new expresions and measures to characterize the effects of changes in mortality on child survival.
+These can be applied to populations at different stages of the demographic transition.
 
@@
 We now consider the consequences of a change $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk.
-
 A discrete approximate of the effect on $CS_a$ given an absolute change in mortality $m_{x,\delta}=m_{x}+\delta$ is (@wrycza2012):
@@
 In this equation, the expected change in child survival is inversely proportional to the difference between maternal age \textit{a} and the mean age of the mother at the birth of her surviving daughters $\overline{x}_{(CS)}$.
-In other words, the magnitude of the change depends negatively on the age distribution of the surviving offspring (as younger children experience longer periods of exposure to risk).
+The magnitude of the change depends negatively on the age distribution of the surviving offspring.
+This is intuitive considering that younger children experience longer periods of exposure to risk.
 
@@
 
-Where $H_{CS_a}$ can be interpreted as a temporary entropy constant until age $a-\alpha$ . 
-Depending the age structure of the survival daughters $C_{CS_a}(x)=\frac{m_{x}l_{a-x}}{CS_{a}}$,  is the way that is taken each age contribution to the disparity measure $H_{CS_a}=\int_\alpha^a{C_{CS_a}(x)H(x)}$ (@Aburto2019).
+Where $H_{CS_a}$ can be interpreted as a temporary entropy constant until age $a-\alpha$.
+The age structure of the surviving daughters $C_{CS_a}(x)=\frac{m_{x}l_{a-x}}{CS_{a}}$ determines the contribution of each age to the disparity measure $H_{CS_a}=\int_\alpha^a{C_{CS_a}(x)H(x)}$ (@Aburto2019).
+The effect of a relative change $\delta$ on $C_{CS_a}$ will be stronger in cases where the mortality is more disperse at ages where fertility is high.
+Consider the two fertility schedules depicted in Figure \ref{fig:relative_change}.
+Against the backdrop of a mortality hump in early adulthood (dashed line), we anticipate a more pronounced change in $C_{CS_a}$ for the younger fertility schedule.
 
-The effect of a relative change $\delta$ will be bigger when the mortality is more disperse in ages where the fertility is more concentrated.
-For example in \ref{fig:relative_change}, with a big hump in first adult ages, the youngest schedule would have a bigger change in than the older one in $C_{CS_a}$.
-
-```{r relative_change, echo=F, fig.align='center', warning=F, out.width = '70%', fig.cap="Two different fertility schedules (black lines) and a big hump mortality case (right axis, line in red).
-In dashed grey lines are the mean age at birth for each pattern."}
+```{r relative_change, echo=F, fig.align='center', warning=F, out.width = '70%', fig.cap="Two different fertility schedules (solid lines) and a big mortality hump (right axis, dashed line).
+The vertical lines show the mean age at birth for each schedule."}
 
@@
 plot(15:40, f_young_p, t="l",xlab="x (mother age)", ylab= TeX(sprintf("$m_x$"))) 
-lines(15:40, f_old_p, lty=2) 
+lines(15:40, f_old_p, lty=1, col=2) 
 abline(v=f_young_mean, lty=2, col="grey")
@@
 par(new = T)
-plot(0:25, rev(m_si_hump_sp(0:25)), axes=F, xlab=NA, ylab=NA, t="l", col=2)
+plot(0:25, rev(m_si_hump_sp(0:25)), axes=F, xlab=NA, ylab=NA, t="l", col=1, lty=2)
 axis(side = 4)
@@
 
-To find the effects on daughter survival of adding $\delta$ to the age-specific death rates of daughters, we get the derivative of $dCS_{(a)}^\delta / d\delta$, evaluated near zero (@KeyfitzMath, section 4.3):
+We get the derivative of $dCS_{(a)}^\delta / d\delta$ evaluated near zero (@KeyfitzMath) to find the effects of adding $\delta$ to the age-specific death rates of daughters:
 
@@
 \label{mort_change_relative}
-\frac{d CS_{a}}{CS_{a}} = \frac{ \int_{\alpha}^{a}m_{x} l_{a-x} \log(l_{a-x}) dx}{CS_{a}} d\delta
+\frac{d CS_{a}}{CS_{a}} = \frac{ \int_{\alpha}^{a}m_{x} l_{a-x} \log(l_{a-x}) dx}{CS_{a}} d\delta.
 \end{equation}
 
-<!-- I think this needs re-weriting for clarity but I didn't want to change it to avoid messing it up! -->
-
-<!-- Podrias dar alguna referencia para lo de entropia?  -->
-
-Calling $C_{CS_a}(x)=\frac{m_{x}l_{a-x}}{CS_{a}}$ the weighting function of the succesful fertility pattern (onset distribution of survival daughters), and making an analogy with the classical entropy function $H$ where the weights are just the survival function (@KeyfitzMath), we can define this information function $H_{CS_a}=\int_\alpha^a{C_{CS_a}H(x)}$, and get:
+We now define the information function $H_{CS_a}=\int_\alpha^a{C_{CS_a}H(x)}$ in analogy to the classic entropy function $H$ (@KeyfitzMath, @Aburto2019).
+Whereas the weights in the latter are given by the survival function, in Equation \ref{eq:relat_change2} they come from the onset distribution of surviving daughters $C_{CS_a}(x)=\frac{m_{x}l_{a-x}}{CS_{a}}$, so that:
 
@@
 
-It is reasonable to think that mothers relates these lost years with the expected time with live daughters, getting the *intensity* of her losts (*Intensity Time Lost* ($ITL$)): ratio between expected time with "lost"" and expected time with "life".
-This makes possible also to compare between fertility regimes:
+It is useful to compare these lost years with the expected time that mother will live with their daughters.
+We call this the *intensity* of loss (*Intensity Time Lost* ($ITL$)): ratio between expected time with "lost"" and expected time with "life".
+This also makes possible to compare between fertility regimes:
 
@@
 
-Different populations could have different mean age at lost ($MAL_a$) for mothers aged $a$ with at least a death child:
+Different populations could have different mean age at child loss ($MAL_a$) for mothers aged $a$ with at least a death child:
 
@@
 
-$K_a$ is the mean age at birth for womens aged $a$, $MAD_{a-x}$ refers to the mean age at death for those newborns who dies before $a-x$, $TFR_a$ is the cumulated fertility for a women aged a, and $f_{x,a}$ is the cumulative age distribution of fertility until age $a$.
-
-Equation \ref{eq:MAL} can be derived by starting with the mother age $x+t$ at each death child age at death $t$, weighted by the fertility and survival function:
+Equation \ref{eq:MAL} can be derived by starting with the mother age $x+t$ at each death child age at death $t$, weighted by the fertility and survival function.
+In it, $K_a$ is the mean age at childbirth for womens aged $a$, $MAD_{a-x}$ refers to the mean age at death for newborns that die before $a-x$, $TFR_a$ is the cumulated fertility for a women aged $a$, and $f_{x,a}$ is the cumulative age distribution of fertility up to age $a$:
 
@@
 
-Arriving to the equation \ref{eq:MAL}.
+<!-- Arriving to the equation \ref{eq:MAL}. -->
 

### 2020-04-07 ALAP2020/ALAP.Rmd
paper for ALAP Improving things

---
title: 'Child survival for mothers: mortality change and related measures'
author: "Ivan Williams and Diego Alburez-Gutierrez"
date: "07 August 2019"
output:
  html_document:
    fig_caption: yes
bibliography: bib.bibtex
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message=F, warning=F)
library(tidyverse)
library(gridExtra)
library(quadprog)
source("QOSplit.R")

options(digits=2)

# age survivor spline
lx_fun <- function(data, age){
  splinefun(data$x, data$lx, method = "monoH.FC")(age)
}
library(wpp2019)
```

## Introduction

Kin count estimation has a long pedigree in mathematical demography. @lotka1931orphanhood proposed to model the orphanhood in a theorical population, comparing the effect across demographic regimes. Equation  \ref{eq:CS} was originally proposed by @brass_derivation_1953 to estimate child mortality. The notion of estimating the expected number of living daughters in a stable population was generalized by @goodman1974 for other kin relations (granddaughters, cousins, etc.). The "counting method" approach, being the net reproduction rate $R_0$ limited to $a$, or the (not complete) ratio between generations (multiplying and dividing by the mother generation $B$), was further popularized by @KeyfitzMath. @bongaarts_projection_1987 used a similar approach to estimate descendants in his 'Family Status Model'. More recently, @wrycza2012 has looked at the formal effect of different kind of changes in mortality by age.

<!---
¿Why is important study survival descendants from mother perspective?¿How impacts mothers health/welfare/experience?
-->

The objective of this work is to give a formal approach in an stable regime context to the timing of descedent mortality (mean age at the event and time spent in that state, absolute and relative) and the effect of mortality change. Finally, a numerical application of these expressions will be done for Latinoamerican countries during the second half of XX Century, relaxing also the stable assumption with an application.

## Method

$CS_a$ is the expected number of surviving children to a woman alive at aged $a$ in a female stable population with fertility rates $m_{x}$, mortality hazard $\mu_{x}$ and survival function $l_x=e^{-\int_{0}^{x}{\mu_t}\,dt}$ ($l_0=1$), as proposed by @goodman1974:

$$CS_a = \int_{0}^a{m_{x}l_{a-x}dx}$$

Following @KeyfitzMath, if we take $a>=\beta$, the fertility pattern by age is concentrated around $\kappa$, and cumulated fertility (gross reproduction rate because of female model) is $F =  \int_{\alpha}^a{m_{x}dx}$, an approximation of $CS_a$ using Taylor´s theorem until second order around $\kappa$ would be:

$$CS_a \approx l_{a-\kappa} \int_{\alpha}^a{m_{x}dx}+(l_{a-\kappa})^{'}\int_{0}^a{(x-\kappa)m_{x}dx}+
(l_{a-\kappa})^{''}\int_{0}^a{\frac{(x-\kappa)^2}{2} m_{x}dx}\\
  \approx F_a l_{a-\kappa} + \frac{\sigma^2}{2}F_a (l_{a-\kappa})^{''}\\
  \approx F_a \, l_{a-\kappa}[1+\frac{\sigma^2}{2}\frac{(l_{a-\kappa})^{''}}{l_{a-\kappa}}]$$

The second Taylor´s term is null because $\int_{\alpha}^a{x \, m_{x}dx} = \kappa \, F_a$. The formula tells us that the survival experience depends of cumulated fertility (possible cases) and child survival (success cases), but also how concentrated is the fertily around $\kappa$ and the suvival curvature around.

From the mother perspective, the time in lost is a terrible situation. We call *Mean Time Spent with Lost* (*MTSL*) the absolute measure of the expected total time with a death daughter that lived a mother aged *a*, which can be expressed in terms of a temporary expected lost years measure in line with $e^\dagger$ ():

$$MTSL_a = \int_0^a{m_x\int_0^{a-x}{d_t e_{t,a-x-t}\, dt}\,dx} = \int_\alpha^a{m_x e_{0|a-x}^\dagger \,dx}$$.

Where $d_t$ is the death distribution from birth, $e_{0,a-x-t}$ is the life expectancy at birth until age $a-x-t$ and $e_{0|a-x}^\dagger$ a temporary dispertion measure. But most interesting would be to compare these lost years with the expected time that mothers would expect to live with their daughters. We call this the *intensity* of loss (*Intensity Time Lost* ($ITL$)): a ratio between expected time in "lost" and expected time in "life", that allows to make comparisons in time and between populations.

$$ITL_a = \frac{\int_\alpha^a{m_x e_{a-x}^\dagger dx}}{\int_\alpha^a{m_x e_{a-x} dx}}$$

Considering woman that losted a child, different mean age at child loss ($MAL$) impacts the mother experience. This relation can be derived by starting with the mother age $x+t$ at each death child age at death $t$, weighted by the fertility and survival function. In it, $\kappa$ is the mean age at childbirth for womens aged $a$, $MAD_{a-x}$ refers to the mean age at death for newborns that die before $a-x$, $F_a$ is the cumulated fertility for a women aged $a$:

$$MAL_a = \frac{\int_0^a{m_x {\int_0^{a-x}\frac{d_t (x+t) dt}{{d_{0,a-x}}}}}}{\int_0^a{m_x}} \\
  MAL_a = \frac{\int_0^a{m_x {\int_0^{a-x}\frac{d_t (x+t) dt}{{d_{0,a-x}}}}}}{\int_0^a{m_x}} \\
  MAL_a = \frac{\int_0^a{m_x x}}{F_a} + \frac{\int_0^a{m_x MAD_{a-x}}}{F_a} \\
  MAL_a = \kappa + \frac{\int_0^a{m_x MAD_{a-x}}}{F_a}$$

This is the sum

We now consider the consequences of a change $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk. A discrete approximate of the effect on $CS_a$ given an absolute change in mortality, considering that $m_{x,\delta}=m_{x}+\delta$ (@wrycza2012), and $l_{a-x} = e^{-\int_{0}^{a-x}{(\mu_t+\delta})}$:

$$CS_{(a)}^\delta = \int_{\alpha}^{a}m_{x} l_{a-x} e^{-\delta (a-x)} dx$$

We get the derivative of $dCS_{(a)}^\delta / d\delta$ evaluated near zero (@KeyfitzMath) to find the effects of adding $\delta$ to the age-specific death rates of daughters:

$$= -a\int_{\alpha}^{a}m_{x} l_{a-x} dx + \int_{\alpha}^{a}m_{x} l_{a-x}x dx$$

Since $CS_{a} = \int_{\alpha}^{a}m_{x} l_{a-x} dx$, we can rewrite equation (\ref{eq2}) as

$$\frac{dCS^\delta}{d\delta} = -a \, CS_{a}  + \int_{\alpha}^{a} x  \, m_{x} l_{a-x} dx$$

Dividing both sides by $CS_{a}$ in an discrete approzimation we get

$$\frac{\Delta CS_{a}}{CS_{a}} \approx -(a - \kappa_{CS}) \Delta\delta$$

In this equation, the expected change in child survival is inversely proportional to the difference between maternal age $a$ and the mean age of the mother at the birth of her surviving daughters $\kappa_{CS}$ (always smaller than $\kappa$). The magnitude of the change depends negatively on the age distribution of the surviving offspring. This is intuitive considering that younger children experience longer periods of exposure to risk.

The relationship is less clear for proportional changes in mortality given the interaction of birth and mortality rates (@KeyfitzMath). Considering that $m_{x,\delta}=m_{x}(1+\delta)$ and $l_{a-x,\delta} = e^{-\int_{0}^{a-x}{\mu_t(1+\delta})dt}=l_{a-x}^{(1+\delta)}$:

$$CS_{a}^\delta = \int_{\alpha}^{a} {m_{x} l_{a-x}^{(1+\delta)}} dx$$

Using the derivative of $dCS_{(a)}^\delta / d\delta = log(l_{a-x}) l_{a-x}^{(1+\delta)}$, reversing between $t$ and $x$:

$$\frac{d CS_{a}}{d \delta} = \int_{\alpha}^{a} {m_{x} l_{a-x} \log(l_{a-x}) dx}\\
= - \int_{0}^{a} {m_{x} l_{a-x} \int_{0}^{a-x}{\mu_x \, dt}\, dx}\\
= - \int_{0}^{a}{\mu_t} \int_{0}^{a-t} {m_{x} l_{a-t-x} \, dx}\, dt$$

Dividing by $CS_a$ and multiplying by $\delta$ get the final expression. This is the negative of cumulative hazard to age *a* times a factor $0\leq \frac{CS_{a-t}}{CS_a}\leq1$ (strictly incremental becasuse $m_x\geq0$ and $l_x\geq0$), that will give more weight to mortality in first ages.

$$\frac{\Delta CS_{a}}{CS_a} = - \left[\int_{0}^{a}{\mu_t  \, \frac{CS_{a-t}}{CS_a} \,dt \,}\right] \Delta \delta$$
```{r main_funtion}
# function to calculate all the measures

CS_results <- function(country, period, a){

  # filter country
  country = LA %>% filter(name == country, Period==period)%>% 
    mutate(dx = c(-diff(lx),lx[101]),
           Lx = ifelse(x==0, Lx_n,
                       ifelse(x==100, ex, 
                              lx-dx/2)),
           Lx = as.numeric(Lx),
           ex = rev(cumsum(rev(Lx)))/lx)

  # calculate
  CS = sum(unlist(sapply(1:a, function(i) country$asfr[i]*country$lx[a-(i-1)]/100000)), na.rm = T)
  Fa = sum(country$asfr[1:a], na.rm = T)
  CS_prob = CS/Fa
  k = sum(country$asfr[1:a]*0:(a-1), na.rm = T)/Fa
  CS_aprox = Fa * country$lx[a-k+1]/100000
  error_aprox = (CS-CS_aprox)/CS*100

  MTSL = 0
  for(i in 1:a){
    for(j in 1:(a-i)){
      MTSL = sum(MTSL,
                 country$asfr[i] * country$dx[j]/100000 * sum(country$Lx[j:a])/country$lx[j],
                 na.rm=T)
    }
  }
  MTSA = 0
  for(i in 1:a){
    for(j in 1:(a-i)){
      MTSA = sum(MTSA,
                 country$asfr[i] * sum(country$Lx[j:a])/country$lx[j],
                 na.rm=T)
    }
  }
  ITL = MTSL/MTSA * 100
  m_MAD = 0
  for(i in 1:a){
    for(j in 1:(a-i)){
      m_MAD = sum(m_MAD,
                  country$asfr[i] * country$dx[j]/100000 * (j-1),
                  na.rm=T)
    }
  }
  MAL = k + m_MAD/Fa
  k_CS = sum(unlist(sapply(1:a, function(i) country$asfr[i]*country$lx[a-(i-1)]/100000*i)), na.rm = T)/CS
  abs_change = -(a-k_CS)

  # Results
  return(data.frame(CS,Fa,CS_prob,k,CS_aprox,error_aprox,MTSL,ITL,MAL))
}
```

## Data

Mortality and fertility data was obtained from the last revision of population prospects by UN (@WPP19). We smoothed female $l_x$ using cubic-splines constrained to monothonic descense, taking later $L_0$ and $T_{100}$ as inputs for year-person calculations. For splitting fertility five groups was used quadratic optimization approach by @Michalski2018, with an desirable property for our purpose which is a good fitting in parity. Also was assumed an unique female percentage of newborns of `r im = 100/204; round(im, 4)`.

```{r Get Data}
# Get wpp locations
data(UNlocations)

# countries of LA
paises_LA_seleccion <- c(192,214,332,388,630,188,222,320,340,484,558,591,32,68,76,152,170,218,600,604,858,862)

# get fertility data
data(tfr)
data(percentASFR)
tfr_paises_LA <- tfr %>% select(-last.observed) %>% gather(Period,tfr,-country_code,-name) %>% filter(country_code %in% paises_LA_seleccion)

asfr_paises_LA <- percentASFR %>% gather(Period, asfr, -name, -country_code, -age) %>% 
  mutate(x = as.numeric(substr(age,1,2)), asfr=asfr/100) %>% 
  left_join(tfr_paises_LA, by=c("country_code","name","Period")) %>% 
  mutate(asfr = asfr/5 * tfr) %>% 
  select(-age) %>% 
  filter(country_code %in% paises_LA_seleccion)

# lt must downloaded manually
lt_paises_LA <- readxl::read_xlsx("Data/WPP2019_MORT_F17_3_ABRIDGED_LIFE_TABLE_FEMALE.xlsx", sheet = 1, range = "A17:T77381") %>% rename(country_code=5,
                                      x = "Age (x)", 
                                      mx = "Central death rate m(x,n)",
                                      qx = "Probability of dying q(x,n)",
                                      dx = "Number of deaths d(x,n)",
                                      lx = "Number of survivors l(x)",
                                      Lx_n = "Number of person-years lived L(x,n)",
                                      ex = "Expectation of life e(x)")  %>% 
  select(country_code,Period,x, mx, qx, dx, lx, Lx_n, ex) %>% 
  filter(country_code %in% paises_LA_seleccion)

# smooth both
asfr_paises_LA_s <- data.frame(country_code=0, Period = " ", x = 0, asfr = 0, stringsAsFactors = F)
lx_paises_LA_s <- data.frame(country_code=0, Period = " ", x = 0, lx = 0, stringsAsFactors = F)
countries = unique(asfr_paises_LA$country_code)
for(c in 1:length(countries)){
  for(p in unique(asfr_paises_LA$Period)){
    cp = asfr_paises_LA %>% filter(country_code==countries[c], Period==p) %>% select(asfr)
    if(any(is.na(cp$asfr))){next}
    out <- tryCatch(QOSplit(c(0,cp$asfr), L=seq(10,45,5), AgeInt=rep(5,8)),
                    error = data.frame(Age=10:49, ASFR=rep(NA,40)))
    asfr_paises_LA_s <- rbind(asfr_paises_LA_s, data.frame(country_code = countries[c], Period = p, x = 
                                                            out$Age, asfr = out$ASFR))
    cp <- lt_paises_LA %>% filter(country_code==countries[c], Period==p) %>% select(x,lx)
    out <- lx_fun(cp, age=0:100)
    lx_paises_LA_s <- rbind(lx_paises_LA_s, data.frame(country_code=countries[c], Period = p, x = 0:100, lx = out))
  }
}

# join everything
LA <- lx_paises_LA_s %>% slice(-1) %>% mutate(country_code=as.integer(country_code)) %>% 
          left_join(asfr_paises_LA_s %>% slice(-1), by=c("country_code","Period","x")) %>%
          left_join(tfr_paises_LA, by=c("country_code","Period")) %>% 
          left_join(lt_paises_LA %>% 
                      select(country_code, Period, x, Lx_n, ex),
                    by = c("country_code","Period","x")) %>% 
          mutate(q = (1-lx/100000),
                 name = as.character(name),
                 tfr = tfr*im,
                 asfr = asfr*im)
LA$name[LA$name == "Bolivia (Plurinational State of)"] = "Bolivia"

```

An negative relation between mortality until 10 and TFR is shown in \@ref(fig:plot_tfr_q010). Specific paths like Argentina and Guatemalare shows very different demographic profiles during second part of XX century.

```{r plot_tfr_q010, fig.cap="Female probability of death in first 10 years old by gross reproduction rate. Latin America countries in period 1950-2015"}
ggplot(LA %>% filter(x==10), aes(x=tfr, y=q))+
                  geom_point(aes(color=Period)) + 
                  geom_smooth(se = F, color = "grey") +
                  geom_line(data = LA %>% filter(x==10, name %in% c("Guatemala")),
                            aes(group=name), colour=2, size = 1.5) +
                  geom_line(data = LA %>% filter(x==10, name %in% c("Argentina")),
                            aes(group=name), colour=4, size = 1.5) +
                  theme_classic() + labs(x="GRR", y="q(0,10)")
```

## Application

```{r get_results}
CS_outs <- as.data.frame(expand.grid(name = unique(LA$name),
                                     Age = c(30,40,50),
                                     Period = unique(LA$Period)))
for(i in 1:nrow(CS_outs)){
          CS_outs[i,4:12] = CS_results(country = CS_outs[i,1], 
                               period = CS_outs[i,3], 
                               a = CS_outs[i,2])}
CS_outs <- CS_outs %>% mutate(Year = as.integer(substr(Period, 1, 4)) + 2.5)
```

Between 1950-1955 and 2010-2015 the portion of surviving daugthers  $\frac{CS_a}{F_a}$ at age 50 passed from a simple mean of `r mean(CS_outs$CS_prob[CS_outs$Period == "1950-1955" & CS_outs$Age==50])` to `r mean(CS_outs$CS_prob[CS_outs$Period == "2010-2015" & CS_outs$Age==50])`, also with a decrease of `r mean(CS_outs$CS[CS_outs$Period == "1950-1955" & CS_outs$Age==50])-mean(CS_outs$CS[CS_outs$Period == "2010-2015" & CS_outs$Age==50])` in the expected descendants at same age. Althought this is not a  real measure but a theorical one if the rates would have been kept constant during enough time.

```{r plot_CS}
a <- ggplot(CS_outs, aes(x=Year, y=CS, color=factor(Age)))+
              geom_point(alpha = .3) + 
              geom_smooth(se = F) +
              theme_classic() +
              labs(y = "CS") + theme(legend.position = "none")
b <- ggplot(CS_outs, aes(x=Year, y=CS_prob, color=factor(Age)))+
              geom_point(alpha = .3) + 
              geom_smooth(se = F) +
              theme_classic() +
              labs(y = "CS/F") + 
              scale_colour_discrete(name="Age")
grid.arrange(a,b,ncol=2)
```

The approximation in () is very accurate, improving with the years. In the case of Guatemala, is atribuitable to the $l_x$ rectangularization process.

```{r plot_CS_aprox, fig.cap = "Error (%) in approximation, 1950 (darker) to 2015 (lighter) (left). Guatemala case (right)"}
a <- ggplot(CS_outs, aes(x=CS, y=error_aprox, color=Year))+
                    geom_point(alpha = .6) + 
                    theme_classic() +
                    theme(legend.position = "none") +
                    labs(y = "%")
b <- LA %>% filter(name=="Guatemala") %>% ggplot() +
            geom_line(aes(x=x, y=lx/100000, color=Period)) +
                    theme_classic() +
                    theme(legend.position = "none") +
                    labs(y = "")
c <- LA %>% filter(name=="Guatemala", between(x, 15,50)) %>% ggplot() +
            geom_line(aes(x=x, y=asfr/tfr*100, color=Period)) +
                    theme_classic() +
                    theme(legend.position = "none") +
                    labs(y = "%")
grid.arrange(a, grid.arrange(b,c,ncol=1), ncol=2)
```

The expected time in years that a mother aged 30 passed with a death son was around 4% in some countries at middle XX Century. When increasing age, the survival experience depends less on infant mortality, and the distribution is around less than 2% for womans aged 50, converging to 0 on time.

```{r plot_ITL, fig.cap = "Intensity Tome Lost of  womans aged 30, 40 and 50. Years 1950 (darker) to 2015 (lighter)"}
ggplot(CS_outs)+  geom_point(aes(x=Fa, y=ITL, color=Year), alpha = .3) +
                  theme_classic()+        
                  theme(legend.position = "none") +
                  labs(y = "%")+
                  facet_grid(~Age) 
```

For those woman that lost a daughter, more fertility more the mean age.

```{r plot_MAL}
ggplot(CS_outs %>% filter(Age==50))+
                    geom_smooth(aes(x=Fa, y=MAL), method = "lm", col="grey", se = F)+
                    geom_point(aes(x=Fa, y=MAL, color=Year), alpha = .3) +
                    theme_classic()+
                    theme(legend.position = "none") +
                    labs(y = "Mean Age at Lost")
```

## Discussion

# pendings

Here an independet assumption is done in the relationship between fertility and descendants survival within population. Both assumptions will be considered at the end.

In terms of mortality, this measure is much more related to first ages (child and youth) than adult or adult ages, so we can model a H-P pattern and derivate some of their parameters...

We ca take cohort rates to compare between countries, relaxing main assumptions in stable model...

Is known that fertility is influenced by child mortality, and that there is an heterogeneus behaviour within. We can add frailty variables to take account of that and express correlation between subgroups behaviour...

```{r relative_change, include = F, echo=F, fig.align='center', warning=F, out.width = '70%', fig.cap="Two different fertility schedules (solid lines) and a big mortality hump (right axis, dashed line). The vertical lines show the mean age at birth for each schedule."}
# 
# # genero m_x (old and young pattern) y l_x (with big hump) 
# f_young = data.frame(x=c(15,20,25,30,35,40), f=c(0,.3,.3,.05,.01,.01))
# f_old = data.frame(  x=c(15,20,25,30,35,40), f=c(0,.04,.06,.25,.1,.01))
# f_young_p = predict(loess(f~x, data = f_young), 15:40)
# f_old_p = predict(loess(f~x, data = f_old), 15:40)
# f_young_mean = sum(f_young_p*15:40/sum(f_young_p))
# f_old_mean = sum(f_old_p*15:40/sum(f_old_p))
# m_si_hump = data.frame(x= c(0,1,5,10,15,20,25), l=c(1,.98,.97,.97,.96,.9,.85))
# m_si_hump_sp = splinefun(m_si_hump, method = "monoH.FC")
# 
# #graph
# library(latex2exp)
# par(mfrow=c(1,1),mar = c(5,5,2,5))
# plot(15:40, f_young_p, t="l",xlab="x (mother age)", ylab= TeX(sprintf("$m_x$"))) 
# lines(15:40, f_old_p, lty=1, col=2) 
# abline(v=f_young_mean, lty=2, col="grey")
# abline(v=f_old_mean, lty=2, col="grey")
# par(new = T)
# plot(0:25, rev(m_si_hump_sp(0:25)), axes=F, xlab=NA, ylab=NA, t="l", col=1, lty=2)
# axis(side = 4)
# mtext(side = 4, line = 3, TeX(sprintf("$l_{40-x}$")))
# 
# #calculate entropy CS
# H_young = -sum(f_young_p*rev(m_si_hump_sp(0:25))*log(rev(m_si_hump_sp(0:25))))/
#   sum(f_young_p*rev(m_si_hump_sp(0:25)))
# H_old = -sum(f_old_p*rev(m_si_hump_sp(0:25))*log(rev(m_si_hump_sp(0:25))))/
#   sum(f_old_p*rev(m_si_hump_sp(0:25)))

```

* The proposed meassures are applied to two different demographic regimes.

[PENDING]

* A computation approach is used to evaluate the relative contribution of each component given a relative change in $\delta$.

[PENDING]

## 5. Bibliography

### 2020-04-07 draft.Rmd
Corrección párrafo
paper for ALAP Improving things

@@
 ---
-title: "Child survival for mothers: mortality change and related measures"
+title: 'Child survival for mothers: mortality change and related measures'
 author: "Ivan Williams and Diego Alburez-Gutierrez"
 date: "07 August 2019"
-output: pdf_document
+output:
+  html_document:
+    df_print: paged
 bibliography: bib.bibtex
@@
 The age structure of the surviving daughters $C_{CS_a}(x)=\frac{m_{x}l_{a-x}}{CS_{a}}$ determines the contribution of each age to the disparity measure $H_{CS_a}=\int_\alpha^a{C_{CS_a}(x)H(x)}$ (@Aburto2019).
-The effect of a relative change $\delta$ on $C_{CS_a}$ will be stronger in cases where the mortality is more disperse at ages where fertility is high.
+The effect of a relative change $\delta$ on $CS_{a}$ will be stronger in cases where the mortality is more disperse at ages where fertility is high.
 Consider the two fertility schedules depicted in Figure \ref{fig:relative_change}.

### 2020-04-08 ALAP2020/ALAP.Rmd

@@
 ```{r setup, include = FALSE}
-knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message=F, warning=F)
+knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
+                      message=F, warning=F)
 library(tidyverse)
@@
 library(quadprog)
+library(wpp2019)
 source("QOSplit.R")
-
 options(digits=2)
-
-# age survivor spline
-lx_fun <- function(data, age){
-  splinefun(data$x, data$lx, method = "monoH.FC")(age)
-}
-library(wpp2019)
 ```
@@
 ```{r main_funtion}
+# age survivor spline
+
+lx_fun <- function(data, age){
+  splinefun(data$x, data$lx, method = "monoH.FC")(age)
+}
+
 # function to calculate all the measures
@@
 # lt must downloaded manually
-lt_paises_LA <- readxl::read_xlsx("Data/WPP2019_MORT_F17_3_ABRIDGED_LIFE_TABLE_FEMALE.xlsx", sheet = 1, range = "A17:T77381") %>% rename(country_code=5,
+lt_paises_LA <- readxl::read_xlsx("WPP2019_MORT_F17_3_ABRIDGED_LIFE_TABLE_FEMALE.xlsx", sheet = 1, range = "A17:T77381") %>% rename(country_code=5,
                                       x = "Age (x)", 
@@
 
-* The proposed meassures are applied to two different demographic regimes.
-
-[PENDING]
-
-* A computation approach is used to evaluate the relative contribution of each component given a relative change in $\delta$.
-
-[PENDING]
-
 ## 5. Bibliography

### 2020-04-15 ALAP2020/ALAP.Rmd
abs change infant mortality

@@
 We now consider the consequences of a change $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk.
-A discrete approximate of the effect on $CS_a$ given an absolute change in mortality, considering that $m_{x,\delta}=m_{x}+\delta$ (@wrycza2012), and $l_{a-x} = e^{-\int_{0}^{a-x}{(\mu_t+\delta})}$:
+A discrete approximate of the effect on $CS_a$ given an absolute change in mortality, considering that $m_{x,\delta}=m_{x}+\delta$ (@wrycza2012), and $l_{a-x} = e^{-\int_{0}^{a-x}{(\mu_t+\delta})dt}$:
 
@@
 
+If the change is only in the age range $\left[0;1\right)$, infant mortality, we can start by splitting the integral^[$a>1$]:
+
+$$CS_{a}^{\delta_0} = \int_{0}^{a-1}
+m_{x} {e{^{-\int_{1}^{a-x}\mu_t \,dt-\int_{0}^{1}(\mu_t + \delta)\,dt}}} dx + 
+\int_{a-1}^{a} m_{x} e^{-\int_{0}^{a-x}(\mu_t+ \delta)\,dt}\\
+=\int_{0}^{a-1} m_{x} l_{a-x}e^{-\delta} dx + \int_{a-1}^{a} m_{x} l_{a-x} e^{-\delta(a-x)}\,dx
+$$
+
+Deriving by $\delta$ and valuating near 0, we get:
+
+$$CS_{a}^{\delta_0} =-\int_{0}^{a-1} m_{x} l_{a-x} dx - a \int_{a-1}^{a} {m_{x} l_{a-x} dx} + \int_{a-1}^{a} {x \, m_{x} l_{a-x} dx}$$
+Expressing $\int_{0}^{a-1} m_{x} l_{a-x} dx$ as $CS_a - CS_{a-1,a}$, and finally dividing by $CS_a$:
+
+$$ = - CS_a + CS_{a-1,a} (1-a+\kappa_{a-1,a})\\
+\frac{\Delta CS_{a}}{CS_{a}} \approx - \left[1-\frac{CS_{a-1,a}}{CS_a} (1-(a-\kappa_{a-1,a}))\right] \Delta\delta$$
+
+Affects proportionally for the range extension of 1, less $P_b$ (if we assume uniform distribution of deaths and births in the first year) times $\frac{CS_{a-1,a}}{CS_a}$ the portion of cumulated alive daughters in the previous year, weighting how important was the last year of reproduction.
+
 The relationship is less clear for proportional changes in mortality given the interaction of birth and mortality rates (@KeyfitzMath).
@@
 $$\frac{\Delta CS_{a}}{CS_a} = - \left[\int_{0}^{a}{\mu_t  \, \frac{CS_{a-t}}{CS_a} \,dt \,}\right] \Delta \delta$$
+
+In a model with heterogeneity, determined at born with a multiplicative effect, $CS_a$ could be interpreted as a conditional expectation with random variables *Y* for fertility heterogeneity and *Z* for frailty.
+
+$$CS_a = \int_{0}^{\infty}f_y \int_{0}^{\infty}f_{z|y}\left[\int_{0}^{\infty}y\,\overline{m}_x\,e^{-\overline{H}_{a-x}\,z}\,dx\right]\,dz\,dy\\
+CS_a = \int_{0}^{\infty}f_y \, y \int_{0}^{\infty}f_{z|y}\left[\int_{0}^{\infty}\,\overline{m}_x\,e^{-\overline{H}_{a-x}\,z}\,dx\right]\,dz\,dy
+$$
+
 ```{r main_funtion}

### 2020-04-23 ALAP2020/ALAP.Rmd
* Add visual cheks to approximations and abs change * Add corollary for infant mort abs change * first mature ideas in heterogeneity * graphs must be improved

--- 2020-04-08 ALAP2020/ALAP.Rmd
@@
 ```{r setup, include = FALSE}
-knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
-                      message=F, warning=F)
+knitr::opts_chunk$set(collapse = TRUE,
+                      message=F, warning=F, echo = F)
 library(tidyverse)
@@
 library(wpp2019)
-source("QOSplit.R")
 options(digits=2)
+source("funs.R")
 ```
@@
 <!---
-¿Why is important study survival descendants from mother perspective?¿How impacts mothers health/welfare/experience?
+Improve Intro: ¿Why is important study survival descendants from mother perspective?¿How impacts mothers health/welfare/experience?
 -->
 
-The objective of this work is to give a formal approach in an stable regime context to the timing of descedent mortality (mean age at the event and time spent in that state, absolute and relative) and the effect of mortality change.
-Finally, a numerical application of these expressions will be done for Latinoamerican countries during the second half of XX Century, relaxing also the stable assumption with an application.
+The objective of this work is to give a formal approach to deomgraphy of surviving childs from mother perspective, looking for relations to characterize this situation: the timing of descedent mortality (mean age at the event and time spent in that state) and the effect of mortality change in the final descendants quantity.
+Numerical applications of the expressions will be done for Latinoamerican countries during the second half of XX Century.
+Details on data handling and smooth procedures are exposed in [data section](#Data_appendix) in the Appendix.
 
-## Method
+## Formulas
 
-$CS_a$ is the expected number of surviving children to a woman alive at aged $a$ in a female stable population with fertility rates $m_{x}$, mortality hazard $\mu_{x}$ and survival function $l_x=e^{-\int_{0}^{x}{\mu_t}\,dt}$ ($l_0=1$), as proposed by @goodman1974:
+$CS_a$ is the expected number of surviving children to a mother^[We´ll refer to mothers, assuming that fertility rates applies to all woman, meaning that all of them are exposed.] alive at aged $a$ in a female stable population with fertility rates $m_{x}$, mortality hazard $\mu_{x}$ and survival function $l_x=e^{-\int_{0}^{x}{\mu_t}\,dt}$ ($l_0=1$ during this work), as proposed by @goodman1974:
 
 $$CS_a = \int_{0}^a{m_{x}l_{a-x}dx}$$
-
-Following @KeyfitzMath, if we take $a>=\beta$, the fertility pattern by age is concentrated around $\kappa$, and cumulated fertility (gross reproduction rate because of female model) is $F =  \int_{\alpha}^a{m_{x}dx}$, an approximation of $CS_a$ using Taylor´s theorem until second order around $\kappa$ would be:
-
-$$CS_a \approx l_{a-\kappa} \int_{\alpha}^a{m_{x}dx}+(l_{a-\kappa})^{'}\int_{0}^a{(x-\kappa)m_{x}dx}+
-(l_{a-\kappa})^{''}\int_{0}^a{\frac{(x-\kappa)^2}{2} m_{x}dx}\\
-  \approx F_a l_{a-\kappa} + \frac{\sigma^2}{2}F_a (l_{a-\kappa})^{''}\\
-  \approx F_a \, l_{a-\kappa}[1+\frac{\sigma^2}{2}\frac{(l_{a-\kappa})^{''}}{l_{a-\kappa}}]$$
-
-The second Taylor´s term is null because $\int_{\alpha}^a{x \, m_{x}dx} = \kappa \, F_a$.
-The formula tells us that the survival experience depends of cumulated fertility (possible cases) and child survival (success cases), but also how concentrated is the fertily around $\kappa$ and the suvival curvature around.
-
-From the mother perspective, the time in lost is a terrible situation.
-We call *Mean Time Spent with Lost* (*MTSL*) the absolute measure of the expected total time with a death daughter that lived a mother aged *a*, which can be expressed in terms of a temporary expected lost years measure in line with $e^\dagger$ ():
-
-$$MTSL_a = \int_0^a{m_x\int_0^{a-x}{d_t e_{t,a-x-t}\, dt}\,dx} = \int_\alpha^a{m_x e_{0|a-x}^\dagger \,dx}$$.
-
-Where $d_t$ is the death distribution from birth, $e_{0,a-x-t}$ is the life expectancy at birth until age $a-x-t$ and $e_{0|a-x}^\dagger$ a temporary dispertion measure.
-But most interesting would be to compare these lost years with the expected time that mothers would expect to live with their daughters.
-We call this the *intensity* of loss (*Intensity Time Lost* ($ITL$)): a ratio between expected time in "lost" and expected time in "life", that allows to make comparisons in time and between populations.
-
-$$ITL_a = \frac{\int_\alpha^a{m_x e_{a-x}^\dagger dx}}{\int_\alpha^a{m_x e_{a-x} dx}}$$
-
-Considering woman that losted a child, different mean age at child loss ($MAL$) impacts the mother experience.
-This relation can be derived by starting with the mother age $x+t$ at each death child age at death $t$, weighted by the fertility and survival function.
-In it, $\kappa$ is the mean age at childbirth for womens aged $a$, $MAD_{a-x}$ refers to the mean age at death for newborns that die before $a-x$, $F_a$ is the cumulated fertility for a women aged $a$:
-
-$$MAL_a = \frac{\int_0^a{m_x {\int_0^{a-x}\frac{d_t (x+t) dt}{{d_{0,a-x}}}}}}{\int_0^a{m_x}} \\
-  MAL_a = \frac{\int_0^a{m_x {\int_0^{a-x}\frac{d_t (x+t) dt}{{d_{0,a-x}}}}}}{\int_0^a{m_x}} \\
-  MAL_a = \frac{\int_0^a{m_x x}}{F_a} + \frac{\int_0^a{m_x MAD_{a-x}}}{F_a} \\
-  MAL_a = \kappa + \frac{\int_0^a{m_x MAD_{a-x}}}{F_a}$$
-
-This is the sum
-
-We now consider the consequences of a change $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk.
-A discrete approximate of the effect on $CS_a$ given an absolute change in mortality, considering that $m_{x,\delta}=m_{x}+\delta$ (@wrycza2012), and $l_{a-x} = e^{-\int_{0}^{a-x}{(\mu_t+\delta})}$:
-
-$$CS_{(a)}^\delta = \int_{\alpha}^{a}m_{x} l_{a-x} e^{-\delta (a-x)} dx$$
-
-We get the derivative of $dCS_{(a)}^\delta / d\delta$ evaluated near zero (@KeyfitzMath) to find the effects of adding $\delta$ to the age-specific death rates of daughters:
-
-$$= -a\int_{\alpha}^{a}m_{x} l_{a-x} dx + \int_{\alpha}^{a}m_{x} l_{a-x}x dx$$
-
-Since $CS_{a} = \int_{\alpha}^{a}m_{x} l_{a-x} dx$, we can rewrite equation (\ref{eq2}) as
-
-$$\frac{dCS^\delta}{d\delta} = -a \, CS_{a}  + \int_{\alpha}^{a} x  \, m_{x} l_{a-x} dx$$
-
-Dividing both sides by $CS_{a}$ in an discrete approzimation we get
-
-$$\frac{\Delta CS_{a}}{CS_{a}} \approx -(a - \kappa_{CS}) \Delta\delta$$
-
-In this equation, the expected change in child survival is inversely proportional to the difference between maternal age $a$ and the mean age of the mother at the birth of her surviving daughters $\kappa_{CS}$ (always smaller than $\kappa$).
-The magnitude of the change depends negatively on the age distribution of the surviving offspring.
-This is intuitive considering that younger children experience longer periods of exposure to risk.
-
-The relationship is less clear for proportional changes in mortality given the interaction of birth and mortality rates (@KeyfitzMath).
-Considering that $m_{x,\delta}=m_{x}(1+\delta)$ and $l_{a-x,\delta} = e^{-\int_{0}^{a-x}{\mu_t(1+\delta})dt}=l_{a-x}^{(1+\delta)}$:
-
-$$CS_{a}^\delta = \int_{\alpha}^{a} {m_{x} l_{a-x}^{(1+\delta)}} dx$$
-
-Using the derivative of $dCS_{(a)}^\delta / d\delta = log(l_{a-x}) l_{a-x}^{(1+\delta)}$, reversing between $t$ and $x$:
-
-$$\frac{d CS_{a}}{d \delta} = \int_{\alpha}^{a} {m_{x} l_{a-x} \log(l_{a-x}) dx}\\
-= - \int_{0}^{a} {m_{x} l_{a-x} \int_{0}^{a-x}{\mu_x \, dt}\, dx}\\
-= - \int_{0}^{a}{\mu_t} \int_{0}^{a-t} {m_{x} l_{a-t-x} \, dx}\, dt$$
-
-Dividing by $CS_a$ and multiplying by $\delta$ get the final expression.
-This is the negative of cumulative hazard to age *a* times a factor $0\leq \frac{CS_{a-t}}{CS_a}\leq1$ (strictly incremental becasuse $m_x\geq0$ and $l_x\geq0$), that will give more weight to mortality in first ages.
-
-$$\frac{\Delta CS_{a}}{CS_a} = - \left[\int_{0}^{a}{\mu_t  \, \frac{CS_{a-t}}{CS_a} \,dt \,}\right] \Delta \delta$$
-```{r main_funtion}
-# age survivor spline
-
-lx_fun <- function(data, age){
-  splinefun(data$x, data$lx, method = "monoH.FC")(age)
-}
-
-# function to calculate all the measures
-
-CS_results <- function(country, period, a){
-
-  # filter country
-  country = LA %>% filter(name == country, Period==period)%>% 
-    mutate(dx = c(-diff(lx),lx[101]),
-           Lx = ifelse(x==0, Lx_n,
-                       ifelse(x==100, ex, 
-                              lx-dx/2)),
-           Lx = as.numeric(Lx),
-           ex = rev(cumsum(rev(Lx)))/lx)
-
-  # calculate
-  CS = sum(unlist(sapply(1:a, function(i) country$asfr[i]*country$lx[a-(i-1)]/100000)), na.rm = T)
-  Fa = sum(country$asfr[1:a], na.rm = T)
-  CS_prob = CS/Fa
-  k = sum(country$asfr[1:a]*0:(a-1), na.rm = T)/Fa
-  CS_aprox = Fa * country$lx[a-k+1]/100000
-  error_aprox = (CS-CS_aprox)/CS*100
-
-  MTSL = 0
-  for(i in 1:a){
-    for(j in 1:(a-i)){
-      MTSL = sum(MTSL,
-                 country$asfr[i] * country$dx[j]/100000 * sum(country$Lx[j:a])/country$lx[j],
-                 na.rm=T)
-    }
-  }
-  MTSA = 0
-  for(i in 1:a){
-    for(j in 1:(a-i)){
-      MTSA = sum(MTSA,
-                 country$asfr[i] * sum(country$Lx[j:a])/country$lx[j],
-                 na.rm=T)
-    }
-  }
-  ITL = MTSL/MTSA * 100
-  m_MAD = 0
-  for(i in 1:a){
-    for(j in 1:(a-i)){
-      m_MAD = sum(m_MAD,
-                  country$asfr[i] * country$dx[j]/100000 * (j-1),
-                  na.rm=T)
-    }
-  }
-  MAL = k + m_MAD/Fa
-  k_CS = sum(unlist(sapply(1:a, function(i) country$asfr[i]*country$lx[a-(i-1)]/100000*i)), na.rm = T)/CS
-  abs_change = -(a-k_CS)
-
-  # Results
-  return(data.frame(CS,Fa,CS_prob,k,CS_aprox,error_aprox,MTSL,ITL,MAL))
-}
-```
-
-## Data
-
-Mortality and fertility data was obtained from the last revision of population prospects by UN (@WPP19).
-We smoothed female $l_x$ using cubic-splines constrained to monothonic descense, taking later $L_0$ and $T_{100}$ as inputs for year-person calculations.
-For splitting fertility five groups was used quadratic optimization approach by @Michalski2018, with an desirable property for our purpose which is a good fitting in parity.
-Also was assumed an unique female percentage of newborns of `r im = 100/204; round(im, 4)`.
-
-```{r Get Data}
-# Get wpp locations
-data(UNlocations)
-
-# countries of LA
-paises_LA_seleccion <- c(192,214,332,388,630,188,222,320,340,484,558,591,32,68,76,152,170,218,600,604,858,862)
-
-# get fertility data
-data(tfr)
-data(percentASFR)
-tfr_paises_LA <- tfr %>% select(-last.observed) %>% gather(Period,tfr,-country_code,-name) %>% filter(country_code %in% paises_LA_seleccion)
-
-asfr_paises_LA <- percentASFR %>% gather(Period, asfr, -name, -country_code, -age) %>% 
-  mutate(x = as.numeric(substr(age,1,2)), asfr=asfr/100) %>% 
-  left_join(tfr_paises_LA, by=c("country_code","name","Period")) %>% 
-  mutate(asfr = asfr/5 * tfr) %>% 
-  select(-age) %>% 
-  filter(country_code %in% paises_LA_seleccion)
-
-# lt must downloaded manually
-lt_paises_LA <- readxl::read_xlsx("WPP2019_MORT_F17_3_ABRIDGED_LIFE_TABLE_FEMALE.xlsx", sheet = 1, range = "A17:T77381") %>% rename(country_code=5,
-                                      x = "Age (x)", 
-                                      mx = "Central death rate m(x,n)",
-                                      qx = "Probability of dying q(x,n)",
-                                      dx = "Number of deaths d(x,n)",
-                                      lx = "Number of survivors l(x)",
-                                      Lx_n = "Number of person-years lived L(x,n)",
-                                      ex = "Expectation of life e(x)")  %>% 
-  select(country_code,Period,x, mx, qx, dx, lx, Lx_n, ex) %>% 
-  filter(country_code %in% paises_LA_seleccion)
-
-# smooth both
-asfr_paises_LA_s <- data.frame(country_code=0, Period = " ", x = 0, asfr = 0, stringsAsFactors = F)
-lx_paises_LA_s <- data.frame(country_code=0, Period = " ", x = 0, lx = 0, stringsAsFactors = F)
-countries = unique(asfr_paises_LA$country_code)
-for(c in 1:length(countries)){
-  for(p in unique(asfr_paises_LA$Period)){
-    cp = asfr_paises_LA %>% filter(country_code==countries[c], Period==p) %>% select(asfr)
-    if(any(is.na(cp$asfr))){next}
-    out <- tryCatch(QOSplit(c(0,cp$asfr), L=seq(10,45,5), AgeInt=rep(5,8)),
-                    error = data.frame(Age=10:49, ASFR=rep(NA,40)))
-    asfr_paises_LA_s <- rbind(asfr_paises_LA_s, data.frame(country_code = countries[c], Period = p, x = 
-                                                            out$Age, asfr = out$ASFR))
-    cp <- lt_paises_LA %>% filter(country_code==countries[c], Period==p) %>% select(x,lx)
-    out <- lx_fun(cp, age=0:100)
-    lx_paises_LA_s <- rbind(lx_paises_LA_s, data.frame(country_code=countries[c], Period = p, x = 0:100, lx = out))
-  }
-}
-
-# join everything
-LA <- lx_paises_LA_s %>% slice(-1) %>% mutate(country_code=as.integer(country_code)) %>% 
-          left_join(asfr_paises_LA_s %>% slice(-1), by=c("country_code","Period","x")) %>%
-          left_join(tfr_paises_LA, by=c("country_code","Period")) %>% 
-          left_join(lt_paises_LA %>% 
-                      select(country_code, Period, x, Lx_n, ex),
-                    by = c("country_code","Period","x")) %>% 
-          mutate(q = (1-lx/100000),
-                 name = as.character(name),
-                 tfr = tfr*im,
-                 asfr = asfr*im)
-LA$name[LA$name == "Bolivia (Plurinational State of)"] = "Bolivia"
-
-```
-
-An negative relation between mortality until 10 and TFR is shown in \@ref(fig:plot_tfr_q010).
-Specific paths like Argentina and Guatemalare shows very different demographic profiles during second part of XX century.
-
-```{r plot_tfr_q010, fig.cap="Female probability of death in first 10 years old by gross reproduction rate. Latin America countries in period 1950-2015"}
-ggplot(LA %>% filter(x==10), aes(x=tfr, y=q))+
-                  geom_point(aes(color=Period)) + 
-                  geom_smooth(se = F, color = "grey") +
-                  geom_line(data = LA %>% filter(x==10, name %in% c("Guatemala")),
-                            aes(group=name), colour=2, size = 1.5) +
-                  geom_line(data = LA %>% filter(x==10, name %in% c("Argentina")),
-                            aes(group=name), colour=4, size = 1.5) +
-                  theme_classic() + labs(x="GRR", y="q(0,10)")
-```
-
-## Application
-
-```{r get_results}
-CS_outs <- as.data.frame(expand.grid(name = unique(LA$name),
-                                     Age = c(30,40,50),
-                                     Period = unique(LA$Period)))
-for(i in 1:nrow(CS_outs)){
-          CS_outs[i,4:12] = CS_results(country = CS_outs[i,1], 
-                               period = CS_outs[i,3], 
-                               a = CS_outs[i,2])}
-CS_outs <- CS_outs %>% mutate(Year = as.integer(substr(Period, 1, 4)) + 2.5)
-```
 
@@
 
-```{r plot_CS}
+<!---
+Interpretation
+-->
+
+```{r plot_CS, fig.cap="Child Survival and Child survival related to cumulated fertility, for women aged 30, 40 and 50. Latinamerican countries in period 1950-2015"}
 a <- ggplot(CS_outs, aes(x=Year, y=CS, color=factor(Age)))+
@@
 
-The approximation in () is very accurate, improving with the years. In the case of Guatemala, is atribuitable to the $l_x$ rectangularization process.
+### An approximation
 
-```{r plot_CS_aprox, fig.cap = "Error (%) in approximation, 1950 (darker) to 2015 (lighter) (left). Guatemala case (right)"}
-a <- ggplot(CS_outs, aes(x=CS, y=error_aprox, color=Year))+
+Following @KeyfitzMath work about living mothers count, we can consider $a>=\beta$ and assume that fertility pattern by age is concentrated around $\kappa$ (mean age at childbearing), and cumulated fertility (or gross reproduction rate because of female model) is $F_a =  \int_{\alpha}^a{m_{x}dx}$.
+An approximation of $l_x$ using Taylor´s theorem until second order around $\kappa$ would be:
+
+$$CS_a \approx l_{a-\kappa} \int_{\alpha}^a{m_{x}dx}+(l_{a-\kappa})^{'}\int_{0}^a{(x-\kappa)m_{x}dx}+
+(l_{a-\kappa})^{''}\int_{0}^a{\frac{(x-\kappa)^2}{2} m_{x}dx}\\
+  \approx F_a l_{a-\kappa} + \frac{\sigma^2}{2}F_a (l_{a-\kappa})^{''}\\
+  \approx F_a \, l_{a-\kappa}[1+\frac{\sigma^2}{2}\frac{(l_{a-\kappa})^{''}}{l_{a-\kappa}}]$$
+
+The second Taylor´s term is null because $\int_{\alpha}^a{x \, m_{x}dx} = \kappa \, F_a$.
+The formula tells us that the survival experience for a woman aged *a* depends mainly of cumulated fertility (possible cases, different from the ancestor count case where this doesn´t affect) and child survival probability (success cases), but also negative proportional affected by how disperse is the fertility by age (variance \sigma^2) and the suvival curvature around (negative sign), in a $l_x$ range where is tipycally very flat (range 20-40 years old) in transitioned populations.
+Guatemala improved the approximation with years due to rectangularization process in $l_x$ (figure \@ref(fig:plot_CS_aprox)).
+
+```{r plot_CS_aprox, fig.cap = "Error in approximation for a=40, from years 1950 (darker) to 2015 (lighter). Guatemala in red (left). Guatemala survival and fertility by age (right)"}
+a <- ggplot(CS_outs %>% filter(Age == 40), 
+                    aes(x=CS, y=error_aprox, color=Year))+
                     geom_point(alpha = .6) + 
+                    geom_point(data = CS_outs %>% filter(name=="Guatemala",Age == 50),
+                               aes(x=CS, y=error_aprox), color="red") +
                     theme_classic() +
@@
                     theme_classic() +
+                    scale_color_grey()+
                     theme(legend.position = "none") +
@@
 
-The expected time in years that a mother aged 30 passed with a death son was around 4% in some countries at middle XX Century.
+### Burden and timing
+
+We call *Mean Time Spent in Bereavement* (*MTSB*) the absolute measure of the expected total lifetime with a death daughter for a mother aged *a*, which can be expressed in terms of a temporary expected lost years index in line with $e^\dagger$ (@Vaupel1986):
+
+$$MTSB_a = \int_0^a{m_x\int_0^{a-x}{d_t e_{0|a-x-t}\, dt}\,dx} = \int_\alpha^a{m_x e_{0|a-x}^\dagger \,dx}$$.
+
+Where $d_t$ is the death distribution from birth, $e_{0|a-x-t}$ is the life expectancy at birth until age $a-x-t$ and $e_{0|a-x}^\dagger$ the temporary dispertion measure.
+But most interesting would be to compare these years with the time that mothers would expect to live with their daughters.
+We call this the *Intensity Time in Bereavement* ($ITB$)): a ratio between expected time with a "lost" and expected time with a "life", that allows to make comparisons between population regimes.
+
+$$ITL_a = \frac{\int_0^a{m_x e_{0|{a-x}}^\dagger dx}}{\int_0^a{m_x e_{0|a-x} dx}}$$
+
+This is a ratio between child-years in two radically different states.
+Looks similar to the trascendental entropy measure $H$ (@KeyfitzMath) but considering all the cohorts born during the the mother´s life, weighted by their relative size $m_x$.
+In the figure \@ref(fig:plot_ITL_MAL) is shown that *ITB* is bigger for youg womans because of the weight of infant mortality in their mother experience, also with more dispersion for same levels of parity.
+Assuming constant period rates the expected time in years that a mother aged 30 passed with a death son was around 4% in some countries at middle XX Century.
 When increasing age, the survival experience depends less on infant mortality, and the distribution is around less than 2% for womans aged 50, converging to 0 on time.
+As extreme cases, in 1950-1955 Haiti women aged 30 would have experienced an intensity of `r round(CS_outs$ITL[CS_outs$name=="Haiti" & CS_outs$Age==30 & CS_outs$Period=="1950-1955"],1)`%, and in 2010-2015 `r round(CS_outs$ITL[CS_outs$name=="Haiti" &CS_outs$Age==30 & CS_outs$Period=="2010-2015"],1)`%, while the women of Costa Rica `r round(CS_outs$ITL[CS_outs$name=="Costa Rica" & CS_outs$Age==30 & CS_outs$Period=="1950-1955"],1)`%, and in 2010-2015 `r round(CS_outs$ITL[CS_outs$name=="Costa Rica" &CS_outs$Age==30 & CS_outs$Period=="2010-2015"],1)`%.
 
-```{r plot_ITL, fig.cap = "Intensity Tome Lost of  womans aged 30, 40 and 50. Years 1950 (darker) to 2015 (lighter)"}
-ggplot(CS_outs)+  geom_point(aes(x=Fa, y=ITL, color=Year), alpha = .3) +
+Another important factor for the child survival experience of mothers, is the mean age at child loss, called here $MAL$.
+This relation can be derived by starting with the mother age $x+t$ at each death child age $t$ at death, weighted by the fertility and survival function.
+In it, $\kappa$ is the mean age at childbirth for womens aged $a$, $MAD_{a-x}$ refers to the mean age at death for newborns that die before $a-x$, $F_a$ is the cumulated fertility for a women aged $a$:
+
+$$MAL_a = \frac{\int_0^a{m_x {\frac{\int_0^{a-x} l_t \, \mu_t (x+t) dt}{\int_0^{a-x} l_t \, \mu_t dt}}dx}}{\int_0^a{m_x}dx} \\
+  MAL_a = \frac{\int_0^a{m_x \left[{x+\frac{\int_0^{a-x} l_t \, \mu_t \, t \, dt}{\int_0^{a-x} l_t \, \mu_t dt}}\right]dx}}{\int_0^a{m_x}} \\
+  MAL_a = \frac{\int_0^a{m_x \,x \,dx}}{F_a} + \frac{\int_0^a{m_x MAD_{a-x}dx}}{F_a} \\
+  MAL_a = \kappa + \frac{\int_0^a{m_x MAD_{a-x}dx}}{F_a}$$
+
+For those latinamerican mothers aged 50 with 3 daughters born who suffered a lost, theyy experienced that at age 30 in average (figure \@ref(fig:plot_ITL_MAL)) (More can be analyzed here).
+
+```{r plot_ITL_MAL, fig.cap = "Intensity Tome Lost of  womans aged 30, 40 and 50. Years 1950 (darker) to 2015 (lighter). Latinamerican countries in period 1950-2015"}
+ITL <- ggplot(CS_outs)+  
+                    geom_smooth(method="lm", aes(x=Fa, y=ITL, color=factor(Age)), 
+                                se = F, formula= (y ~ exp(x)))+
+                    geom_point(aes(x=Fa, y=ITL, color=factor(Age)), alpha = .3) +
                     theme_classic()+        
-                  theme(legend.position = "none") +
-                  labs(y = "%")+
-                  facet_grid(~Age) 
-```
+                    theme(legend.position = "right") +
+                    labs(y = "%")
 
-For those woman that lost a daughter, more fertility more the mean age.
-
-```{r plot_MAL}
-ggplot(CS_outs %>% filter(Age==50))+
+MAL <- ggplot(CS_outs %>% filter(Age==50))+
                     geom_smooth(aes(x=Fa, y=MAL), method = "lm", col="grey", se = F)+
@@
                     labs(y = "Mean Age at Lost")
+grid.arrange(ITL, MAL, ncol=2)
+```
+
+### Mortality changes
+
+We now consider the consequences of an absolute change $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk.
+Considering that $m_{x,\delta}=m_{x}+\delta$ (@wrycza2012), and $l_{a-x}^\delta = e^{-\int_{0}^{a-x}{(\mu_t+\delta})dt}$:
+
+$$CS_{a}^\delta = \int_{\alpha}^{a}m_{x} l_{a-x} e^{-\delta (a-x)} dx$$
+
+We get the derivative of $dCS_{(a)}^\delta / d\delta$ evaluated near zero (@KeyfitzMath) to find the effects of adding $\delta$ to hazard rates:
+
+$$\frac{dCS^{\delta}}{d\delta} = -a\int_{\alpha}^{a}m_{x}{l_{a-x} dx} + \int_{\alpha}^{a} m_{x} {l_{a-x}x dx}\\
+= -a \, CS_{a}  + \int_{\alpha}^{a} {x  \, m_{x} l_{a-x} dx}$$
+
+Dividing both sides by $CS_{a}$ in an discrete approzimation we get
+
+$$\frac{\Delta CS_{a}}{CS_{a}} \approx -(a - k_a) \Delta\delta$$
+
+The expected change in descendants survival is inversely proportional to the difference between maternal age $a$ and the mean age of the mother at the birth of her surviving daughters $k_a$ (always smaller than $\kappa_a$).
+The magnitude of the change depends negatively on the age distribution of the surviving offspring.
+This is intuitive considering that younger descendants experiences longer periods of exposure to risk.
+The figure \@ref(fig:CS_abs_app) shows that goodness of fit is descreasing with the change size, given that $\delta$ is assumed near zero (as an example considering $\delta=0.01$ means at age 50 that $\frac{l_50}{e^{\delta\,50}} = \frac{l_50}{1.6}$, a big change).
+
+```{r CS_abs_app, fig.cap= "Absolute change and its approximation for age CS in women aged 50. Latinamerican countries in period 1950-2015"}
+# application of absolute changes
+age = 50
+CS_abs_change_app = data.frame(delta = NA,Country = NA, Period = NA, change = 0, change_app = 0)
+for(delta in seq(0,.01,.001)){
+  for (country in unique(LA$name)){
+    for (period in unique(LA$Period)){
+      # country = "Argentina"; period = "2010-2015"; delta = .01
+      change_app = -(age - CS_outs %>% filter(name == country, 
+                                              Period == period, 
+                                              Age == age) %>% select(k)) * delta
+      country_i = LA %>% filter(name == country, Period == period) 
+      CS = CS_outs %>% filter(name == country, 
+                              Period == period, 
+                              Age == age) %>% select(CS)
+      CS_changed = sum(unlist(sapply(1:age, function(i) 
+        country_i$asfr[i]*country_i$lx[age-(i-1)]/100000*exp(-delta*(age-(i-1))))), 
+        na.rm = T)
+      change = (CS_changed-CS)/CS
+      CS_abs_change_app = rbind(CS_abs_change_app, 
+                                data.frame(delta = delta,
+                                           Country = country, Period = period, 
+                                           change = as.numeric(change), 
+                                           change_app = as.numeric(change_app)))
+    }
+  } 
+}
+CS_abs_change_app %>% slice(-1) %>% 
+  ggplot() + 
+  geom_point(aes(x=change, y=change_app, color=delta)) +
+  geom_abline(slope=1, intercept = 0) +
+  theme_classic() +
+  # theme(legend.position = "none")+
+  xlim(-.3, 0) + ylim(-.3, 0)
+```
+
+That an equal change in mortality could happens at all ages is unlikelihood. As a [corollary](#Corollary) we inspect the case thtat only infant mortality is affected.
+
+The relationship is less clear for a proportional change given the interaction of birth and mortality rates (@KeyfitzMath).
+Considering that $\mu_{x,\delta}=\mu_{x}(1+\delta)$ and $l_{a-x}^\delta = e^{-\int_{0}^{a-x}{\mu_t(1+\delta})dt}=(l_{a-x})^{(1+\delta)}$:
+
+$$CS_{a}^\delta = \int_{\alpha}^{a} {m_{x} l_{a-x}^{(1+\delta)}} dx$$
+
+Using the derivative $\frac{dl_{a-x}^{(1+\delta)}}{d\delta} = log(l_{a-x}) l_{a-x}^{(1+\delta)}$, and in the third row reversing integrals between $t$ and $x$:
+
+$$\frac{d CS_{a}^\delta}{d \delta} = \int_{\alpha}^{a} {m_{x} l_{a-x} \log(l_{a-x}) dx}\\
+= - \int_{0}^{a} {m_{x} l_{a-x} \int_{0}^{a-x}{\mu_x \, dt}\, dx}\\
+= - \int_{0}^{a}{\mu_t} \int_{0}^{a-t} {m_{x} l_{a-t-x} \, dx}\, dt$$
+
+Dividing by $CS_a$ and multiplying by $\delta$ get one useful expression.
+This is the negative of cumulative hazard $H_a$ but considering a factor $0\leq \frac{CS_{a-t}}{CS_a}\leq1$ that takes in acccount the amount of survinvig descendatns that would be lost for a mother aged *a* in each age x of risk.
+
+$$\frac{\Delta CS_{a}}{CS_a} = - \left[\int_{0}^{a}{\mu_t  \, \frac{CS_{a-t}}{CS_a} \,dt \,}\right] \Delta \delta$$
+
+This factor has a S-shape because $l_x$ curvature fertility acumulation, that gives more weight to first ages, and which is grater than 1 only for ages *a* with complete reproductive life and high infant mortality (with no aditional childs and higher risk for one age additional of survive).
+
+<!---
+An Application here
+-->
+
+```{r chek_CS_ratio, include = F}
+# test factor
+countries = c("Guatemala", "Argentina", "Haiti")
+period = "1950-1955"
+Age = 40
+plot(NULL, xlim=c(15,Age), ylim=c(0,1.1),xlab="Age", ylab="Ratio")
+abline(h=1, lty=2, col="grey")
+for (country in countries){
+  points(15:Age,
+         sapply(15:Age, function(age)
+         CS_results(country = country, 
+             period = period, 
+             a = age)$CS)/
+          CS_results(country = country, 
+             period = period, 
+             a = Age)$CS,
+          col = which(country==countries), t="l")
+}
+```
+
+### Heterogeneity
+
+In a model with heterogeneity, determined at born with a multiplicative effect, $CS_a$ could be interpreted as a conditional expectation with random variables *K* for fertility heterogeneity and *Z* for mortality frailty.
+
+$$CS_a(k,z) = \int_{0}^{\infty}\,{m(k)}_x\,{l(z)}_{a-x}\,dx$$
+
+Following @Coresh1988 we can express the fertility part in a multiplicative way with variability in the level but not in the shape, as ${m(k)}_x = F_\beta\,k\,r_x$, where $F_\beta$ is the baseline cumulated fertility, $K$ is a random variable with mean 1 that allows variability between groups and $r_x$ is the fertility structure by age ($\int_{0}^{\beta}{r_x}=1$).
+Frailty part can be thought in a cohort effect way, also in the multiplicative assumption as @Vaupel2014, with $l(z)_x=e^{-H_{x}Z}=l_x^Z$, with baseline hazard $\mu$.
+Replacing conveniently the last two expressions and considering the joint distribution $f_{kz}=f_{z/k} \,f_k$, the mean $\overline{CS}_a$ would be:
+
+$$\overline{CS}_a = \int_{0}^{\infty}\int_{0}^{\infty}\left[\int_{0}^{a}{ {m(k)}_x l(z)_{a-x}\,dx}\right]f_{kz}\,dz \,dk\\
+= F_\beta \int_{0}^{a-x} r_x \int_{0}^{\infty} \int_{0}^{\infty} {l_{a-x}}^z \, k\, f_{kz} \,dk\,dz\,dx$$
+
+The way to chain dependence is convenient to the expected herarchy $f_{zk}=f_{k|z}f_z$: groups with higher descendant´s mortality would adjust their fertility level, with positive correlation.
+Isolating one age and defining $f_{kz}$ we can explore effects in $CS_a$.
+
+<!---
+More development here
+-->
+
+```{r, include=F}
+# frailty distributions
+db = LA %>% filter(x==10) %>% select(q,tfr)
+cor.test(LA$q[LA$x==10],LA$tfr[LA$x==10])
+plot(db)
+model = lm(tfr~log(q), db)
+lines(seq(0,0.4,0.01), predict(model,newdata = data.frame(q=seq(0,0.4,0.01))))
 ```
@@
 
-# pendings
+## Appendix
 
-Here an independet assumption is done in the relationship between fertility and descendants survival within population. Both assumptions will be considered at the end.
+### Data {#Data_appendix}
 
-In terms of mortality, this measure is much more related to first ages (child and youth) than adult or adult ages, so we can model a H-P pattern and derivate some of their parameters...
+Mortality and fertility data was obtained from the last revision of population prospects by UN (@WPP19).
+We smoothed female $l_x$ using cubic-splines constrained to monothonic descense, taking later $L_0$ and $T_{100}$ as inputs for year-person calculations.
+For splitting fertility five groups was used quadratic optimization approach by @Michalski2018, with an desirable property for our purpose which is a good fitting in parity.
+Also was assumed an unique female percentage of newborns of `r round(im, 4)` for all period-country cases.
 
-We ca take cohort rates to compare between countries, relaxing main assumptions in stable model...
+An negative relation between mortality until 10 and TFR is shown in \@ref(fig:plot_tfr_q010).
+Specific paths like Argentina and Guatemala shows very different demographic profiles during second part of XX century.
 
-Is known that fertility is influenced by child mortality, and that there is an heterogeneus behaviour within.
-We can add frailty variables to take account of that and express correlation between subgroups behaviour...
-
-```{r relative_change, include = F, echo=F, fig.align='center', warning=F, out.width = '70%', fig.cap="Two different fertility schedules (solid lines) and a big mortality hump (right axis, dashed line).
-The vertical lines show the mean age at birth for each schedule."}
-# 
-# # genero m_x (old and young pattern) y l_x (with big hump) 
-# f_young = data.frame(x=c(15,20,25,30,35,40), f=c(0,.3,.3,.05,.01,.01))
-# f_old = data.frame(  x=c(15,20,25,30,35,40), f=c(0,.04,.06,.25,.1,.01))
-# f_young_p = predict(loess(f~x, data = f_young), 15:40)
-# f_old_p = predict(loess(f~x, data = f_old), 15:40)
-# f_young_mean = sum(f_young_p*15:40/sum(f_young_p))
-# f_old_mean = sum(f_old_p*15:40/sum(f_old_p))
-# m_si_hump = data.frame(x= c(0,1,5,10,15,20,25), l=c(1,.98,.97,.97,.96,.9,.85))
-# m_si_hump_sp = splinefun(m_si_hump, method = "monoH.FC")
-# 
-# #graph
-# library(latex2exp)
-# par(mfrow=c(1,1),mar = c(5,5,2,5))
-# plot(15:40, f_young_p, t="l",xlab="x (mother age)", ylab= TeX(sprintf("$m_x$"))) 
-# lines(15:40, f_old_p, lty=1, col=2) 
-# abline(v=f_young_mean, lty=2, col="grey")
-# abline(v=f_old_mean, lty=2, col="grey")
-# par(new = T)
-# plot(0:25, rev(m_si_hump_sp(0:25)), axes=F, xlab=NA, ylab=NA, t="l", col=1, lty=2)
-# axis(side = 4)
-# mtext(side = 4, line = 3, TeX(sprintf("$l_{40-x}$")))
-# 
-# #calculate entropy CS
-# H_young = -sum(f_young_p*rev(m_si_hump_sp(0:25))*log(rev(m_si_hump_sp(0:25))))/
-#   sum(f_young_p*rev(m_si_hump_sp(0:25)))
-# H_old = -sum(f_old_p*rev(m_si_hump_sp(0:25))*log(rev(m_si_hump_sp(0:25))))/
-#   sum(f_old_p*rev(m_si_hump_sp(0:25)))
-
+```{r plot_tfr_q010, fig.cap="Female probability of death in first 10 years old by gross reproduction rate. Latin America countries in period 1950-2015"}
+ggplot(LA %>% filter(x==10), aes(x=tfr * 2.04, y=q))+
+                  geom_point(aes(color=Period)) + 
+                  geom_smooth(se = F, color = "grey", method = "lm", 
+                              formula = y ~ exp(x)) +
+                  geom_line(data = LA %>% filter(x==10, name %in% c("Guatemala")),
+                            aes(group=name), colour=2, size = 1) +
+                  geom_line(data = LA %>% filter(x==10, name %in% c("Argentina")),
+                            aes(group=name), colour=4, size = 1) +
+                  theme_classic() + labs(x="GRR", y="q(0,10)")
 ```
 
-## 5. Bibliography
+### Corollary {#Corollary}
+
+That an equal change in mortality could happens at all ages is unlikelihood.
+This measure is much more related to first ages (child and youth) than adult or ones.
+As a corollary, if the change only happens in the age range $\left[0;1\right)$, infant mortality, we can inspect the effect starting with splitting the integral (of course, for $a>1$)]:
+
+$$CS_{a}^{\delta_0} = \int_{0}^{a-1}
+m_{x} {e{^{-\int_{1}^{a-x}\mu_t \,dt-\int_{0}^{1}(\mu_t + \delta)\,dt}}} dx + 
+\int_{a-1}^{a} m_{x} e^{-\int_{0}^{a-x}(\mu_t+ \delta)\,dt}\\
+=\int_{0}^{a-1} m_{x} l_{a-x}e^{-\delta} dx + \int_{a-1}^{a} m_{x} l_{a-x} e^{-\delta(a-x)}\,dx$$
+
+Deriving by $\delta$ and valuating near 0, we get:
+
+$$CS_{a}^{\delta_0} =-\int_{0}^{a-1} m_{x} l_{a-x} dx - a \int_{a-1}^{a} {m_{x} l_{a-x} dx} + \int_{a-1}^{a} {x \, m_{x} l_{a-x} dx}$$
+If we name $CS_{a-1,a}=\int_{a-1}^{a} {m_{x} l_{a-x} dx}$, we can express the first term in the right as $CS_a - CS_{a-1,a}$, and finally divide by $CS_a$: 
+$$ = - CS_a + CS_{a-1,a} (1-a+\kappa_{a-1,a})\\
+\frac{\Delta CS_{a}}{CS_{a}} \approx - \left[1-\frac{CS_{a-1,a}}{CS_a} (1-(a-\kappa_{a-1,a}))\right] \Delta\delta$$
+
+This means that an absolute change in infant hazard rate affects proportionally for all the age range extension (because of all the cohorts aged more than 1), except for the mean time exposure for those who born and survive in the last year $1-(a-k_{a-1,a})$, times $\frac{CS_{a-1,a}}{CS_a}$ the portion of cumulated alive daughters for the same period, weighting how important was this in total succesful experience (in terms of alive descendants).
+
+## References

### 2020-04-23 R/garbage.R

@@ -1,53 +1,0 @@
-
-library(HMDHFDplus)
-user = "ivanwilliams1985@gmail.com"
-pwHMD ="volveroman"
-pwHFD = 52962
-
-Austria_F <- readHFDweb("AUT","asfrRR", user, pwHFD)
-Austria_M <- readHMDweb("AUT","fltper_1x1", user, pwHMD)
-alpha = 15
-a = 40
-range = a - alpha
-year = 1951
-Austria_F2017 <- subset(Austria_F, Year==year & Age %in% alpha:a)
-Austria_M2017 <- subset(Austria_M, Year==year & Age %in% 0:range)
-plot(Austria_F2017$Age, Austria_F2017$ASFR, t="l")
-par(new = T)
-plot(Austria_M2017$Age, rev(Austria_M2017$lx)/100000, ylim = c(.9,1), 
-     col=2, axes=F, xlab=NA, ylab=NA, t="l")
-
-
-
-# theorical cases
-
-# big hump case
-
-#input
-par(mfrow=c(1,1),mar = c(5,5,2,5))
-f_young = data.frame(x=c(15,20,25,30,35,40), f=c(0,.3,.3,.05,.01,.01))
-f_old = data.frame(  x=c(15,20,25,30,35,40), f=c(0,.04,.06,.25,.1,.01))
-f_young_p = predict(loess(f~x, data = f_young), 15:40); sum(f_young_p)
-f_old_p = predict(loess(f~x, data = f_old), 15:40); sum(f_old_p)
-f_young_mean = sum(f_young_p*15:40/sum(f_young_p)); f_young_mean
-f_old_mean = sum(f_old_p*15:40/sum(f_old_p)); f_old_mean
-library(latex2exp)
-m_si_hump = data.frame(x= c(0,1,5,10,15,20,25), l=c(1,.98,.97,.97,.96,.9,.85))
-m_si_hump_sp = splinefun(m_si_hump, method = "monoH.FC")
-#graph
-plot(15:40, f_young_p, 
-     t="l",xlab="x (mother age)", ylab= TeX(sprintf("$m_x$"))) 
-lines(15:40, f_old_p, lty=2) 
-abline(v=f_young_mean, lty=2, col="grey")
-abline(v=f_old_mean, lty=2, col="grey")
-par(new = T)
[…]

### 2020-04-28 ALAP2020/ALAP.Rmd
worked on intro

@@
 
-Kin count estimation has a long pedigree in mathematical demography.
-@lotka1931orphanhood proposed to model the orphanhood in a theorical population, comparing the effect across demographic regimes.
-Equation  \ref{eq:CS} was originally proposed by @brass_derivation_1953 to estimate child mortality.
+Kin count estimation has a long and distinguished pedigree in mathematical demography, starting with the work of @lotka1931orphanhood on modelling orphanhood in theorical populations across demographic regimes.
+To the best of our knowledge, @brass_derivation_1953 first proposed an equation akin to Eq.
+\ref{eq:CS} to estimate child survival over age.
 The notion of estimating the expected number of living daughters in a stable population was generalized by @goodman1974 for other kin relations (granddaughters, cousins, etc.).
-The "counting method" approach, being the net reproduction rate $R_0$ limited to $a$, or the (not complete) ratio between generations (multiplying and dividing by the mother generation $B$), was further popularized by @KeyfitzMath.
-@bongaarts_projection_1987 used a similar approach to estimate descendants in his 'Family Status Model'.
-More recently, @wrycza2012 has looked at the formal effect of different kind of changes in mortality by age.
+<!-- We should clarify this sentence a bit: -->
+The "counting method" approach was further popularized in @KeyfitzMath, where the net reproduction rate $R_0$ was limited to $a$ or to the (not complete) ratio between generations (multiplying and dividing by the mother generation $B$).
+@bongaarts_projection_1987 used a similar approach to estimate descendants in his 'Family Status Model' and, more recently, @wrycza2012 has used formal techniques to describe the effect of different kind of changes in mortality by age.
 
-<!---
-Improve Intro: ¿Why is important study survival descendants from mother perspective?¿How impacts mothers health/welfare/experience?
--->
+The question of child survival, particularly the survival of daughters for mothers, sits at the very center of demographic theory.
+This is evidenced by the omnipresensce of the 'net reproductive rate' in accounts of human and non-human populations. 
+Historical demographers draw liberally on assumptions about kin availability and individual's exposure to offspring mortality to explain fertility decisions, especially around the demographic transition.
+These assumptions are often untested given the data sparcity [@livi_bacci_concise_1997;@volk_infant_2013]. 
+Economic theories of fertility that consider children as 'investment' for old-age support also rely on notions of child survival [@preston1978effects].
 
-The objective of this work is to give a formal approach to deomgraphy of surviving childs from mother perspective, looking for relations to characterize this situation: the timing of descedent mortality (mean age at the event and time spent in that state) and the effect of mortality change in the final descendants quantity.
-Numerical applications of the expressions will be done for Latinoamerican countries during the second half of XX Century.
-Details on data handling and smooth procedures are exposed in [data section](#Data_appendix) in the Appendix.
+Child survival - and it's counterpart, child death - gain new significance in the wake of the fertility transition.
+Depressed fertility means that fewer children are tasked with providing key emotional, social, and financial transfers to aging parents for ever increasing periods of time.
+With increasing periods of generational overlapp, individuals find themselves 'sandwiched' between aging parents and young children requiring their simultaneous attention and care [@daatland_norwegian_2010].
+In the context of global population aging, elderly parents without access to formal social security and pension systems are particularly reliant on these transfers to make ends meet [@smith-greenaway_maternal_2020].
+Losing an only child may become a more common experience in the context of global fertility decline, something particularly worrying for parents suddendly rendered childless.
+Given the known psychological, health, and social consequences of child loss for bereaved parents and families [@Hendrickson2009;@lee_death_2014], it is surprising that the demographic processes that shape parental bereavement remain very poorly understood.
+
+How can we characterize the timing of offspring mortality as experienced by a mother using a formal demographic perspective? 
+How do changes in mortality affect the availability of offspring over age from the point of view of a prospective mother?
+In this paper, we uncover formal relationships to answer these questions and provide numerical applications using data from countries in Latin America during the second half of the twentieth century.
+Please see the [data section](#Data_appendix) in the Appendix for details on the data and analytical procedures.
 

### 2020-04-28 ALAP2020/ALAP.Rmd
worked on intro

@@
 
-Kin count estimation has a long and distinguished pedigree in mathematical demography, starting with the work of @lotka1931orphanhood on modelling orphanhood in theorical populations across demographic regimes.
+Kin count estimation has a long and distinguished pedigree in mathematical demography, starting with the work of @lotka1931orphanhood on modelling orphanhood in theoretical populations across demographic regimes.
 To the best of our knowledge, @brass_derivation_1953 first proposed an equation akin to Eq.
@@
 The question of child survival, particularly the survival of daughters for mothers, sits at the very center of demographic theory.
-This is evidenced by the omnipresensce of the 'net reproductive rate' in accounts of human and non-human populations. 
+This is evidenced by the omnipresence of the 'net reproductive rate' in accounts of human and non-human populations. 
 Historical demographers draw liberally on assumptions about kin availability and individual's exposure to offspring mortality to explain fertility decisions, especially around the demographic transition.
-These assumptions are often untested given the data sparcity [@livi_bacci_concise_1997;@volk_infant_2013]. 
+These assumptions are often untested given the data sparsity [@livi_bacci_concise_1997;@volk_infant_2013]. 
 Economic theories of fertility that consider children as 'investment' for old-age support also rely on notions of child survival [@preston1978effects].
@@
 Depressed fertility means that fewer children are tasked with providing key emotional, social, and financial transfers to aging parents for ever increasing periods of time.
-With increasing periods of generational overlapp, individuals find themselves 'sandwiched' between aging parents and young children requiring their simultaneous attention and care [@daatland_norwegian_2010].
+With increasing periods of generational overlap, individuals find themselves 'sandwiched' between aging parents and young children requiring their simultaneous attention and care [@daatland_norwegian_2010].
 In the context of global population aging, elderly parents without access to formal social security and pension systems are particularly reliant on these transfers to make ends meet [@smith-greenaway_maternal_2020].
-Losing an only child may become a more common experience in the context of global fertility decline, something particularly worrying for parents suddendly rendered childless.
+Losing an only child may become a more common experience in the context of global fertility decline, something particularly worrying for parents suddenly rendered childless.
 Given the known psychological, health, and social consequences of child loss for bereaved parents and families [@Hendrickson2009;@lee_death_2014], it is surprising that the demographic processes that shape parental bereavement remain very poorly understood.
 
-How can we characterize the timing of offspring mortality as experienced by a mother using a formal demographic perspective? 
-How do changes in mortality affect the availability of offspring over age from the point of view of a prospective mother?
+How can we characterize the timing of offspring mortality as experienced by a mother from a formal demographic perspective? 
+<!-- How do changes in mortality affect the availability of offspring over age from the point of view of a prospective mother? -->
+How do changing mortality rates affect the availability of offspring over age for mothers, if fertility remains constant?
 In this paper, we uncover formal relationships to answer these questions and provide numerical applications using data from countries in Latin America during the second half of the twentieth century.
-Please see the [data section](#Data_appendix) in the Appendix for details on the data and analytical procedures.
+Please see the [data section](#Data_appendix) in the Appendix for more details on the data and on the analytical procedures.
 
-## Formulas
+## Formal development
 
-$CS_a$ is the expected number of surviving children to a mother^[We´ll refer to mothers, assuming that fertility rates applies to all woman, meaning that all of them are exposed.] alive at aged $a$ in a female stable population with fertility rates $m_{x}$, mortality hazard $\mu_{x}$ and survival function $l_x=e^{-\int_{0}^{x}{\mu_t}\,dt}$ ($l_0=1$ during this work), as proposed by @goodman1974:
+Let $CS_a$ be the expected number of surviving children to a mother^[We use woman and mother interchangeably in this paper, assuming that all women are exposed to the same fertility rates.] alive at aged $a$ in a female stable population with fertility rates $m_{x}$, mortality hazard $\mu_{x}$ and survival function $l_x=e^{-\int_{0}^{x}{\mu_t}\,dt}$ (with unit radix $l_0=1$), as proposed by @goodman1974:
 
-$$CS_a = \int_{0}^a{m_{x}l_{a-x}dx}$$
+$$CS_a = \int_{0}^a{m_{x}l_{a-x}dx}.$$
 
-Between 1950-1955 and 2010-2015 the portion of surviving daugthers  $\frac{CS_a}{F_a}$ at age 50 passed from a simple mean of `r mean(CS_outs$CS_prob[CS_outs$Period == "1950-1955" & CS_outs$Age==50])` to `r mean(CS_outs$CS_prob[CS_outs$Period == "2010-2015" & CS_outs$Age==50])`, also with a decrease of `r mean(CS_outs$CS[CS_outs$Period == "1950-1955" & CS_outs$Age==50])-mean(CS_outs$CS[CS_outs$Period == "2010-2015" & CS_outs$Age==50])` in the expected descendants at same age.
-Althought this is not a  real measure but a theorical one if the rates would have been kept constant during enough time.
+<!-- Que te parece algo asi? -->
 
-<!---
-Interpretation
--->
+We motivate this paper with an empirical example using fertility and mortality rates from the 2019 Revision of the UN World Population Prospects (UN WPP) for the Latin American Region. 
+The example makes the simplifying assumption that the UN WPP period rates remain in place during sufficient time for the populations to approach demographic stability.
+We hope that this inaccuracy will be forgiven for the sake of simplicity. 
+A more sophisticated model using the appropriate age-specific cohort rates in the subscripts has been proposed by @alburez_gutierrez_womens_2019.
 
-```{r plot_CS, fig.cap="Child Survival and Child survival related to cumulated fertility, for women aged 30, 40 and 50. Latinamerican countries in period 1950-2015"}
+Consider a Latin American woman standing before us.
+If this were 1950-1955, she could reasonable expect to have `r mean(CS_outs$CS[CS_outs$Period == "1950-1955" & CS_outs$Age==50])` surviving children on her 50$^{th}$ birthday.
+In 2010-2015 a woman the same age would only have `r mean(CS_outs$CS[CS_outs$Period == "2010-2015" & CS_outs$Age==50])` living children. 
+The difference of `r mean(CS_outs$CS[CS_outs$Period == "1950-1955" & CS_outs$Age==50])-mean(CS_outs$CS[CS_outs$Period == "2010-2015" & CS_outs$Age==50])` children is explained by reduced fertility and improved mortality in the region [give stats].
+We now remove the effect of changing fertility by considering the number of daughters surviving up to maternal age $a$ as a proportion of the daughters ever born to a woman that age, $\frac{CS_a}{F_a}$.
+Given Eq.
+\ref{eq:CS}, the increase from `r mean(CS_outs$CS_prob[CS_outs$Period == "1950-1955" & CS_outs$Age==50])` in 1950-1955 to `r mean(CS_outs$CS_prob[CS_outs$Period == "2010-2015" & CS_outs$Age==50])` in 2010-2015 must be explained by a change in mortality.
+In the remainder of the paper, we aim to formalize the relationship population-level changes in mortality rates and perceived change in the lived experience of child death from the perspective of a mother.
+We consider the case when these changes are additive, mutliplicative, and when they are not evenly distributed across age.
+
+<!-- Add a line for TFR in plot? -->
+<!-- Por que no llegas hasta 2020? -->
+
+```{r plot_CS, fig.cap="Child Survival and Child survival as a share of cummulative fertility by age, for women aged 30, 40 and 50.
+Estimates using UN WPP data for Latin American countries in the period 1950-2015 period"}
 a <- ggplot(CS_outs, aes(x=Year, y=CS, color=factor(Age)))+
@@
 
-### An approximation
+### An approximation of $CS_a$, the expected number of surviving daughters
 
-Following @KeyfitzMath work about living mothers count, we can consider $a>=\beta$ and assume that fertility pattern by age is concentrated around $\kappa$ (mean age at childbearing), and cumulated fertility (or gross reproduction rate because of female model) is $F_a =  \int_{\alpha}^a{m_{x}dx}$.
-An approximation of $l_x$ using Taylor´s theorem until second order around $\kappa$ would be:
+<!-- QUe te parece esta reorganizacion de esta seccion? -->
+
+As an initial step, we seek an intuitive understanding of the effect of changing mortality on child survival. 
+Building on work by @KeyfitzMath for the probability of a living mother, we show which features of the child survival function explain the expected number of children. 
+For this we use an approximate of $l_x$ using Taylor´s theorem until second order around the mean age of childbearing $\kappa$ for all $a>=\alpha$:
 
@@
 
+where the fertility pattern by age is concentrated around $\kappa$ and the accumulated fertility (or gross reproduction rate in our same-sex model) is $F_a =  \int_{\alpha}^a{m_{x}dx}$. 
 The second Taylor´s term is null because $\int_{\alpha}^a{x \, m_{x}dx} = \kappa \, F_a$.
-The formula tells us that the survival experience for a woman aged *a* depends mainly of cumulated fertility (possible cases, different from the ancestor count case where this doesn´t affect) and child survival probability (success cases), but also negative proportional affected by how disperse is the fertility by age (variance \sigma^2) and the suvival curvature around (negative sign), in a $l_x$ range where is tipycally very flat (range 20-40 years old) in transitioned populations.
+
+We find that, seen from the perspective of a mother, child survival mainly depends on the cummulative fertility function^[This runs contrary to the findings of @KeyfitzMath, who found that levels of fertility did not affect the probability of having living ancestors.] and the mean age at childbearing $\kappa$ acts through the factor $l_{a-\kappa}$.
+The dispersion of of fertility $\sigma^2$ is negatively proportional to child survival. 
+All else constant, a higher dispersion of fertility $\sigma^2$ is associated with a decrease in $CS_a$ during reproductive age.
+The negative relationship stems from the fact that the second derivative of $l_{a-\kappa}$ is negative in the 15-49 age range, given a concave downwards survival function.
+
+<!-- No cambie esto porque no me queda tan claro cual es la conexion con el resto del texto - podemos discutir -->
+
 Guatemala improved the approximation with years due to rectangularization process in $l_x$ (figure \@ref(fig:plot_CS_aprox)).
 
-```{r plot_CS_aprox, fig.cap = "Error in approximation for a=40, from years 1950 (darker) to 2015 (lighter). Guatemala in red (left). Guatemala survival and fertility by age (right)"}
+```{r plot_CS_aprox, fig.cap = "Left: Error in approximation for a=40, from years 1950 (darker) to 2015 (lighter) for all Latin American countries (blue) and Guatemala (red).
+Right: Change in survival and fertility by age in Guatemala."}
 a <- ggplot(CS_outs %>% filter(Age == 40), 
@@
 
-Where $d_t$ is the death distribution from birth, $e_{0|a-x-t}$ is the life expectancy at birth until age $a-x-t$ and $e_{0|a-x}^\dagger$ the temporary dispertion measure.
+Where $d_t$ is the death distribution from birth, $e_{0|a-x-t}$ is the life expectancy at birth until age $a-x-t$ and $e_{0|a-x}^\dagger$ the temporary dispersion measure.
 But most interesting would be to compare these years with the time that mothers would expect to live with their daughters.
@@
 This is a ratio between child-years in two radically different states.
-Looks similar to the trascendental entropy measure $H$ (@KeyfitzMath) but considering all the cohorts born during the the mother´s life, weighted by their relative size $m_x$.
-In the figure \@ref(fig:plot_ITL_MAL) is shown that *ITB* is bigger for youg womans because of the weight of infant mortality in their mother experience, also with more dispersion for same levels of parity.
+Looks similar to the transcendental entropy measure $H$ (@KeyfitzMath) but considering all the cohorts born during the the mother´s life, weighted by their relative size $m_x$.
+In the figure \@ref(fig:plot_ITL_MAL) is shown that *ITB* is bigger for young women because of the weight of infant mortality in their mother experience, also with more dispersion for same levels of parity.
 Assuming constant period rates the expected time in years that a mother aged 30 passed with a death son was around 4% in some countries at middle XX Century.
-When increasing age, the survival experience depends less on infant mortality, and the distribution is around less than 2% for womans aged 50, converging to 0 on time.
+When increasing age, the survival experience depends less on infant mortality, and the distribution is around less than 2% for women aged 50, converging to 0 on time.
 As extreme cases, in 1950-1955 Haiti women aged 30 would have experienced an intensity of `r round(CS_outs$ITL[CS_outs$name=="Haiti" & CS_outs$Age==30 & CS_outs$Period=="1950-1955"],1)`%, and in 2010-2015 `r round(CS_outs$ITL[CS_outs$name=="Haiti" &CS_outs$Age==30 & CS_outs$Period=="2010-2015"],1)`%, while the women of Costa Rica `r round(CS_outs$ITL[CS_outs$name=="Costa Rica" & CS_outs$Age==30 & CS_outs$Period=="1950-1955"],1)`%, and in 2010-2015 `r round(CS_outs$ITL[CS_outs$name=="Costa Rica" &CS_outs$Age==30 & CS_outs$Period=="2010-2015"],1)`%.
@@
 This relation can be derived by starting with the mother age $x+t$ at each death child age $t$ at death, weighted by the fertility and survival function.
-In it, $\kappa$ is the mean age at childbirth for womens aged $a$, $MAD_{a-x}$ refers to the mean age at death for newborns that die before $a-x$, $F_a$ is the cumulated fertility for a women aged $a$:
+In it, $\kappa$ is the mean age at childbirth for women aged $a$, $MAD_{a-x}$ refers to the mean age at death for newborns that die before $a-x$, $F_a$ is the accumulated fertility for a women aged $a$:
 
@@
 
-For those latinamerican mothers aged 50 with 3 daughters born who suffered a lost, theyy experienced that at age 30 in average (figure \@ref(fig:plot_ITL_MAL)) (More can be analyzed here).
+For those Latin American mothers aged 50 with 3 daughters born who suffered a lost, they experienced that at age 30 in average (figure \@ref(fig:plot_ITL_MAL)) (More can be analyzed here).
 
@@
 
-Dividing both sides by $CS_{a}$ in an discrete approzimation we get
+Dividing both sides by $CS_{a}$ in an discrete approximation we get
 
@@
 This is intuitive considering that younger descendants experiences longer periods of exposure to risk.
-The figure \@ref(fig:CS_abs_app) shows that goodness of fit is descreasing with the change size, given that $\delta$ is assumed near zero (as an example considering $\delta=0.01$ means at age 50 that $\frac{l_50}{e^{\delta\,50}} = \frac{l_50}{1.6}$, a big change).
+The figure \@ref(fig:CS_abs_app) shows that goodness of fit is decreasing with the change size, given that $\delta$ is assumed near zero (as an example considering $\delta=0.01$ means at age 50 that $\frac{l_50}{e^{\delta\,50}} = \frac{l_50}{1.6}$, a big change).
 
@@
 
-That an equal change in mortality could happens at all ages is unlikelihood. As a [corollary](#Corollary) we inspect the case thtat only infant mortality is affected.
+That an equal change in mortality could happens at all ages is unlikelihood. As a [corollary](#Corollary) we inspect the case that only infant mortality is affected.
 
@@
 Dividing by $CS_a$ and multiplying by $\delta$ get one useful expression.
-This is the negative of cumulative hazard $H_a$ but considering a factor $0\leq \frac{CS_{a-t}}{CS_a}\leq1$ that takes in acccount the amount of survinvig descendatns that would be lost for a mother aged *a* in each age x of risk.
+This is the negative of cumulative hazard $H_a$ but considering a factor $0\leq \frac{CS_{a-t}}{CS_a}\leq1$ that takes in account the amount of surviving descendants that would be lost for a mother aged *a* in each age x of risk.
 
@@
 
-This factor has a S-shape because $l_x$ curvature fertility acumulation, that gives more weight to first ages, and which is grater than 1 only for ages *a* with complete reproductive life and high infant mortality (with no aditional childs and higher risk for one age additional of survive).
+This factor has a S-shape because $l_x$ curvature fertility accumulation, that gives more weight to first ages, and which is grater than 1 only for ages *a* with complete reproductive life and high infant mortality (with no additional children and higher risk for one age additional of survive).
 
@@
 
-Following @Coresh1988 we can express the fertility part in a multiplicative way with variability in the level but not in the shape, as ${m(k)}_x = F_\beta\,k\,r_x$, where $F_\beta$ is the baseline cumulated fertility, $K$ is a random variable with mean 1 that allows variability between groups and $r_x$ is the fertility structure by age ($\int_{0}^{\beta}{r_x}=1$).
+Following @Coresh1988 we can express the fertility part in a multiplicative way with variability in the level but not in the shape, as ${m(k)}_x = F_\beta\,k\,r_x$, where $F_\beta$ is the baseline accumulated fertility, $K$ is a random variable with mean 1 that allows variability between groups and $r_x$ is the fertility structure by age ($\int_{0}^{\beta}{r_x}=1$).
 Frailty part can be thought in a cohort effect way, also in the multiplicative assumption as @Vaupel2014, with $l(z)_x=e^{-H_{x}Z}=l_x^Z$, with baseline hazard $\mu$.
@@
 
-The way to chain dependence is convenient to the expected herarchy $f_{zk}=f_{k|z}f_z$: groups with higher descendant´s mortality would adjust their fertility level, with positive correlation.
+The way to chain dependence is convenient to the expected hierarchy $f_{zk}=f_{k|z}f_z$: groups with higher descendant´s mortality would adjust their fertility level, with positive correlation.
 Isolating one age and defining $f_{kz}$ we can explore effects in $CS_a$.
@@
 Mortality and fertility data was obtained from the last revision of population prospects by UN (@WPP19).
-We smoothed female $l_x$ using cubic-splines constrained to monothonic descense, taking later $L_0$ and $T_{100}$ as inputs for year-person calculations.
+We smoothed female $l_x$ using cubic-splines constrained to monotonic decrease, taking later $L_0$ and $T_{100}$ as inputs for year-person calculations.
 For splitting fertility five groups was used quadratic optimization approach by @Michalski2018, with an desirable property for our purpose which is a good fitting in parity.
@@
 
-This means that an absolute change in infant hazard rate affects proportionally for all the age range extension (because of all the cohorts aged more than 1), except for the mean time exposure for those who born and survive in the last year $1-(a-k_{a-1,a})$, times $\frac{CS_{a-1,a}}{CS_a}$ the portion of cumulated alive daughters for the same period, weighting how important was this in total succesful experience (in terms of alive descendants).
+This means that an absolute change in infant hazard rate affects proportionally for all the age range extension (because of all the cohorts aged more than 1), except for the mean time exposure for those who born and survive in the last year $1-(a-k_{a-1,a})$, times $\frac{CS_{a-1,a}}{CS_a}$ the portion of cummulative alive daughters for the same period, weighting how important was this in total successful experience (in terms of alive descendants).
 

### 2020-04-30 ALAP2020/ALAP.Rmd
sugerencias antes de charla el viernes

@@
 
-We find that, seen from the perspective of a mother, child survival mainly depends on the cummulative fertility function^[This runs contrary to the findings of @KeyfitzMath, who found that levels of fertility did not affect the probability of having living ancestors.] and the mean age at childbearing $\kappa$ acts through the factor $l_{a-\kappa}$.
+We find that, seen from the perspective of a mother, child survival mainly depends on the cummulative fertility function^[Note that this runs contrary to the findings of @KeyfitzMath, who found that levels of fertility did not affect the probability of having living ancestors.] and the mean age at childbearing $\kappa$ acts through the factor $l_{a-\kappa}$.
 The dispersion of of fertility $\sigma^2$ is negatively proportional to child survival. 
@@
 
-### Burden and timing
+### Burden and timing of maternal bereavement
 
@@
 ```{r plot_ITL_MAL, fig.cap = "Intensity Tome Lost of  womans aged 30, 40 and 50. Years 1950 (darker) to 2015 (lighter). Latinamerican countries in period 1950-2015"}
-ITL <- ggplot(CS_outs)+  
-                    geom_smooth(method="lm", aes(x=Fa, y=ITL, color=factor(Age)), 
-                                se = F, formula= (y ~ exp(x)))+
+ITL <- 
+  ggplot(CS_outs) +
+  geom_smooth(
+    method="lm"
+    , aes(x=Fa, y=ITL, color=factor(Age))
+    , se = F
+    , formula= (y ~ exp(x))
+    ) +
   geom_point(aes(x=Fa, y=ITL, color=factor(Age)), alpha = .3) +
-                    theme_classic()+        
+  theme_classic() +
   theme(legend.position = "right") +
@@
 We now consider the consequences of an absolute change $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk. 
-Considering that $m_{x,\delta}=m_{x}+\delta$ (@wrycza2012), and $l_{a-x}^\delta = e^{-\int_{0}^{a-x}{(\mu_t+\delta})dt}$:
+As a [corollary](#Corollary) we consider the case where mortality only changes for infants, considering that it is unlikelily for a potential change in mortality to affect all ages equally in real populations.
+For now, let us consider $m_{x,\delta}=m_{x}+\delta$ (@wrycza2012) and $l_{a-x}^\delta = e^{-\int_{0}^{a-x}{(\mu_t+\delta})dt}$:
 
@@
 
-```{r CS_abs_app, fig.cap= "Absolute change and its approximation for age CS in women aged 50. Latinamerican countries in period 1950-2015"}
-# application of absolute changes
-age = 50
-CS_abs_change_app = data.frame(delta = NA,Country = NA, Period = NA, change = 0, change_app = 0)
-for(delta in seq(0,.01,.001)){
-  for (country in unique(LA$name)){
-    for (period in unique(LA$Period)){
-      # country = "Argentina"; period = "2010-2015"; delta = .01
-      change_app = -(age - CS_outs %>% filter(name == country, 
-                                              Period == period, 
-                                              Age == age) %>% select(k)) * delta
-      country_i = LA %>% filter(name == country, Period == period) 
-      CS = CS_outs %>% filter(name == country, 
-                              Period == period, 
-                              Age == age) %>% select(CS)
-      CS_changed = sum(unlist(sapply(1:age, function(i) 
-        country_i$asfr[i]*country_i$lx[age-(i-1)]/100000*exp(-delta*(age-(i-1))))), 
-        na.rm = T)
-      change = (CS_changed-CS)/CS
-      CS_abs_change_app = rbind(CS_abs_change_app, 
-                                data.frame(delta = delta,
-                                           Country = country, Period = period, 
-                                           change = as.numeric(change), 
-                                           change_app = as.numeric(change_app)))
-    }
-  } 
-}
-CS_abs_change_app %>% slice(-1) %>% 
+```{r CS_abs_app, fig.cap= "Effect of an absolute absolute change in mortality on child survival by maternal age of women in Latin America  for the 1950-2015 period.
+Goodness of fit of approximation compared to direct estimation"}
+
+# NEW CODE ----
+
+abs30 <- absolute_change_fit(30)
+abs40 <- absolute_change_fit(40)
+abs50 <- absolute_change_fit(50)
+abs70 <- absolute_change_fit(70)
+
+bind_rows(abs30, abs40, abs50, abs70) %>% 
+  mutate(age = paste("Maternal age", age)) %>% 
   ggplot() + 
@@
   geom_abline(slope=1, intercept = 0) +
-  theme_classic() +
-  # theme(legend.position = "none")+
-  xlim(-.3, 0) + ylim(-.3, 0)
+  scale_x_continuous("Actual change in CS", limits = c(-.3, 0)) +
+  scale_y_continuous("Approximated change in CS", limits = c(-.3, 0)) +
+  scale_color_continuous("Absolute change\nin delta") +
+  facet_wrap(~age) +
+  theme_classic()
+
+# ORIGINAL CODE ----
+# application of absolute changes
+
+# age <- 50
+# 
+# CS_abs_change_app <- data.frame(delta = NA,Country = NA, Period = NA, change = 0, change_app = 0)
+# 
+# for(delta in seq(0,.01,.001)){
+# # for(delta in seq(0,.02,.005)){
+#   for (country in unique(LA$name)){
+#     for (period in unique(LA$Period)){
+#       # country = "Argentina"; period = "2010-2015"; delta = .01
+#       change_app = -(age - CS_outs %>% filter(name == country, 
+#                                               Period == period, 
+#                                               Age == age) %>% 
+#                        select(k)) * delta
+#       
+#       country_i = LA %>% filter(name == country, Period == period) 
+#       CS = CS_outs %>% filter(name == country, 
+#                               Period == period, 
+#                               Age == age) %>% select(CS)
+#       CS_changed = sum(unlist(sapply(1:age, function(i) 
+#         country_i$asfr[i]*country_i$lx[age-(i-1)]/100000*exp(-delta*(age-(i-1))))), 
+#         na.rm = T)
+#       change = (CS_changed-CS)/CS
+#       CS_abs_change_app = rbind(CS_abs_change_app, 
+#                                 data.frame(delta = delta,
+#                                            Country = country, Period = period, 
+#                                            change = as.numeric(change), 
+#                                            change_app = as.numeric(change_app)))
+#     }
+#   } 
+# }
+# 
+# CS_abs_change_app %>% 
+#   slice(-1) %>% 
+#   ggplot() + 
+#   geom_point(aes(x=change, y=change_app, color=delta)) +
+#   geom_abline(slope=1, intercept = 0) +
+#   scale_x_continuous("Actual change in CS", limits = c(-.3, 0)) +
+#   scale_y_continuous("Approximated change in CS", limits = c(-.3, 0)) +
+#   scale_color_continuous("Absolute change\nin delta") +
+#   theme_classic()
+
 ```
 
-That an equal change in mortality could happens at all ages is unlikelihood. As a [corollary](#Corollary) we inspect the case that only infant mortality is affected.
+We now consider the consequences of a proportional change in mortality on child survival.
+Given a proportional change in mortality $\mu_{x,\delta}=\mu_{x}(1+\delta)$ so that $l_{a-x}^\delta = e^{-\int_{0}^{a-x}{\mu_t(1+\delta})dt}=(l_{a-x})^{(1+\delta)}$, it follows that:
 
-The relationship is less clear for a proportional change given the interaction of birth and mortality rates (@KeyfitzMath).
-Considering that $\mu_{x,\delta}=\mu_{x}(1+\delta)$ and $l_{a-x}^\delta = e^{-\int_{0}^{a-x}{\mu_t(1+\delta})dt}=(l_{a-x})^{(1+\delta)}$:
-
-$$CS_{a}^\delta = \int_{\alpha}^{a} {m_{x} l_{a-x}^{(1+\delta)}} dx$$
+$$CS_{a}^\delta = \int_{\alpha}^{a} {m_{x} l_{a-x}^{(1+\delta)}} dx.$$
 
@@
 
+Grasping this relationship intuitively is more difficult to  given the interaction of birth and mortality rates (@KeyfitzMath). 
 This factor has a S-shape because $l_x$ curvature fertility accumulation, that gives more weight to first ages, and which is grater than 1 only for ages *a* with complete reproductive life and high infant mortality (with no additional children and higher risk for one age additional of survive).
+
+<!-- Do we get any inuition about how proportional changes affects CS? -->
 
@@
 
-In a model with heterogeneity, determined at born with a multiplicative effect, $CS_a$ could be interpreted as a conditional expectation with random variables *K* for fertility heterogeneity and *Z* for mortality frailty.
+In a model with heterogeneity, determined at birth with a multiplicative effect, $CS_a$ could be interpreted as a conditional expectation with random variables *K* for fertility heterogeneity and *Z* for mortality frailty.
 
@@
 
-Following @Coresh1988 we can express the fertility part in a multiplicative way with variability in the level but not in the shape, as ${m(k)}_x = F_\beta\,k\,r_x$, where $F_\beta$ is the baseline accumulated fertility, $K$ is a random variable with mean 1 that allows variability between groups and $r_x$ is the fertility structure by age ($\int_{0}^{\beta}{r_x}=1$).
+Following @Coresh1988 we can express the fertility component in a multiplicative way with variability in the level but not in the shape, as ${m(k)}_x = F_\beta\,k\,r_x$, where $F_\beta$ is the baseline accumulated fertility, $K$ is a random variable with mean 1 that allows variability between groups and $r_x$ is the fertility structure by age ($\int_{0}^{\beta}{r_x}=1$).
 Frailty part can be thought in a cohort effect way, also in the multiplicative assumption as @Vaupel2014, with $l(z)_x=e^{-H_{x}Z}=l_x^Z$, with baseline hazard $\mu$.

### 2020-04-30 ALAP2020/funs.R

@@ -128,13 +128,73 @@ LA <- lx_paises_LA_s %>% slice(-1) %>% mutate(country_code=as.integer(country_co
 LA$name[LA$name == "Bolivia (Plurinational State of)"] = "Bolivia"
 
 
-# get_results
+# get_results ---------------------
+
 CS_outs <- as.data.frame(expand.grid(name = unique(LA$name),
-                                     Age = c(30,40,50),
+                                     # Age = c(30,40,50),
+                                     Age = c(30,40,50, 70),
                                      Period = unique(LA$Period)))
 for(i in 1:nrow(CS_outs)){
-  CS_outs[i,4:12] = CS_results(country = CS_outs[i,1], 
-                               period = CS_outs[i,3], 
-                               a = CS_outs[i,2])}
+  CS_outs[i, 4:12] <- CS_results(
+    country = CS_outs[i,1],
+    period = CS_outs[i,3], 
+    a = CS_outs[i,2]
+    )
+  }
+
 CS_outs <- CS_outs %>% mutate(Year = as.integer(substr(Period, 1, 4)) + 2.5)
 
+# Other functions ----
+
+age <- 50
+
+absolute_change_fit <- function(age) {
+
+  CS_abs_change_app <- data.frame(delta = NA,Country = NA, Period = NA, change = 0, change_app = 0)
+  
+  for(delta in seq(0,.01,.001)){
+    # for(delta in seq(0,.02,.005)){
+    for (country in unique(LA$name)){
+      for (period in unique(LA$Period)){
+        # country = "Argentina"; period = "2010-2015"; delta = .01
+        change_app = -(age - CS_outs %>% filter(name == country, 
+                                                Period == period, 
+                                                Age == age) %>% 
+                         select(k)) * delta
+        
+        country_i = LA %>% filter(name == country, Period == period) 
+        CS = CS_outs %>% filter(name == country, 
+                                Period == period, 
+                                Age == age) %>% select(CS)
+        CS_changed = sum(unlist(sapply(1:age, function(i) 
+          country_i$asfr[i]*country_i$lx[age-(i-1)]/100000*exp(-delta*(age-(i-1))))), 
+          na.rm = T)
+        change = (CS_changed-CS)/CS
+        CS_abs_change_app = rbind(CS_abs_change_app, 
+                                  data.frame(delta = delta,
+                                             Country = country, Period = period, 
[…]

### 2020-05-01 ALAP2020/ALAP.Rmd
added covid paragraph

@@
 
+How do changing mortality rates affect the availability of offspring over age for mothers, if fertility remains constant?
 How can we characterize the timing of offspring mortality as experienced by a mother from a formal demographic perspective? 
 <!-- How do changes in mortality affect the availability of offspring over age from the point of view of a prospective mother? -->
-How do changing mortality rates affect the availability of offspring over age for mothers, if fertility remains constant?
 In this paper, we uncover formal relationships to answer these questions and provide numerical applications using data from countries in Latin America during the second half of the twentieth century.
-Please see the [data section](#Data_appendix) in the Appendix for more details on the data and on the analytical procedures.
+See the [data section](#Data_appendix) in the Appendix for more details on the data and on the analytical procedures.
+
+Recent development have underscored the need to understand how sudden changes in mortality affect the availability of kin. 
+In the context of the global pandemic caused by the Covid-19 disease, elderly people depend on support provided by their relatives, especially during periods of lockdown and in the absence of governmental support mechanisms.
+Elderly parents may be at a higher risk of losing the key support provided by their adult children given the known age-gradient in the Covid-19 case fatality rate.
+Our approach is key for understanding the impact of age-specific excess mortality on the resilience or otherwise of kinship networks for the vulnerable elderly population.
 
@@
 The example makes the simplifying assumption that the UN WPP period rates remain in place during sufficient time for the populations to approach demographic stability.
-We hope that this inaccuracy will be forgiven for the sake of simplicity. 
+<!-- We hope that this inaccuracy will be forgiven for the sake of simplicity.  -->
 A more sophisticated model using the appropriate age-specific cohort rates in the subscripts has been proposed by @alburez_gutierrez_womens_2019.

### 2020-05-01 ALAP2020/ALAP.Rmd
before merge charla 1/5

--- 2020-04-28 ALAP2020/ALAP.Rmd
@@
 
-Kin count estimation has a long and distinguished pedigree in mathematical demography, starting with the work of @lotka1931orphanhood on modelling orphanhood in theoretical populations across demographic regimes.
-To the best of our knowledge, @brass_derivation_1953 first proposed an equation akin to Eq.
-\ref{eq:CS} to estimate child survival over age.
+Kin count estimation has a long pedigree in mathematical demography.
+@lotka1931orphanhood proposed to model the orphanhood in a theorical population, comparing the effect across demographic regimes.
+Equation  \ref{eq:CS} was originally proposed by @brass_derivation_1953 to estimate child mortality.
 The notion of estimating the expected number of living daughters in a stable population was generalized by @goodman1974 for other kin relations (granddaughters, cousins, etc.).
-<!-- We should clarify this sentence a bit: -->
-The "counting method" approach was further popularized in @KeyfitzMath, where the net reproduction rate $R_0$ was limited to $a$ or to the (not complete) ratio between generations (multiplying and dividing by the mother generation $B$).
-@bongaarts_projection_1987 used a similar approach to estimate descendants in his 'Family Status Model' and, more recently, @wrycza2012 has used formal techniques to describe the effect of different kind of changes in mortality by age.
+The "counting method" approach, being the net reproduction rate $R_0$ limited to $a$, or the (not complete) ratio between generations (multiplying and dividing by the mother generation $B$), was further popularized by @KeyfitzMath.
+@bongaarts_projection_1987 used a similar approach to estimate descendants in his 'Family Status Model'.
+More recently, @wrycza2012 has looked at the formal effect of different kind of changes in mortality by age.
 
-The question of child survival, particularly the survival of daughters for mothers, sits at the very center of demographic theory.
-This is evidenced by the omnipresence of the 'net reproductive rate' in accounts of human and non-human populations. 
-Historical demographers draw liberally on assumptions about kin availability and individual's exposure to offspring mortality to explain fertility decisions, especially around the demographic transition.
-These assumptions are often untested given the data sparsity [@livi_bacci_concise_1997;@volk_infant_2013]. 
-Economic theories of fertility that consider children as 'investment' for old-age support also rely on notions of child survival [@preston1978effects].
+<!---
+Improve Intro: ¿Why is important study survival descendants from mother perspective?¿How impacts mothers health/welfare/experience?
+-->
 
-Child survival - and it's counterpart, child death - gain new significance in the wake of the fertility transition.
-Depressed fertility means that fewer children are tasked with providing key emotional, social, and financial transfers to aging parents for ever increasing periods of time.
-With increasing periods of generational overlap, individuals find themselves 'sandwiched' between aging parents and young children requiring their simultaneous attention and care [@daatland_norwegian_2010].
-In the context of global population aging, elderly parents without access to formal social security and pension systems are particularly reliant on these transfers to make ends meet [@smith-greenaway_maternal_2020].
-Losing an only child may become a more common experience in the context of global fertility decline, something particularly worrying for parents suddenly rendered childless.
-Given the known psychological, health, and social consequences of child loss for bereaved parents and families [@Hendrickson2009;@lee_death_2014], it is surprising that the demographic processes that shape parental bereavement remain very poorly understood.
+The objective of this work is to give a formal approach to deomgraphy of surviving childs from mother perspective, looking for relations to characterize this situation: the timing of descedent mortality (mean age at the event and time spent in that state) and the effect of mortality change in the final descendants quantity.
+Numerical applications of the expressions will be done for Latinoamerican countries during the second half of XX Century.
+Details on data handling and smooth procedures are exposed in [data section](#Data_appendix) in the Appendix.
 
-How can we characterize the timing of offspring mortality as experienced by a mother from a formal demographic perspective? 
-<!-- How do changes in mortality affect the availability of offspring over age from the point of view of a prospective mother? -->
-How do changing mortality rates affect the availability of offspring over age for mothers, if fertility remains constant?
-In this paper, we uncover formal relationships to answer these questions and provide numerical applications using data from countries in Latin America during the second half of the twentieth century.
-Please see the [data section](#Data_appendix) in the Appendix for more details on the data and on the analytical procedures.
+## Formulas
 
-## Formal development
+$CS_a$ is the expected number of surviving children to a mother^[We´ll refer to mothers, assuming that fertility rates applies to all woman, meaning that all of them are exposed.] alive at aged $a$ in a female stable population with fertility rates $m_{x}$, mortality hazard $\mu_{x}$ and survival function $l_x=e^{-\int_{0}^{x}{\mu_t}\,dt}$ ($l_0=1$ during this work), as proposed by @goodman1974:
 
-Let $CS_a$ be the expected number of surviving children to a mother^[We use woman and mother interchangeably in this paper, assuming that all women are exposed to the same fertility rates.] alive at aged $a$ in a female stable population with fertility rates $m_{x}$, mortality hazard $\mu_{x}$ and survival function $l_x=e^{-\int_{0}^{x}{\mu_t}\,dt}$ (with unit radix $l_0=1$), as proposed by @goodman1974:
+$$CS_a = \int_{0}^a{m_{x}l_{a-x}dx}$$
 
-$$CS_a = \int_{0}^a{m_{x}l_{a-x}dx}.$$
+Between 1950-1955 and 2010-2015 the portion of surviving daugthers  $\frac{CS_a}{F_a}$ at age 50 passed from a simple mean of `r mean(CS_outs$CS_prob[CS_outs$Period == "1950-1955" & CS_outs$Age==50])` to `r mean(CS_outs$CS_prob[CS_outs$Period == "2010-2015" & CS_outs$Age==50])`, also with a decrease of `r mean(CS_outs$CS[CS_outs$Period == "1950-1955" & CS_outs$Age==50])-mean(CS_outs$CS[CS_outs$Period == "2010-2015" & CS_outs$Age==50])` in the expected descendants at same age.
+Althought this is not a  real measure but a theorical one if the rates would have been kept constant during enough time.
 
-<!-- Que te parece algo asi? -->
+<!---
+Interpretation
+-->
 
-We motivate this paper with an empirical example using fertility and mortality rates from the 2019 Revision of the UN World Population Prospects (UN WPP) for the Latin American Region. 
-The example makes the simplifying assumption that the UN WPP period rates remain in place during sufficient time for the populations to approach demographic stability.
-We hope that this inaccuracy will be forgiven for the sake of simplicity. 
-A more sophisticated model using the appropriate age-specific cohort rates in the subscripts has been proposed by @alburez_gutierrez_womens_2019.
-
-Consider a Latin American woman standing before us.
-If this were 1950-1955, she could reasonable expect to have `r mean(CS_outs$CS[CS_outs$Period == "1950-1955" & CS_outs$Age==50])` surviving children on her 50$^{th}$ birthday.
-In 2010-2015 a woman the same age would only have `r mean(CS_outs$CS[CS_outs$Period == "2010-2015" & CS_outs$Age==50])` living children. 
-The difference of `r mean(CS_outs$CS[CS_outs$Period == "1950-1955" & CS_outs$Age==50])-mean(CS_outs$CS[CS_outs$Period == "2010-2015" & CS_outs$Age==50])` children is explained by reduced fertility and improved mortality in the region [give stats].
-We now remove the effect of changing fertility by considering the number of daughters surviving up to maternal age $a$ as a proportion of the daughters ever born to a woman that age, $\frac{CS_a}{F_a}$.
-Given Eq.
-\ref{eq:CS}, the increase from `r mean(CS_outs$CS_prob[CS_outs$Period == "1950-1955" & CS_outs$Age==50])` in 1950-1955 to `r mean(CS_outs$CS_prob[CS_outs$Period == "2010-2015" & CS_outs$Age==50])` in 2010-2015 must be explained by a change in mortality.
-In the remainder of the paper, we aim to formalize the relationship population-level changes in mortality rates and perceived change in the lived experience of child death from the perspective of a mother.
-We consider the case when these changes are additive, mutliplicative, and when they are not evenly distributed across age.
-
-<!-- Add a line for TFR in plot? -->
-<!-- Por que no llegas hasta 2020? -->
-
-```{r plot_CS, fig.cap="Child Survival and Child survival as a share of cummulative fertility by age, for women aged 30, 40 and 50.
-Estimates using UN WPP data for Latin American countries in the period 1950-2015 period"}
+```{r plot_CS, fig.cap="Child Survival and Child survival related to cumulated fertility, for women aged 30, 40 and 50. Latinamerican countries in period 1950-2015"}
 a <- ggplot(CS_outs, aes(x=Year, y=CS, color=factor(Age)))+
@@
 
-### An approximation of $CS_a$, the expected number of surviving daughters
+### An approximation
 
-<!-- QUe te parece esta reorganizacion de esta seccion? -->
-
-As an initial step, we seek an intuitive understanding of the effect of changing mortality on child survival. 
-Building on work by @KeyfitzMath for the probability of a living mother, we show which features of the child survival function explain the expected number of children. 
-For this we use an approximate of $l_x$ using Taylor´s theorem until second order around the mean age of childbearing $\kappa$ for all $a>=\alpha$:
+Following @KeyfitzMath work about living mothers count, we can consider $a>=\beta$ and assume that fertility pattern by age is concentrated around $\kappa$ (mean age at childbearing), and cumulated fertility (or gross reproduction rate because of female model) is $F_a =  \int_{\alpha}^a{m_{x}dx}$.
+An approximation of $l_x$ using Taylor´s theorem until second order around $\kappa$ would be:
 
@@
 
-where the fertility pattern by age is concentrated around $\kappa$ and the accumulated fertility (or gross reproduction rate in our same-sex model) is $F_a =  \int_{\alpha}^a{m_{x}dx}$. 
 The second Taylor´s term is null because $\int_{\alpha}^a{x \, m_{x}dx} = \kappa \, F_a$.
+The formula tells us that the survival experience for a woman aged *a* depends mainly of cumulated fertility (possible cases, different from the ancestor count case where this doesn´t affect) and child survival probability (success cases), but also negative proportional affected by how disperse is the fertility by age (variance \sigma^2) and the suvival curvature around (negative sign), in a $l_x$ range where is tipycally very flat (range 20-40 years old) in transitioned populations.
+Guatemala improved the approximation with years due to rectangularization process in $l_x$ (figure \@ref(fig:plot_CS_aprox)).
+And change in $\kappa$, taking logs and deriving: $\frac{dCS_a}{d\kappa}\frac{1}{CS_a} \approx \mu_{a-\kappa}F_a \Delta\kappa$
 
-We find that, seen from the perspective of a mother, child survival mainly depends on the cummulative fertility function^[This runs contrary to the findings of @KeyfitzMath, who found that levels of fertility did not affect the probability of having living ancestors.] and the mean age at childbearing $\kappa$ acts through the factor $l_{a-\kappa}$.
-The dispersion of of fertility $\sigma^2$ is negatively proportional to child survival. 
-All else constant, a higher dispersion of fertility $\sigma^2$ is associated with a decrease in $CS_a$ during reproductive age.
-The negative relationship stems from the fact that the second derivative of $l_{a-\kappa}$ is negative in the 15-49 age range, given a concave downwards survival function.
-
-<!-- No cambie esto porque no me queda tan claro cual es la conexion con el resto del texto - podemos discutir -->
-
-Guatemala improved the approximation with years due to rectangularization process in $l_x$ (figure \@ref(fig:plot_CS_aprox)).
-
-```{r plot_CS_aprox, fig.cap = "Left: Error in approximation for a=40, from years 1950 (darker) to 2015 (lighter) for all Latin American countries (blue) and Guatemala (red).
-Right: Change in survival and fertility by age in Guatemala."}
+```{r plot_CS_aprox, fig.cap = "Error in approximation for a=40, from years 1950 (darker) to 2015 (lighter). Guatemala in red (left). Guatemala survival and fertility by age (right)"}
 a <- ggplot(CS_outs %>% filter(Age == 40), 
@@
 
-Where $d_t$ is the death distribution from birth, $e_{0|a-x-t}$ is the life expectancy at birth until age $a-x-t$ and $e_{0|a-x}^\dagger$ the temporary dispersion measure.
+Where $d_t$ is the death distribution from birth, $e_{0|a-x-t}$ is the life expectancy at birth until age $a-x-t$ and $e_{0|a-x}^\dagger$ the temporary dispertion measure.
 But most interesting would be to compare these years with the time that mothers would expect to live with their daughters.
@@
 This is a ratio between child-years in two radically different states.
-Looks similar to the transcendental entropy measure $H$ (@KeyfitzMath) but considering all the cohorts born during the the mother´s life, weighted by their relative size $m_x$.
-In the figure \@ref(fig:plot_ITL_MAL) is shown that *ITB* is bigger for young women because of the weight of infant mortality in their mother experience, also with more dispersion for same levels of parity.
+Looks similar to the trascendental entropy measure $H$ (@KeyfitzMath) but considering all the cohorts born during the the mother´s life, weighted by their relative size $m_x$.
+In the figure \@ref(fig:plot_ITL_MAL) is shown that *ITB* is bigger for youg womans because of the weight of infant mortality in their mother experience, also with more dispersion for same levels of parity.
 Assuming constant period rates the expected time in years that a mother aged 30 passed with a death son was around 4% in some countries at middle XX Century.
-When increasing age, the survival experience depends less on infant mortality, and the distribution is around less than 2% for women aged 50, converging to 0 on time.
+When increasing age, the survival experience depends less on infant mortality, and the distribution is around less than 2% for womans aged 50, converging to 0 on time.
 As extreme cases, in 1950-1955 Haiti women aged 30 would have experienced an intensity of `r round(CS_outs$ITL[CS_outs$name=="Haiti" & CS_outs$Age==30 & CS_outs$Period=="1950-1955"],1)`%, and in 2010-2015 `r round(CS_outs$ITL[CS_outs$name=="Haiti" &CS_outs$Age==30 & CS_outs$Period=="2010-2015"],1)`%, while the women of Costa Rica `r round(CS_outs$ITL[CS_outs$name=="Costa Rica" & CS_outs$Age==30 & CS_outs$Period=="1950-1955"],1)`%, and in 2010-2015 `r round(CS_outs$ITL[CS_outs$name=="Costa Rica" &CS_outs$Age==30 & CS_outs$Period=="2010-2015"],1)`%.
@@
 This relation can be derived by starting with the mother age $x+t$ at each death child age $t$ at death, weighted by the fertility and survival function.
-In it, $\kappa$ is the mean age at childbirth for women aged $a$, $MAD_{a-x}$ refers to the mean age at death for newborns that die before $a-x$, $F_a$ is the accumulated fertility for a women aged $a$:
+In it, $\kappa$ is the mean age at childbirth for womens aged $a$, $MAD_{a-x}$ refers to the mean age at death for newborns that die before $a-x$, $F_a$ is the cumulated fertility for a women aged $a$:
 
@@
 
-For those Latin American mothers aged 50 with 3 daughters born who suffered a lost, they experienced that at age 30 in average (figure \@ref(fig:plot_ITL_MAL)) (More can be analyzed here).
+For those latinamerican mothers aged 50 with 3 daughters born who suffered a lost, theyy experienced that at age 30 in average (figure \@ref(fig:plot_ITL_MAL)) (More can be analyzed here).
 
@@
 
-Dividing both sides by $CS_{a}$ in an discrete approximation we get
+Dividing both sides by $CS_{a}$ in an discrete approzimation we get
 
@@
 This is intuitive considering that younger descendants experiences longer periods of exposure to risk.
-The figure \@ref(fig:CS_abs_app) shows that goodness of fit is decreasing with the change size, given that $\delta$ is assumed near zero (as an example considering $\delta=0.01$ means at age 50 that $\frac{l_50}{e^{\delta\,50}} = \frac{l_50}{1.6}$, a big change).
+The figure \@ref(fig:CS_abs_app) shows that goodness of fit is descreasing with the change size, given that $\delta$ is assumed near zero (as an example considering $\delta=0.01$ means at age 50 that $\frac{l_50}{e^{\delta\,50}} = \frac{l_50}{1.6}$, a big change).
 
@@
 
-That an equal change in mortality could happens at all ages is unlikelihood. As a [corollary](#Corollary) we inspect the case that only infant mortality is affected.
+That an equal change in mortality could happens at all ages is unlikelihood. As a [corollary](#Corollary) we inspect the case thtat only infant mortality is affected.
 
@@
 Dividing by $CS_a$ and multiplying by $\delta$ get one useful expression.
-This is the negative of cumulative hazard $H_a$ but considering a factor $0\leq \frac{CS_{a-t}}{CS_a}\leq1$ that takes in account the amount of surviving descendants that would be lost for a mother aged *a* in each age x of risk.
+This is the negative of cumulative hazard $H_a$ but considering a factor $0\leq \frac{CS_{a-t}}{CS_a}\leq1$ that takes in acccount the amount of survinvig descendatns that would be lost for a mother aged *a* in each age x of risk.
 
@@
 
-This factor has a S-shape because $l_x$ curvature fertility accumulation, that gives more weight to first ages, and which is grater than 1 only for ages *a* with complete reproductive life and high infant mortality (with no additional children and higher risk for one age additional of survive).
+This factor has a S-shape because $l_x$ curvature fertility acumulation, that gives more weight to first ages, and which is grater than 1 only for ages *a* with complete reproductive life and high infant mortality (with no aditional childs and higher risk for one age additional of survive).
 
@@
 
-Following @Coresh1988 we can express the fertility part in a multiplicative way with variability in the level but not in the shape, as ${m(k)}_x = F_\beta\,k\,r_x$, where $F_\beta$ is the baseline accumulated fertility, $K$ is a random variable with mean 1 that allows variability between groups and $r_x$ is the fertility structure by age ($\int_{0}^{\beta}{r_x}=1$).
+Following @Coresh1988 we can express the fertility part in a multiplicative way with variability in the level but not in the shape, as ${m(k)}_x = F_\beta\,k\,r_x$, where $F_\beta$ is the baseline cumulated fertility, $K$ is a random variable with mean 1 that allows variability between groups and $r_x$ is the fertility structure by age ($\int_{0}^{\beta}{r_x}=1$).
 Frailty part can be thought in a cohort effect way, also in the multiplicative assumption as @Vaupel2014, with $l(z)_x=e^{-H_{x}Z}=l_x^Z$, with baseline hazard $\mu$.
-Replacing conveniently the last two expressions and considering the joint distribution $f_{kz}=f_{z/k} \,f_k$, the mean $\overline{CS}_a$ would be:
+Replacing conveniently the last two expressions and considering the joint distribution $f_{kz}=f_{z/k} \,f_k$, the unconditional mean $\overline{CS}_a$ would be:
 
@@
 
-The way to chain dependence is convenient to the expected hierarchy $f_{zk}=f_{k|z}f_z$: groups with higher descendant´s mortality would adjust their fertility level, with positive correlation.
+The way to chain dependence is convenient to the expected herarchy $f_{zk}=f_{k|z}f_z$: groups with higher descendant´s mortality would adjust their fertility level, with positive correlation.
 Isolating one age and defining $f_{kz}$ we can explore effects in $CS_a$.
+
+given K and Z values are fertility and child survival are independent
+K is a partity level assgined to cohorts of women.
 
@@
 Mortality and fertility data was obtained from the last revision of population prospects by UN (@WPP19).
-We smoothed female $l_x$ using cubic-splines constrained to monotonic decrease, taking later $L_0$ and $T_{100}$ as inputs for year-person calculations.
+We smoothed female $l_x$ using cubic-splines constrained to monothonic descense, taking later $L_0$ and $T_{100}$ as inputs for year-person calculations.
 For splitting fertility five groups was used quadratic optimization approach by @Michalski2018, with an desirable property for our purpose which is a good fitting in parity.
@@
 
-This means that an absolute change in infant hazard rate affects proportionally for all the age range extension (because of all the cohorts aged more than 1), except for the mean time exposure for those who born and survive in the last year $1-(a-k_{a-1,a})$, times $\frac{CS_{a-1,a}}{CS_a}$ the portion of cummulative alive daughters for the same period, weighting how important was this in total successful experience (in terms of alive descendants).
+This means that an absolute change in infant hazard rate affects proportionally for all the age range extension (because of all the cohorts aged more than 1), except for the mean time exposure for those who born and survive in the last year $1-(a-k_{a-1,a})$, times $\frac{CS_{a-1,a}}{CS_a}$ the portion of cumulated alive daughters for the same period, weighting how important was this in total succesful experience (in terms of alive descendants).
 

### 2020-05-01 ALAP2020/ALAP.Rmd

--- 2020-05-01 ALAP2020/ALAP.Rmd
@@
 <!-- No cambie esto porque no me queda tan claro cual es la conexion con el resto del texto - podemos discutir -->
+
+The second Taylor´s term is null because $\int_{\alpha}^a{x \, m_{x}dx} = \kappa \, F_a$.
+The formula tells us that the survival experience for a woman aged *a* depends mainly of cumulated fertility (possible cases, different from the ancestor count case where this doesn´t affect) and child survival probability (success cases), but also negative proportional affected by how disperse is the fertility by age (variance \sigma^2) and the suvival curvature around (negative sign), in a $l_x$ range where is tipycally very flat (range 20-40 years old) in transitioned populations.
+Guatemala improved the approximation with years due to rectangularization process in $l_x$ (figure \@ref(fig:plot_CS_aprox)).
+And change in $\kappa$, taking logs and deriving: $\frac{dCS_a}{d\kappa}\frac{1}{CS_a} \approx \mu_{a-\kappa}F_a \Delta\kappa$
 
@@
 
-Following @Coresh1988 we can express the fertility component in a multiplicative way with variability in the level but not in the shape, as ${m(k)}_x = F_\beta\,k\,r_x$, where $F_\beta$ is the baseline accumulated fertility, $K$ is a random variable with mean 1 that allows variability between groups and $r_x$ is the fertility structure by age ($\int_{0}^{\beta}{r_x}=1$).
+Following @Coresh1988 we can express the fertility part in a multiplicative way with variability in the level but not in the shape, as ${m(k)}_x = F_\beta\,k\,r_x$, where $F_\beta$ is the baseline cumulated fertility, $K$ is a random variable with mean 1 that allows variability between groups and $r_x$ is the fertility structure by age ($\int_{0}^{\beta}{r_x}=1$).
 Frailty part can be thought in a cohort effect way, also in the multiplicative assumption as @Vaupel2014, with $l(z)_x=e^{-H_{x}Z}=l_x^Z$, with baseline hazard $\mu$.
-Replacing conveniently the last two expressions and considering the joint distribution $f_{kz}=f_{z/k} \,f_k$, the mean $\overline{CS}_a$ would be:
+Replacing conveniently the last two expressions and considering the joint distribution $f_{kz}=f_{z/k} \,f_k$, the unconditional mean $\overline{CS}_a$ would be:
 
@@
 Isolating one age and defining $f_{kz}$ we can explore effects in $CS_a$.
+
+given K and Z values are fertility and child survival are independent
+K is a partity level assgined to cohorts of women.
 

### 2020-05-12 covid/covid_proposal.Rmd
covid proposal 1

---
title: 'Research proposal: Mortality exposure in times of Covid-19'
author: "Diego Alburez-Gutierrez and Ivan Williams"
date: "12 May 2019"
output:
  pdf_document:
    fig_caption: yes
bibliography: bib.bibtex
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      message=F, warning=F, echo = F)

# library(tidyverse)
# library(gridExtra)
# library(quadprog)
# library(wpp2019)
# options(digits=2)
# source("funs.R")
```

## Introduction

The subject of death has, once again, come to the forefront of public debate as the world grapples with a global pandemic of the Covid-19 disease. 
Much attention has been given to estimating age-specific case fatality rates and mortality rates in general [ref].
The degree to which the pandemic will increase the exposure to mortality in the population as a whole has recieved little or no attention in the demographic literature. 
Studies in sociology and public health have consistently shown the negative and long-term consequences of bereavement on mental, physical, and emotional health, especially for women [ref].
Yet, no study to date has considered the degree to which covid-19 excess mortality will increase expossure to death for people around the world.

Given the age-gradient of the Covid-19 case fatality rates [ref], we expect that the pandemic will result in the death of thousands of elderly relatives, such as parents, grandparents, great-grandparents, aunts and uncles. 
More worringly, it might result in the death of adult children for even older parents who depend on the crucial support given by their offspring or grandchildren.
For many in the Global North this will be their first close encouter with death, as historically low mortality rates have implied a reduction in the exposure to mortality at all ages in high-income countries [@murphy_long-term_2011, @alburez_gutierrez_womens_2019].

## Project description

This study will combine existing and future data on Covid-19 excess mortality [ref] with powerful methods from formal demography [@goodman1974] to estimate the potential burden of covid-19 bereavement.
As such, we are not interested in the total number of people who will die from Covid-19, but in the number of people who will be affected by the death of a relative because of the disease.
In other words, this is a study on `excess bereavement'.
Given the range of covid-19 mortality rates in different countries, we ask: (1) How many people will lose a relative to Covid-19, either in the nuclear or in the extended family? (2) What is the age distribution of the bereaved population?

The project comprises a methodological and an empirical component.
In terms of methodology, we seek to generalize the Kin-Cohort Method, which is itself an extension of the Goodman, Keyfitz Pullum kinship equations [@goodman1974], to estimate the potential increase in the expossure to the death of relatives. 
this means deriving the mathematical expressions needed to model the effect of changes in mortality on the population-level prevalence of excess bereavement.

We exemplify this for the case of parental bereavement, which we understand as the probability of losing a child. 
Let $CS_a$ be the expected number of surviving children to a mother alive at aged $a$ in a female stable population with fertility rates $m_{x}$, mortality hazard $\mu_{x}$ and survival function $l_x=e^{-\int_{0}^{x}{\mu_t}\,dt}$ (with unit radix $l_0=1$), as proposed by @goodman1974:

$$CS_a = \int_{0}^a{m_{x}l_{a-x}dx}.$$

Now consider an absolute increase in mortality $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk.
We can think of this hypothetical increase in mortality as the excess mortality derived from the Covid-19 disease so that $m_{x,\delta}=m_{x}+\delta$ (@wrycza2012) and $l_{a-x}^\delta = e^{-\int_{0}^{a-x}{(\mu_t+\delta})dt}$:

$$CS_{a}^\delta = \int_{\alpha}^{a}m_{x} l_{a-x} e^{-\delta (a-x)} dx.$$

As we have shown elsewhere [ref], we can conceptualize the effects of this absolute change $\delta$ in mortality as the difference between maternal age $a$ and the mean age of the mother at the birth of her surviving daughters $k_a$:

$$\frac{\Delta CS_{a}}{CS_{a}} \approx -(a - k_a) \Delta\delta.$$

This is a first step towards developing a formal understanding the consequences of an increase in age-specific mortality on child survival but further development is needed, especially for the case in which mortality does not increase by the same factor at all ages. 
This is, of course, the case of Covid-19, which is known to affect people in older ages [ref]. 
Further work is also needed to generalize these relationships to other types of kin, such as mothers, grandmothers, cousins and aunts.
A final strand of methodological component of the project will seek to develop the introduce heterogeneity in the Kin-Cohort models. 
In their current state, these models ignore the role of heterogeneity in the experience of kin death, a particular limitation when considering that deaths from the infectious disease are likely to be clustered in the population.

The empirical componente of the project will apply the mathematical models to month-specific mortality and fertility rates (interpolated, if needed, from the available data taking the seasonsality of mortality and fertility into account) to estimate the excess burden of kin death derived from Covid-19.
Combining this with data on the age and sex structure of the population will allow us to estimate excess bereavement, which we will compare to a baseline counter-factual scenario using rates that exclude the excess mortality from Covid-19.
The study will produce a range of estimates, given the uncertainty inherent to the Covid-19 mortality statistics that are required as input for our mathematical models.

As an initial step, the analysis will focus on Germany and Italy, for which quality data on covid-19 death rates [ref] is already available at the sub-national level and where more quality data is expected in the future [ref].
Our methodology is country-agnostic, meaning that it can be applied to other settings once quality data on Covid-19 excess mortality is becomes available.

## Expected main results

- A set of academic publications describing the methodology and the main results of the study
- A flexible and scalable methodology for estimating the excess burden of covid-19 bereavement. The code to produce the estimates will be made freely available on the code-sharing platform [Github](Github.com).
- The first set of estimates of how the Covid-19 disease will affect the expossure to mortality at a population-level
- A online platform for scientists and policy makers to access and download the data easily (see [this interactive app](https://research-app.shinyapps.io/child_death_paa/))

## Work plan

## References

### 2020-05-14 paper.Rmd
commit for Emilio´s revision

---
title: 'Child survival for mothers: mortality change and related measures'
author: "Ivan Williams and Diego Alburez-Gutierrez"
date: "May 2020"
output:
  html_document:
    fig_caption: yes
bibliography: bib.bibtex
---

<!--
Comentarios:
- Puede sonar que la instroducción es un poco larga para el resto del cuerpo. Lo dejo a tu criterio porque me cuesta sacar algo de allí, aprece todo bastante relevante.
- hay dos comentarios más abajo: uno sobre algo que agregaría a la introducción y otra cosa sobre una idea interesante que podríamos agregar como mini corolario (line 217 aprox).
- Perdón por mi spanglish.
- Los gráficos aún requieren trabajo. Quizás sea algo que podamos mejorar post-vista de Emilio.
- Traté de dejar expuesto que la parte de heterogeneidad es solo el planteo.
- Las referencias cruzadas no funcionan en html_output pero tengo alta confianza que lo harán en pdf.
- Resolví que en funs.R se graben los objetos numéricos, y que en cada sesión del markdown se lean. Correr el script cada vez se hace knitr hace poco dinámico el trabajo (tarda mucho, hay formas de catche pero no las exploré en profundidad). Es decir, tenemos solo dos archivos de trabajo: este markdown y funs.R.
-->

```{r setup, include = FALSE}
library(tidyverse)
library(gridExtra)
knitr::opts_chunk$set(collapse = TRUE,  echo = F, message=F, warning=F)
options(digits=2)
load("R/CS_outs.RData")
```

## Introduction

The question of child survival, particularly the survival of daughters for mothers, sits at the very center of demographic theory.
This is evidenced by the omnipresence of the 'net reproductive rate' in accounts of human and non-human populations. 
Historical demographers draw liberally on assumptions about kin availability and individual's exposure to offspring mortality to explain fertility decisions, especially around the demographic transition. 
These assumptions are often untested given the data sparsity [@livi_bacci_concise_1997;@volk_infant_2013]. 
Economic theories of fertility that consider children as 'investment' for old-age support also rely on notions of child survival [@preston1978effects].

Kin count estimation has a long and distinguished pedigree in mathematical demography, starting with the work of @lotka1931orphanhood on modelling orphanhood in theoretical populations across demographic regimes. To the best of our knowledge, @brass_derivation_1953 first proposed a formula to represent child survival over age. The notion of estimating the expected number of living daughters in a stable population was generalized by @goodman1974 for other kin relations (granddaughters, cousins, etc.). The "counting method" approach was further popularized in @KeyfitzMath. @bongaarts_projection_1987 used a similar approach to estimate descendants in his 'Family Status Model'.

Child survival - and it's counterpart, child death - gain new significance in the wake of the fertility transition.
Depressed fertility means that fewer children are tasked with providing key emotional, social, and financial transfers to aging parents for ever increasing periods of time.
With increasing periods of generational overlap, individuals find themselves 'sandwiched' between aging parents and young children requiring their simultaneous attention and care [@daatland_norwegian_2010].
In the context of global population aging, elderly parents without access to formal social security and pension systems are particularly reliant on these transfers to make ends meet [@smith-greenaway_maternal_2020].  
Losing an only child may become a more common experience in the context of global fertility decline, something particularly worrying for parents suddenly rendered childless.
Given the known psychological, health, and social consequences of child loss for bereaved parents and families [@Hendrickson2009;@lee_death_2014], it is surprising that the demographic processes that shape parental bereavement remain very poorly understood.
Recent development have underscored the need to understand how sudden changes in mortality affect the availability of kin. 
In the context of the global pandemic caused by the Covid-19 disease, elderly people depend on support provided by their relatives, especially during periods of lockdown and in the absence of governmental support mechanisms.<!-- Maybe here are repeated --> Elderly parents may be at a higher risk of losing the key support provided by their adult children given the known age-gradient in the Covid-19 case fatality rate. 
How do changes in mortality affect the availability of offspring over age from the point of view of a prospective mother? How can we characterize the timing of offspring mortality as experienced by a mother from a formal demographic perspective? How this would work in an heterogeneity population?
In the remainder of the paper, we aim to formalize the relationship population-level changes in mortality rates and perceived change in the lived experience of child death from the perspective of a mother. Our approach is key for understanding the impact of age-specific excess mortality on the resilience or otherwise of kinship networks for the vulnerable elderly population.
We consider the case when these changes are additive and mutliplicative. In this paper, we uncover formal relationships to answer these questions and provide numerical applications using data from countries in Latin America during the second half of the twentieth century. See the [Applications](#Applications) section for more details on the source, calculation procedures and main numerical results. We used period measures, in an stable context, so the objetive is give moment measures. A more sophisticated model using the appropriate age-specific cohort rates in the subscripts has been proposed by @alburez_gutierrez_womens_2019.

<!-- In some place (maybe here) I would say something like this: "
These work can be read like an humble extension of the kin expressions developed by formal demograperhs mentioned before, motivated by the study of beaverement experience in mothers.
(como una forma de decir: miren lo dificil ya esta hecho, nosotros continuamos un camino, dando algun que otro pequeño paso motivados por...)" -->

## Relationship (and proofs)

### Child Survival

Let $CS_a$ be the expected number of surviving children to a mother^[We use woman and mother interchangeably in this paper, assuming that all women are exposed to the same fertility rates.] alive at aged $a$ in a female stable population with fertility rates $m_{x}$, mortality hazard $\mu_{x}$ and survival function $l_x=e^{-\int_{0}^{x}{\mu_t}\,dt}$ (with unit radix $l_0=1$), as proposed by @goodman1974:

$$CS_a = \int_{0}^a{m_{x}l_{a-x}dx}$$

### An approximation

As an initial step, we seek an intuitive understanding of the effect of changing mortality on child survival. 
Building on work by @KeyfitzMath for the probability of a living mother, we can show which features of the child survival function explain the expected number of children. 
For this we use an approximate of $l_x$ using Taylor´s theorem until second order around the mean age of childbearing $\kappa$:

$$CS_a \approx l_{a-\kappa} \int_{0}^a{m_{x}dx}+(l_{a-\kappa})^{'}\int_{0}^a{(x-\kappa)m_{x}dx}+
(l_{a-\kappa})^{''}\int_{0}^a{\frac{(x-\kappa)^2}{2} m_{x}dx}\\
  \approx F_a l_{a-\kappa} + \frac{\sigma^2}{2}F_a (l_{a-\kappa})^{''}\\
  \approx F_a \, l_{a-\kappa}[1+\frac{\sigma^2}{2}\frac{(l_{a-\kappa})^{''}}{l_{a-\kappa}}]$$

where the fertility pattern by age is concentrated around $\kappa$ and the accumulated fertility (or gross reproduction rate in our female-dominant scenario) is $F_a =  \int_{\alpha}^a{m_{x}dx}$. The second Taylor´s term is null because $\int_{\alpha}^a{x \, m_{x}dx} = \kappa \, F_a$.

We find that, seen from the perspective of a mother, child survival mainly depends on the cummulative fertility function^[Different from the findings of @KeyfitzMath of having living ancestors.] and the survival from birth to $a-\kappa$. Also that the approximation is affected (negativally) by how disperse is the fertility by age (variance $\sigma^2$) and the suvival curvature around (with negative sign), in an age range where $l_x$ is tipycally very flat (range 20-40 years old) in transitioned populations^[This approximation is useful also for get a dimension changes in the fertility average age \kappa: taking logs in $F_a \, l_{a-\kappa}$ and deriving we get $\frac{\Delta CS_a}{CS_a} \approx - \mu_{a-\kappa} \Delta\kappa$].

### Mortality Changes

We now consider the consequences of an absolute change $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk.
For now, let us consider $m_{x,\delta}=m_{x}+\delta$ (@wrycza2012) and $l_{a-x}^\delta = e^{-\int_{0}^{a-x}{(\mu_t+\delta})dt}$:

$$CS_{a}^\delta = \int_{\alpha}^{a}m_{x} l_{a-x} e^{-\delta (a-x)} dx$$

We get the derivative of $dCS_{(a)}^\delta / d\delta$ evaluated near zero (@KeyfitzMath) to find the effects of adding $\delta$ to hazard rates:

$$\frac{dCS^{\delta}}{d\delta} = -a\int_{\alpha}^{a}m_{x}{l_{a-x} dx} + \int_{\alpha}^{a} m_{x} {l_{a-x}x dx}\\
= -a \, CS_{a}  + \int_{\alpha}^{a} {x  \, m_{x} l_{a-x} dx}$$

Dividing both sides by $CS_{a}$ in an discrete approximation we get:

$$\frac{\Delta CS_{a}}{CS_{a}} \approx -(a - k_a) \Delta\delta$$

The expected change in descendants survival is inversely proportional to the difference between maternal age $a$ and the mean age of the mother at the birth of her surviving daughters $k_a$. The magnitude of the change depends negatively on the age distribution of the surviving offspring. This is intuitive considering that younger descendants experiences longer periods of exposure to risk.

That an equal change in mortality could happens at all ages is unlikelihood. This measure could be much related a change in first ages (child and youth) than adult or ones. As a corollary, if the change only happens in the age range $\left[0;1\right)$, infant mortality, we can inspect the effect starting with splitting the integral (considering those born last year and not, and of course for $a>1$):

$$CS_{a}^{\delta_0} = \int_{0}^{a-1} m_{x} {e{^{-\int_{1}^{a-x}\mu_t \,dt-\int_{0}^{1}(\mu_t + \delta)\,dt}}} dx + \int_{a-1}^{a} m_{x} e^{-\int_{0}^{a-x}(\mu_t+ \delta)\,dt}dx\\
=\int_{0}^{a-1} m_{x} l_{a-x}e^{-\delta} dx + \int_{a-1}^{a} m_{x} l_{a-x} e^{-\delta(a-x)}\,dx$$

Deriving by $\delta$ and valuating near 0, we get:

$$CS_{a}^{\delta_0} =-\int_{0}^{a-1} m_{x} l_{a-x} dx - a \int_{a-1}^{a} {m_{x} l_{a-x} dx} + \int_{a-1}^{a} {x \, m_{x} l_{a-x} dx}$$
If we name $CS_{a-1,a}=\int_{a-1}^{a} {m_{x} l_{a-x} dx}$, we can express the first term in the right as $CS_a - CS_{a-1,a}$, and finally divide by $CS_a$: 
$$ = - CS_a + CS_{a-1,a} (1-a+\kappa_{a-1,a})\\
\frac{\Delta CS_{a}}{CS_{a}} \approx - \left[1-\frac{CS_{a-1,a}}{CS_a} (1-(a-\kappa_{a-1,a}))\right] \Delta\delta$$

This means that an absolute change in infant hazard rate affects proportionally for all the age range extension (because of all the cohorts aged more than 1), except for the mean time exposure for those who born and survive in the last year $1-(a-k_{a-1,a})$, times $\frac{CS_{a-1,a}}{CS_a}$ the portion of cummulative alive daughters for the same period, weighting how important was this in total successful experience (in terms of alive descendants).

We now consider the consequences of a proportional change in mortality on child survival.
Given a proportional change in mortality $\mu_{x,\delta}=\mu_{x}(1+\delta)$ so that $l_{a-x}^\delta = e^{-\int_{0}^{a-x}{\mu_t(1+\delta})dt}=(l_{a-x})^{(1+\delta)}$, it follows that:

$$CS_{a}^\delta = \int_{\alpha}^{a} {m_{x} l_{a-x}^{(1+\delta)}} dx.$$

Using the derivative $\frac{dl_{a-x}^{(1+\delta)}}{d\delta} = log(l_{a-x}) l_{a-x}^{(1+\delta)}$, and in the third row reversing integrals between $t$ and $x$:

$$\frac{d CS_{a}^\delta}{d \delta} = \int_{0}^{a} {m_{x} l_{a-x} \log(l_{a-x}) dx}\\
= - \int_{0}^{a} {m_{x} l_{a-x} \int_{0}^{a-x}{\mu_t \, dt}\, dx}\\
= - \int_{0}^{a}{\mu_t} \int_{0}^{a-t} {m_{x} l_{a-t-x} \, dx}\, dt$$

The last integral is $CS_{a-t}$. Dividing by $CS_a$ and multiplying by $\delta$ get one useful expression. This is the negative of cumulative hazard $H_a$ but considering a positive factor $\frac{CS_{a-t}}{CS_a}$ that takes in account the relative amount of surviving descendants that would be lost at each child age at risk.

$$\frac{\Delta CS_{a}}{CS_a} \approx - \left[\int_{0}^{a}{\mu_x  \, \frac{CS_{a-x}}{CS_a} \,dx \,}\right] \Delta \delta$$

Grasping this relationship intuitively is more difficult given the interaction of birth and mortality rates (@KeyfitzMath). 
This factor has a S-shape because $l_x$ curvature and fertility accumulation, that gives more weight to first ages, generally less than 1 except for mother ages close to age valuation *a* with complete reproductive life and high infant mortality (no additional children and high risk for one age additional of survive), typically pretransitional countries with a near 50. A numerical evaluation of both analytical expressions (absolute an relative) are done in [Applications](#Applications) section.

## Related Measures

### Burden and timing of maternal bereavement

We call *Mean Time Spent in Bereavement* (*MTSB*) the absolute measure of the expected total lifetime with a death daughter for a mother aged *a*, which can be expressed in terms of a temporary expected lost years index, in line with $e^\dagger$ (@Vaupel1986):

$$MTSB_a = \int_0^a{m_x\int_0^{a-x}{d_t e_{0|a-x-t}\, dt}\,dx} = \int_\alpha^a{m_x e_{0|a-x}^\dagger \,dx}$$.

Where $d_t$ is the death distribution from birth, $e_{0|a-x-t}$ is the life expectancy at birth until age $a-x-t$ and $e_{0|a-x}^\dagger$ the temporary dispersion measure. But most interesting would be to compare these years with the time that mothers would expect to live with their daughters. We call this the *Intensity Time in Bereavement* ($ITB$): a ratio between expected time with a "lost" and expected time with a "life", that allows to make comparisons between population regimes.

$$ITL_a = \frac{\int_0^a{m_x e_{0|{a-x}}^\dagger dx}}{\int_0^a{m_x e_{0|a-x} dx}}$$

This is a ratio between child-years in two radically different states. Looks similar to the transcendental entropy measure $H$ (@KeyfitzMath) but considering all the cohorts born during the mother´s life, weighted by their relative size $m_x$. In the figure \@ref(fig:plot_ITL_MAL) is shown that *ITB* is bigger for young women because of the weight of infant mortality in their mother experience, also with more dispersion for same levels of parity.

Another important factor for the child survival experience of mothers, is the mean age at child loss, called here $MAL$. This relation can be derived by starting with the mother age $x+t$ at each death child age $t$ at death, weighted by the fertility and survival function. In it, $\kappa$ is the mean age at childbirth for women aged $a$, $MAD_{a-x}$ refers to the mean age at death for newborns that die before $a-x$ and $F_a$ is the accumulated fertility for a women aged $a$:

$$MAL_a = \frac{\int_0^a{m_x {\frac{\int_0^{a-x} l_t \, \mu_t (x+t) dt}{\int_0^{a-x} l_t \, \mu_t dt}}dx}}{\int_0^a{m_x}dx} \\
  MAL_a = \frac{\int_0^a{m_x \left[{x+\frac{\int_0^{a-x} l_t \, \mu_t \, t \, dt}{\int_0^{a-x} l_t \, \mu_t dt}}\right]dx}}{\int_0^a{m_x}} \\
  MAL_a = \frac{\int_0^a{m_x \,x \,dx}}{F_a} + \frac{\int_0^a{m_x MAD_{a-x}dx}}{F_a} \\
  MAL_a = \kappa + \frac{\int_0^a{m_x MAD_{a-x}dx}}{F_a}$$

A numerical approximation of both measures are done in [Applications](#Applications) section.

### Heterogeneity (first ideas, some guidance is needed)

In a model with heterogeneity, determined at birth by a multiplicative effect, $CS_a$ could be interpreted as a conditional expectation with random variables *K* for fertility heterogeneity and *Z* for mortality frailty.

$$CS_a(k,z) = \int_{0}^{\infty}\,{m(k)}_x\,{l(z)}_{a-x}\,dx$$

Following @Coresh1988 we can express the fertility part in a multiplicative way with variability in the level but not in the shape, as ${m(k)}_x = F_\beta\,k\,r_x$, where $F_\beta$ is the baseline cumulated fertility until upper limit in all age range until $\beta$, $k$ is a random variable with mean 1 that allows variability between groups and $r_x$ is the fertility structure by age ($\int_{0}^{\beta}{r_x}=1$). 
In the other side, frailty part can be thought in a cohort effect way, also in the multiplicative assumption as @Vaupel2014, with $l(z)_x=e^{-H_{x}z}=l_x^z$, with baseline hazard $\mu$.
Considering the joint distribution $f_{kz}$, the intuition could be expressed as  $f_{zk}=f_{k|z}f_z$: groups with higher descendant´s mortality would adjust their fertility level, with positive correlation.
Replacing both variables, the unconditional mean $\overline{CS}_a$ would be^[Note that given k and z values, fertility and child survival are independent.]:

$$\overline{CS}_a = \int_{0}^{\infty}\int_{0}^{\infty}\left[\int_{0}^{a}{ {m(k)}_x l(z)_{a-x}\,dx}\right]f_{kz}\,dz \,dk\\
= F_\beta \int_{0}^{a-x} r_x \int_{0}^{\infty} \int_{0}^{\infty} {(l_{a-x})^z} \, k\, f_{kz} \,dk\,dz\,dx$$

Isolating for an age *x* we can inspect the part related to z and k: $(l_{a-x})^z \, k\, f_{kz}$. 
Following @Gupta2018, one way to create a correlated bivariate distribution is assume for example that $Z = Y_0 + Y_1$ and $K = e^{(Y_0 + Y_2)}$, given the fact that the historical relation is not linear between both components (see figure \@ref(fig:plot_tfr_q0), a funtion *f* like $F_\beta=f(ln(q_{0,5}))$. The auxiliar variables $Y_0$, $Y_1$ and $Y_2$ are independet and Gamma distribuited. Then we can express, using Laplace transform properties:

$$\int_{0}^{\infty} \int_{0}^{\infty} {(l_{a-x})^z \, k\, f_{kz} \,dk\,dz}\\
\int_{0}^{\infty} \int_{0}^{\infty}{e^{-H_x(Y_0 + Y_1)}e^{Y_0 + Y_2}}\\
L_{Y_0}[1-H_x]L_{Y_1}[-H_x]L_{Y_2}\,dY_0\,dY_1\,dY_2$$

...

*What can we look for here?*:
* _a formalization of covariance_

* _estimate joint distribution using historical data. Idea: ¿maybe taking all the period/country observations as realized values of an hypothetical stable population, and using bayesian procedures?_

## Applications {#Applications}
### Data {#Data}

We motivate this paper with an empirical example using fertility and mortality rates from the 2019 Revision of the UN World Population Prospects (UN WPP) for the Latin American Region (@WPP19). We smoothed female $l_x$ in quinquennial ages, using cubic-splines constrained to monotonic decrease, taking $L_0$ and $T_{100}$ from raw life tables as inputs for year-person calculations. For splitting fertility five groups was used quadratic optimization approach by @Michalski2018, with an desirable property for our purpose which is a good fitting in parity. Also was assumed an unique female percentage of newborns of `r round(im, 4)` for all period-country cases. Calculations were done in a discrete way assuming that the $m_x$ live births are born at exact mother´s age *x*.
An already known log relation between infant mortality and TFR is shown in Latinoamerican experience (\@ref(fig:plot_tfr_q0)).

<!-- Specific paths like Argentina and Guatemala shows very different demographic profiles during second part of XX century. -->

```{r plot_tfr_q0, fig.cap="Female probability of death at birth by gross reproduction rate. Latin America countries in period 1950-2015"}

ggplot(LA %>% filter(x==0, name!="Latin America and the Caribbean"), 
                  aes(x=qx, y=tfr))+
                  geom_point(aes(color=Year)) +
                  geom_smooth(se = F, color = "grey", method = "lm",
                              formula = y ~ log(x)) +
                  geom_line(data = LA %>% filter(x==0, name=="Latin America and the Caribbean"),
                            aes(x=qx, y=tfr, color=Year), size=1)+
                  theme_classic() + labs(x="q(0)", y="GRR") + 
                  theme(legend.position = "none") 
                  # geom_line(data = LA %>% filter(x==0, name %in% c("Guatemala")),
                  #           aes(group=name), colour=2, size = 1) +
                  # geom_line(data = LA %>% filter(x==0, name %in% c("Argentina")),
                  #           aes(group=name), colour=4, size = 1)
```

### Numerical Results

Consider a Latin American woman standing before us.
If this were 1950-1955, she could reasonable expect to have `r mean(CS_outs$CS[CS_outs$Period == "1950-1955" & CS_outs$Age==50])` surviving daughters on her 50$^{th}$ birthday.
In 2010-2015 a woman the same age would only have `r mean(CS_outs$CS[CS_outs$Period == "2010-2015" & CS_outs$Age==50])` living children. 
The difference of `r mean(CS_outs$CS[CS_outs$Period == "1950-1955" & CS_outs$Age==50])-mean(CS_outs$CS[CS_outs$Period == "2010-2015" & CS_outs$Age==50])` children is explained by reduced fertility and improved mortality in the region [give stats]. 
We now remove the effect of changing fertility by considering the number of daughters surviving up to maternal age $a$ as a proportion of the daughters ever born to a woman that age, $\frac{CS_a}{F_a}$. 
Given Eq. \ref{eq:CS}, the increase from `r mean(CS_outs$CS_prob[CS_outs$Period == "1950-1955" & CS_outs$Age==50])` in 1950-1955 to `r mean(CS_outs$CS_prob[CS_outs$Period == "2010-2015" & CS_outs$Age==50])` in 2010-2015 must be explained by a change in mortality.

<!-- this next paragrap es una idea.can be easily solve with optimization. Let´s discuss what you think.
$CS_a$ is usually an increasong function of age, with a pick An interesting corollary of these results is the question: at wich ages a mother would have the same *CS*, like a mirror effect? if we consider age *a*, the question could be answered by solving numerically *y* for: $\int_{0}^{a}{m_xl_{a-x}}dx=\int_{0}^{y}{m_xl_{y-x}}dx$, or with our approximation estimating *y* that $\frac{F_a}{F_y} = \frac{l_{a-\kappa_a}}{l_{y-\kappa_y}}$.
```{r, include=F}
# Por ejemplo... Una mujer de Bolivia en 1950-1955 esperaba tener la misma cantidad de hijos vivos a los 70 que a los 37.
CS_outs %>% filter(Period=="1950-1955", name=="Bolivia") %>% select(Age,CS)
```
End idea-->

```{r plot_CS, fig.cap="Child Survival and Child survival as a share of cummulative fertility by age, for women aged 30, 40 and 50. Estimates using UN WPP data for Latin American countries in the period 1950-2015 period"}
a <- ggplot(CS_outs %>% filter(name!="Latin America and the Caribbean"), 
            aes(x=Year, y=CS, color=factor(Age)))+
            geom_point(alpha = .3) + 
            geom_line(data = CS_outs %>% filter(name=="Latin America and the Caribbean"), 
                      aes(x=Year, y=CS, color=factor(Age))) +
            theme_classic() +
            labs(y = "CS") + theme(legend.position = "none")
b <- ggplot(CS_outs %>% filter(name!="Latin America and the Caribbean"), 
            aes(x=Year, y=CS_prob, color=factor(Age)))+
            geom_point(alpha = .3) + 
            geom_line(data = CS_outs %>% filter(name=="Latin America and the Caribbean"), 
                      aes(x=Year, y=CS_prob, color=factor(Age))) +
            theme_classic() +
            labs(y = "CS/F") + 
            scale_colour_discrete(name="Age")
grid.arrange(a,b,ncol=2)
```

The approximation proposed for CS_a is reaalyprecise, especially for recent periods. For exaple, Guatemala improved the approximation with years due to rectangularization process in $l_x$ (figure \@ref(fig:plot_CS_aprox)).

```{r plot_CS_aprox, fig.cap = "Left: Error in approximation for a=40, from years 1950 (darker) to 2015 (lighter) for all Latin American countries (blue) and Guatemala (red). Right: Change in survival and fertility by age in Guatemala."}
age = 40
CS_app <- ggplot(CS_outs %>% filter(Age == age, name!="Latin America and the Caribbean"), 
            aes(x=CS, y=error_aprox, color=Year))+
            geom_point(alpha = .6) + 
            geom_point(data = CS_outs %>% filter(name=="Guatemala",Age == age),
                       aes(x=CS, y=error_aprox), color="red") +
            geom_line(data = CS_outs %>% filter(Age == age, name=="Latin America and the Caribbean"), 
                      aes(x=CS, y=error_aprox), color="magenta1") +
            geom_hline(yintercept = 0, color="grey", linetype = 2) +
            theme_classic() +
            theme(legend.position = "none") +
            labs(y = "%")
Guate_lt <- LA %>% filter(name=="Guatemala") %>% ggplot() +
            geom_line(aes(x=x, y=lx/100000, color=Period)) +
            theme_classic() +
            scale_color_grey()+
            theme(legend.position = "none") +
            labs(y = "")
Guate_mx <- LA %>% filter(name=="Guatemala", between(x, 15,49)) %>% ggplot() +
            geom_line(aes(x=x, y=asfr/tfr*100, color=Period)) +
            theme_classic() +
            scale_color_grey()+
            theme(legend.position = "none") +
            labs(y = "%")
grid.arrange(CS_app, grid.arrange(Guate_lt,Guate_mx,ncol=1), ncol=2)
```

The figure \@ref(fig:CS_abs_app) shows that goodness of fit is decreasing with the change size, given that $\delta$ is assumed near zero. To ilustrate that considering $\delta=0.01$ means at age 50, it means for the absolute change case that $\frac{l_{50}}{e^{\delta\,50}} = l_{50}/$`r round(exp((50*0.01)),1)`, a big change). For the relative case was assumed a constant within each interval $\mu_{x+t}=\mu_x$ for $t$ between 0 and 1, $\int_{x}^{x+1}{\mu_t dt} = \log(l_{x+1})-\log(l_{x})$, and was used this approximation on the empirical data $\frac{\Delta CS_{a}}{CS_a} \approx - \left[\sum_{0}^{a-1}{[\log(l_{x+1})-\log(l_{x})]\, \frac{CS_{a-1-x}}{CS_a}\,}\right] \Delta \delta$.

```{r CS_abs_app, fig.cap= "Effect of an absolute absolute change in mortality on child survival by maternal age of women in Latin America  for the 1950-2015 period. Goodness of fit of approximation compared to direct estimation"}

ch_abs <-  CS_change_app %>%
              ggplot() +
             geom_point(aes(x=change_obs_abs, y=change_app_abs, color=delta)) +
             geom_abline(slope=1, intercept = 0) +
             scale_x_continuous("Empirical change in CS") +
             scale_y_continuous("Approximated change in CS") +
             scale_color_continuous("Absolute change\nin delta") +
             theme_classic()

ch_rel <-  CS_change_app %>%
              ggplot() +
             geom_point(aes(x=change_obs_rel, y=change_app_rel, color=delta)) +
             geom_abline(slope=1, intercept = 0) +
             scale_x_continuous("Empirical change in CS") +
             scale_y_continuous("Approximated change in CS") +
             scale_color_continuous("Relative change\nin delta") +
             theme_classic()

grid.arrange(ch_abs, ch_rel, ncol=2)

```

Assuming constant period rates the expected time in years that a mother aged 30 passed with a death son was around 4% in some countries at middle XX Century. When increasing age, the survival experience depends less on infant mortality, and the distribution is around less than 2% for women aged 50, converging to 0 on time. As extreme cases, in 1950-1955 Haiti women aged 30 would have experienced an intensity of `r round(CS_outs$ITL[CS_outs$name=="Haiti" & CS_outs$Age==30 & CS_outs$Period=="1950-1955"],1)`%, and in 2010-2015 `r round(CS_outs$ITL[CS_outs$name=="Haiti" &CS_outs$Age==30 & CS_outs$Period=="2010-2015"],1)`%, while the women of Costa Rica `r round(CS_outs$ITL[CS_outs$name=="Costa Rica" & CS_outs$Age==30 & CS_outs$Period=="1950-1955"],1)`%, and in 2010-2015 `r round(CS_outs$ITL[CS_outs$name=="Costa Rica" &CS_outs$Age==30 & CS_outs$Period=="2010-2015"],1)`%.
For those Latin American mothers aged 50 with 3 daughters born who suffered a lost, they experienced that at age 30 in average (figure \@ref(fig:plot_ITL_MAL)).

```{r plot_ITL_MAL, fig.cap = "a) Intensity Tome Lost of  women aged 30, 40 and 50. b) Mean age at Lost for women aged 40. Years 1950 (darker) to 2015 (lighter). Latinamerican countries in period 1950-2015"}
ITL <- ggplot(CS_outs %>% filter(name!="Latin America and the Caribbean"), 
          aes(x=Fa, y=ITL, color=factor(Age)), alpha = .3) + # alpha not working
          geom_point() +
          geom_line(data = CS_outs %>% filter(name=="Latin America and the Caribbean"), 
                    aes(x=Fa, y=ITL, color=factor(Age)), size =1)+
          theme_classic() +
          theme(legend.position = "right") +
          labs(y = "%")+
          scale_color_discrete(name = "Age")

age = 50
MAL <- ggplot(CS_outs %>% filter(Age==age, name!="Latin America and the Caribbean"))+
        geom_point(aes(x=Fa, y=MAL, color=Year), alpha = .3) +
        geom_line(data = CS_outs %>% filter(Age==age, name=="Latin America and the Caribbean"),
                  aes(x=Fa, y=MAL, color=Year),size= 1) +
        theme_classic()+
        theme(legend.position = "none") +
        labs(y = "Mean Age at Lost")
grid.arrange(ITL, MAL, ncol=2)
```

## References

### 2020-05-14 paper.tex
commit for Emilio´s revision

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Child survival for mothers: mortality change and related measures},
  pdfauthor={Ivan Williams and Diego Alburez-Gutierrez},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {\setlength{\parindent}{0pt}%
  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
  {\par}

\title{Child survival for mothers: mortality change and related
measures}
\author{Ivan Williams and Diego Alburez-Gutierrez}
\date{07 August 2019}

\begin{document}
\maketitle

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

Kin count estimation has a long and distinguished pedigree in
mathematical demography, starting with the work of Lotka (1931) on
modelling orphanhood in theoretical populations across demographic
regimes. To the best of our knowledge, Brass (1953) first proposed a
formula to represent child survival over age. The notion of estimating
the expected number of living daughters in a stable population was
generalized by Goodman, Keyfitz, and Pullum (1974) for other kin
relations (granddaughters, cousins, etc.). The ``counting method''
approach was further popularized in Keyfitz and Caswell (2005).
Bongaarts (1987) used a similar approach to estimate descendants in his
`Family Status Model'.

The question of child survival, particularly the survival of daughters
for mothers, sits at the very center of demographic theory. This is
evidenced by the omnipresence of the `net reproductive rate' in accounts
of human and non-human populations. Historical demographers draw
liberally on assumptions about kin availability and individual's
exposure to offspring mortality to explain fertility decisions,
especially around the demographic transition. These assumptions are
often untested given the data sparsity (Livi Bacci 1997; Volk and
Atkinson 2013). Economic theories of fertility that consider children as
`investment' for old-age support also rely on notions of child survival
(Preston and others 1978).

Child survival - and it's counterpart, child death - gain new
significance in the wake of the fertility transition. Depressed
fertility means that fewer children are tasked with providing key
emotional, social, and financial transfers to aging parents for ever
increasing periods of time. With increasing periods of generational
overlap, individuals find themselves `sandwiched' between aging parents
and young children requiring their simultaneous attention and care
(Daatland, Veenstra, and Lima 2010). In the context of global population
aging, elderly parents without access to formal social security and
pension systems are particularly reliant on these transfers to make ends
meet (Smith-Greenaway and Trinitapoli 2020).\\
Losing an only child may become a more common experience in the context
of global fertility decline, something particularly worrying for parents
suddenly rendered childless. Given the known psychological, health, and
social consequences of child loss for bereaved parents and families
(Hendrickson 2009; Lee et al. 2014), it is surprising that the
demographic processes that shape parental bereavement remain very poorly
understood. Recent development have underscored the need to understand
how sudden changes in mortality affect the availability of kin. In the
context of the global pandemic caused by the Covid-19 disease, elderly
people depend on support provided by their relatives, especially during
periods of lockdown and in the absence of governmental support
mechanisms. Elderly parents may be at a higher risk of losing the key
support provided by their adult children given the known age-gradient in
the Covid-19 case fatality rate. How do changes in mortality affect the
availability of offspring over age from the point of view of a
prospective mother? How can we characterize the timing of offspring
mortality as experienced by a mother from a formal demographic
perspective? How this would work in an heterogeneity population? In the
remainder of the paper, we aim to formalize the relationship
population-level changes in mortality rates and perceived change in the
lived experience of child death from the perspective of a mother. Our
approach is key for understanding the impact of age-specific excess
mortality on the resilience or otherwise of kinship networks for the
vulnerable elderly population. We consider the case when these changes
are additive, mutliplicative, and when they are not evenly distributed
across age. In this paper, we uncover formal relationships to answer
these questions and provide numerical applications using data from
countries in Latin America during the second half of the twentieth
century. See the \protect\hyperlink{Data}{data section} in the
\protect\hyperlink{Applications}{Applications} section for more details
on the source, calculation procedures and main results. We used period
measures, in an stable context, so the objetive is give moment measures.
A more sophisticated model using the appropriate age-specific cohort
rates in the subscripts has been proposed by Alburez-Gutierrez, Kolk,
and Zagheni (2019).

\hypertarget{relationship-proof}{%
\subsection{Relationship \& Proof}\label{relationship-proof}}

\hypertarget{child-survival}{%
\subsubsection{Child Survival}\label{child-survival}}

Let \(CS_a\) be the expected number of surviving children to a
mother\footnote{We use woman and mother interchangeably in this paper,
  assuming that all women are exposed to the same fertility rates.}
alive at aged \(a\) in a female stable population with fertility rates
\(m_{x}\), mortality hazard \(\mu_{x}\) and survival function
\(l_x=e^{-\int_{0}^{x}{\mu_t}\,dt}\) (with unit radix \(l_0=1\)), as
proposed by Goodman, Keyfitz, and Pullum (1974):

\[
\begin{aligned}
CS_a = \int_{0}^a{m_{x}l_{a-x}dx}
\end{aligned}
\]

\hypertarget{an-approximation}{%
\subsubsection{An approximation}\label{an-approximation}}

As an initial step, we seek an intuitive understanding of the effect of
changing mortality on child survival. Building on work by Keyfitz and
Caswell (2005) for the probability of a living mother, we show which
features of the child survival function explain the expected number of
children. For this we use an approximate of \(l_x\) using Taylor´s
theorem until second order around the mean age of childbearing
\(\kappa\) for all \(a>=\alpha\):

\[CS_a \approx l_{a-\kappa} \int_{\alpha}^a{m_{x}dx}+(l_{a-\kappa})^{'}\int_{0}^a{(x-\kappa)m_{x}dx}+
(l_{a-\kappa})^{''}\int_{0}^a{\frac{(x-\kappa)^2}{2} m_{x}dx}\\
  \approx F_a l_{a-\kappa} + \frac{\sigma^2}{2}F_a (l_{a-\kappa})^{''}\\
  \approx F_a \, l_{a-\kappa}[1+\frac{\sigma^2}{2}\frac{(l_{a-\kappa})^{''}}{l_{a-\kappa}}]\]

where the fertility pattern by age is concentrated around \(\kappa\) and
the accumulated fertility (or gross reproduction rate in our same-sex
model) is \(F_a = \int_{\alpha}^a{m_{x}dx}\). The second Taylor´s term
is null because \(\int_{\alpha}^a{x \, m_{x}dx} = \kappa \, F_a\).

We find that, seen from the perspective of a mother, child survival
mainly depends on the cummulative fertility function\footnote{Note that
  this runs contrary to the findings of Keyfitz and Caswell (2005), who
  found that levels of fertility did not affect the probability of
  having living ancestors.} and the mean age at childbearing \(\kappa\)
acts through the factor \(l_{a-\kappa}\). The dispersion of of fertility
\(\sigma^2\) is negatively proportional to child survival. All else
constant, a higher dispersion of fertility \(\sigma^2\) is associated
with an increase in \(CS_a\) during reproductive age. The negative
relationship stems from the fact that the second derivative of
\(l_{a-\kappa}\) is negative in the 15-49 age range, given a concave
downwards survival function.

The second Taylor´s term is null because
\(\int_{\alpha}^a{x \, m_{x}dx} = \kappa \, F_a\). The formula tells us
that the survival experience for a woman aged \emph{a} depends mainly of
cumulated fertility (possible cases, different from the ancestor count
case where this doesn´t affect) and child survival probability (success
cases). Also that the approximation is affected by how disperse is the
fertility by age (variance \(\sigma^2\)) and the suvival curvature
around (with negative sign), in a \(l_x\) range where is tipycally very
flat (range 20-40 years old) in transitioned populations.

And change in \(\kappa\), taking logs and deriving:
\(\frac{dCS_a}{d\kappa}\frac{1}{CS_a} \approx \mu_{a-\kappa}F_a \Delta\kappa\)

\hypertarget{mortality-changes}{%
\subsubsection{Mortality Changes}\label{mortality-changes}}

We now consider the consequences of an absolute change \(\delta\) in
mortality in the range \([0,a-\alpha]\), where \(\alpha\) is the start
age of fertility risk. As a \protect\hyperlink{Corollary}{corollary} we
consider the case where mortality only changes for infants, considering
that it is unlikelily for a potential change in mortality to affect all
ages equally in real populations. For now, let us consider
\(m_{x,\delta}=m_{x}+\delta\) (Wrycza and Baudisch (2012)) and
\(l_{a-x}^\delta = e^{-\int_{0}^{a-x}{(\mu_t+\delta})dt}\):

\[CS_{a}^\delta = \int_{\alpha}^{a}m_{x} l_{a-x} e^{-\delta (a-x)} dx\]

We get the derivative of \(dCS_{(a)}^\delta / d\delta\) evaluated near
zero (Keyfitz and Caswell (2005)) to find the effects of adding
\(\delta\) to hazard rates:

\[\frac{dCS^{\delta}}{d\delta} = -a\int_{\alpha}^{a}m_{x}{l_{a-x} dx} + \int_{\alpha}^{a} m_{x} {l_{a-x}x dx}\\
= -a \, CS_{a}  + \int_{\alpha}^{a} {x  \, m_{x} l_{a-x} dx}\]

Dividing both sides by \(CS_{a}\) in an discrete approximation we get

\[\frac{\Delta CS_{a}}{CS_{a}} \approx -(a - k_a) \Delta\delta\]

The expected change in descendants survival is inversely proportional to
the difference between maternal age \(a\) and the mean age of the mother
at the birth of her surviving daughters \(k_a\) (always smaller than
\(\kappa_a\)). The magnitude of the change depends negatively on the age
distribution of the surviving offspring. This is intuitive considering
that younger descendants experiences longer periods of exposure to risk.
The figure @ref(fig:CS\_abs\_app) shows that goodness of fit is
decreasing with the change size, given that \(\delta\) is assumed near
zero (as an example considering \(\delta=0.01\) means at age 50 that
\(\frac{l_50}{e^{\delta\,50}} = \frac{l_50}{1.6}\), a big change).

That an equal change in mortality could happens at all ages is
unlikelihood. This measure is much more related to first ages (child and
youth) than adult or ones. As a corollary, if the change only happens in
the age range \(\left[0;1\right)\), infant mortality, we can inspect the
effect starting with splitting the integral (of course, for \(a>1\)){]}:

\[CS_{a}^{\delta_0} = \int_{0}^{a-1} m_{x} {e{^{-\int_{1}^{a-x}\mu_t \,dt-\int_{0}^{1}(\mu_t + \delta)\,dt}}} dx + \int_{a-1}^{a} m_{x} e^{-\int_{0}^{a-x}(\mu_t+ \delta)\,dt}dx\\
=\int_{0}^{a-1} m_{x} l_{a-x}e^{-\delta} dx + \int_{a-1}^{a} m_{x} l_{a-x} e^{-\delta(a-x)}\,dx\]

Deriving by \(\delta\) and valuating near 0, we get:

\[CS_{a}^{\delta_0} =-\int_{0}^{a-1} m_{x} l_{a-x} dx - a \int_{a-1}^{a} {m_{x} l_{a-x} dx} + \int_{a-1}^{a} {x \, m_{x} l_{a-x} dx}\]
If we name \(CS_{a-1,a}=\int_{a-1}^{a} {m_{x} l_{a-x} dx}\), we can
express the first term in the right as \(CS_a - CS_{a-1,a}\), and
finally divide by \(CS_a\):
\[ = - CS_a + CS_{a-1,a} (1-a+\kappa_{a-1,a})\\
\frac{\Delta CS_{a}}{CS_{a}} \approx - \left[1-\frac{CS_{a-1,a}}{CS_a} (1-(a-\kappa_{a-1,a}))\right] \Delta\delta\]

This means that an absolute change in infant hazard rate affects
proportionally for all the age range extension (because of all the
cohorts aged more than 1), except for the mean time exposure for those
who born and survive in the last year \(1-(a-k_{a-1,a})\), times
\(\frac{CS_{a-1,a}}{CS_a}\) the portion of cummulative alive daughters
for the same period, weighting how important was this in total
successful experience (in terms of alive descendants).

We now consider the consequences of a proportional change in mortality
on child survival. Given a proportional change in mortality
\(\mu_{x,\delta}=\mu_{x}(1+\delta)\) so that
\(l_{a-x}^\delta = e^{-\int_{0}^{a-x}{\mu_t(1+\delta})dt}=(l_{a-x})^{(1+\delta)}\),
it follows that:

\[CS_{a}^\delta = \int_{\alpha}^{a} {m_{x} l_{a-x}^{(1+\delta)}} dx.\]

Using the derivative
\(\frac{dl_{a-x}^{(1+\delta)}}{d\delta} = log(l_{a-x}) l_{a-x}^{(1+\delta)}\),
and in the third row reversing integrals between \(t\) and \(x\):

\[\frac{d CS_{a}^\delta}{d \delta} = \int_{0}^{a} {m_{x} l_{a-x} \log(l_{a-x}) dx}\\
= - \int_{0}^{a} {m_{x} l_{a-x} \int_{0}^{a-x}{\mu_t \, dt}\, dx}\\
= - \int_{0}^{a}{\mu_t} \int_{0}^{a-t} {m_{x} l_{a-t-x} \, dx}\, dt\]

The last integral is \(CS_{a-t}\). Dividing by \(CS_a\) and multiplying
by \(\delta\) get one useful expression. This is the negative of
cumulative hazard \(H_a\) but considering a positive factor
\(\frac{CS_{a-t}}{CS_a}\) that takes in account the relative amount of
surviving descendants that would be lost at each child age at risk.

\[\frac{\Delta CS_{a}}{CS_a} \approx - \left[\int_{0}^{a}{\mu_x  \, \frac{CS_{a-x}}{CS_a} \,dx \,}\right] \Delta \delta\]

Grasping this relationship intuitively is more difficult given the
interaction of birth and mortality rates (Keyfitz and Caswell (2005)).
This factor has a S-shape because \(l_x\) curvature and fertility
accumulation, that gives more weight to first ages, generally less than
1 except for mother ages close to age valuation \emph{a} with complete
reproductive life and high infant mortality (no additional children and
high risk for one age additional of survive), typically some countries
with \(a=50\) and first periods.\\

\hypertarget{related-measures}{%
\subsection{Related Measures}\label{related-measures}}

\hypertarget{burden-and-timing-of-maternal-bereavement}{%
\subsubsection{Burden and timing of maternal
bereavement}\label{burden-and-timing-of-maternal-bereavement}}

We call \emph{Mean Time Spent in Bereavement} (\emph{MTSB}) the absolute
measure of the expected total lifetime with a death daughter for a
mother aged \emph{a}, which can be expressed in terms of a temporary
expected lost years index in line with \(e^\dagger\) (Vaupel (1986)):

\[MTSB_a = \int_0^a{m_x\int_0^{a-x}{d_t e_{0|a-x-t}\, dt}\,dx} = \int_\alpha^a{m_x e_{0|a-x}^\dagger \,dx}\].

Where \(d_t\) is the death distribution from birth, \(e_{0|a-x-t}\) is
the life expectancy at birth until age \(a-x-t\) and
\(e_{0|a-x}^\dagger\) the temporary dispersion measure. But most
interesting would be to compare these years with the time that mothers
would expect to live with their daughters. We call this the
\emph{Intensity Time in Bereavement} (\(ITB\))): a ratio between
expected time with a ``lost'' and expected time with a ``life'', that
allows to make comparisons between population regimes.

\[ITL_a = \frac{\int_0^a{m_x e_{0|{a-x}}^\dagger dx}}{\int_0^a{m_x e_{0|a-x} dx}}\]

This is a ratio between child-years in two radically different states.
Looks similar to the transcendental entropy measure \(H\) (Keyfitz and
Caswell (2005)) but considering all the cohorts born during the the
mother´s life, weighted by their relative size \(m_x\). In the figure
@ref(fig:plot\_ITL\_MAL) is shown that \emph{ITB} is bigger for young
women because of the weight of infant mortality in their mother
experience, also with more dispersion for same levels of parity.

Another important factor for the child survival experience of mothers,
is the mean age at child loss, called here \(MAL\). This relation can be
derived by starting with the mother age \(x+t\) at each death child age
\(t\) at death, weighted by the fertility and survival function. In it,
\(\kappa\) is the mean age at childbirth for women aged \(a\),
\(MAD_{a-x}\) refers to the mean age at death for newborns that die
before \(a-x\), \(F_a\) is the accumulated fertility for a women aged
\(a\):

\[MAL_a = \frac{\int_0^a{m_x {\frac{\int_0^{a-x} l_t \, \mu_t (x+t) dt}{\int_0^{a-x} l_t \, \mu_t dt}}dx}}{\int_0^a{m_x}dx} \\
  MAL_a = \frac{\int_0^a{m_x \left[{x+\frac{\int_0^{a-x} l_t \, \mu_t \, t \, dt}{\int_0^{a-x} l_t \, \mu_t dt}}\right]dx}}{\int_0^a{m_x}} \\
  MAL_a = \frac{\int_0^a{m_x \,x \,dx}}{F_a} + \frac{\int_0^a{m_x MAD_{a-x}dx}}{F_a} \\
  MAL_a = \kappa + \frac{\int_0^a{m_x MAD_{a-x}dx}}{F_a}\]

\hypertarget{heterogeneity}{%
\subsubsection{Heterogeneity}\label{heterogeneity}}

In a model with heterogeneity, determined at birth with a multiplicative
effect, \(CS_a\) could be interpreted as a conditional expectation with
random variables \emph{K} for fertility heterogeneity and \emph{Z} for
mortality frailty.

\[CS_a(k,z) = \int_{0}^{\infty}\,{m(k)}_x\,{l(z)}_{a-x}\,dx\]

Following Coresh and Goldman (1988) we can express the fertility part in
a multiplicative way with variability in the level but not in the shape,
as \({m(k)}_x = F_\beta\,k\,r_x\), where \(F_\beta\) is the baseline
cumulated fertility until upper limit in fecundity age range \(\beta\),
\(K\) is a random variable with mean 1 that allows variability between
groups and \(r_x\) is the fertility structure by age
(\(\int_{0}^{\beta}{r_x}=1\)). Frailty part can be thought in a cohort
effect way, also in the multiplicative assumption as Vaupel and Missov
(2014), with \(l(z)_x=e^{-H_{x}Z}=l_x^Z\), with baseline hazard \(\mu\).
Considering the joint distribution \(f_{kz}\), the intuition could be
expressed as \(f_{zk}=f_{k|z}f_z\): groups with higher descendant´s
mortality would adjust their fertility level, with positive correlation.
Replacing both variables, the unconditional mean \(\overline{CS}_a\)
would be\footnote{Note that given K and Z values fertility and child
  survival are independent.}:

\[\overline{CS}_a = \int_{0}^{\infty}\int_{0}^{\infty}\left[\int_{0}^{a}{ {m(k)}_x l(z)_{a-x}\,dx}\right]f_{kz}\,dz \,dk\\
= F_\beta \int_{0}^{a-x} r_x \int_{0}^{\infty} \int_{0}^{\infty} {l_{a-x}}^z \, k\, f_{kz} \,dk\,dz\,dx\]

Isolating for an age \emph{x} we can inspect the part related to Z and
K. Assume \(Z = Y_0 + Y_1\) and \(K = e^{(Y_0 + Y_2)}\), given the fact
that the the historical relation is not linear (see figure
@ref(fig:heterogeneity)), like \(F_\beta=ln(q_{0,5})\). \emph{Y\_0},
\emph{Y\_1} and \emph{Y\_2} are independet and gamma distribuited. Then
we can express:

\[{l_{a-x}}^z \, k\, f_{kz} \,dk\,dz\\
e^{-H_x(Y_0 + Y_1)}e^{Y_0 + Y_2}\\
L_{Y_0}[1-H_x]L_{Y_1}[-H_x]L_{Y_2}\]

\hypertarget{Applications}{%
\subsection{Applications}\label{Applications}}

\hypertarget{Data}{%
\subsubsection{Data}\label{Data}}

We motivate this paper with an empirical example using fertility and
mortality rates from the 2019 Revision of the UN World Population
Prospects (UN WPP) for the Latin American Region (Nations (2019)). We
smoothed female \(l_x\) in quinquennial ages, using cubic-splines
constrained to monotonic decrease, taking \(L_0\) and \(T_{100}\) from
raw life tables as inputs for year-person calculations. For splitting
fertility five groups was used quadratic optimization approach by
Michalski and Gorlishchev (2018), with an desirable property for our
purpose which is a good fitting in parity. Also was assumed an unique
female percentage of newborns of 0.49 for all period-country cases.
Calculations was done in a discrete way assuming that the \emph{m\_x}
live borths are born at exact mother´s age \emph{x}. A negative relation
between mortality until 10 and TFR is shown in
@ref(fig:plot\_tfr\_q010). Specific paths like Argentina and Guatemala
shows very different demographic profiles during second part of XX
century.

\hypertarget{numerical-results}{%
\subsubsection{Numerical Results}\label{numerical-results}}

Consider a Latin American woman standing before us. If this were
1950-1955, she could reasonable expect to have 1.87 surviving children
on her 50\(^{th}\) birthday. In 2010-2015 a woman the same age would
only have 1.87 living children. The difference of 0 children is
explained by reduced fertility and improved mortality in the region
{[}give stats{]}. We now remove the effect of changing fertility by
considering the number of daughters surviving up to maternal age \(a\)
as a proportion of the daughters ever born to a woman that age,
\(\frac{CS_a}{F_a}\). Given Eq. \ref{eq:CS}, the increase from 0.82 in
1950-1955 to 0.82 in 2010-2015 must be explained by a change in
mortality.

An interesting corollary of these results is the question: at wich age,
a mother would have the same \emph{CS} than \emph{a=20}, like a mirror
effect, \(\int_{0}^{a}{m_xl_{a-x}}dx=\int_{0}^{y}{m_xl_{y-x}}dx\) with
\(a<\beta<y\). Could be solved numerically or with our approximation
taking \(F_al_{a-\kappa_a}=F_yl_{y-\kappa_y}\), solving for \emph{y}
which \(\frac{F_a}{F_y} = \frac{l_{a-\kappa_a}}{l_{y-\kappa_y}}\).

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/plot_CS-1.pdf}
\caption{Child Survival and Child survival as a share of cummulative
fertility by age, for women aged 30, 40 and 50. Estimates using UN WPP
data for Latin American countries in the period 1950-2015 period}
\end{figure}

Guatemala improved the approximation with years due to
rectangularization process in \(l_x\) (figure
@ref(fig:plot\_CS\_aprox)).

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/plot_CS_aprox-1.pdf}
\caption{Left: Error in approximation for a=40, from years 1950 (darker)
to 2015 (lighter) for all Latin American countries (blue) and Guatemala
(red). Right: Change in survival and fertility by age in Guatemala.}
\end{figure}

With \(\mu_{x+t}=\mu_x\) for \(t\) between 0 and 1,
\(\int_{x}^{x+1}{\mu_t dt} = M_x\), so discretizing
\(\frac{\Delta CS_{a}}{CS_a} \approx - \left[\sum_{0}^{a-1}{M_x \, \frac{CS_{a-1-x}}{CS_a}\,}\right] \Delta \delta\).

Assuming constant period rates the expected time in years that a mother
aged 30 passed with a death son was around 4\% in some countries at
middle XX Century. When increasing age, the survival experience depends
less on infant mortality, and the distribution is around less than 2\%
for women aged 50, converging to 0 on time. As extreme cases, in
1950-1955 Haiti women aged 30 would have experienced an intensity of
0.5\%, and in 2010-2015 0.5\%, while the women of Costa Rica 0.5\%, and
in 2010-2015 0.5\%. For those Latin American mothers aged 50 with 3
daughters born who suffered a lost, they experienced that at age 30 in
average (figure @ref(fig:plot\_ITL\_MAL)) (More can be analyzed here).

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/plot_ITL_MAL-1.pdf}
\caption{Intensity Tome Lost of womans aged 30, 40 and 50. Years 1950
(darker) to 2015 (lighter). Latinamerican countries in period 1950-2015}
\end{figure}

\hypertarget{references}{%
\subsection*{References}\label{references}}
\addcontentsline{toc}{subsection}{References}

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-alburez_gutierrez_womens_2019}{}%
Alburez-Gutierrez, Diego, Martin Kolk, and Emilio Zagheni. 2019.
``Women's Experience of Child Death over the Life Course: A Global
Demographic Perspective.'' Preprint. SocArXiv.
\url{https://doi.org/10.31235/osf.io/s69fz}.

\leavevmode\hypertarget{ref-bongaarts_projection_1987}{}%
Bongaarts, John. 1987. ``The Projection of Family Composition over the
Life Course with Family Status Life Tables.'' In \emph{Family
Demography: Methods and Their Application}, edited by John Bongaarts and
Thomas Burch. Oxford: IUSSP Series, Oxford Univer- sity Pres.

\leavevmode\hypertarget{ref-brass_derivation_1953}{}%
Brass, W. 1953. ``The Derivation of Fertility and Reproduction Rates
from Restricted Data on Reproductive Histories.'' \emph{Population
Studies} 7 (2): 137. \url{https://doi.org/10.2307/2172029}.

\leavevmode\hypertarget{ref-Coresh1988}{}%
Coresh, Josef, and Noreen Goldman. 1988. ``The effect of variability in
the fertility schedule on numbers of kin.'' \emph{Math. Popul. Stud.} 1
(2): 137--56. \url{https://doi.org/10.1080/08898488809525268}.

\leavevmode\hypertarget{ref-daatland_norwegian_2010}{}%
Daatland, Svein Olav, Marijke Veenstra, and Ivar A. Lima. 2010.
``Norwegian Sandwiches: On the Prevalence and Consequences of Family and
Work Role Squeezes over the Life Course.'' \emph{European Journal of
Ageing} 7 (4): 271--81. \url{https://doi.org/10.1007/s10433-010-0163-3}.

\leavevmode\hypertarget{ref-goodman1974}{}%
Goodman, Leo A, Nathan Keyfitz, and Thomas W Pullum. 1974. ``Family
Formation and the Frequency of Various Kinship Relationships.''
\emph{Theoretical Population Biology} 5 (1): 1--27.

\leavevmode\hypertarget{ref-Hendrickson2009}{}%
Hendrickson, K. C. 2009. ``Morbidity, Mortality, and Parental Grief: A
Review of the Literature on the Relationship Between the Death of a
Child and the Subsequent Health of Parents.'' \emph{Palliat Support
Care} 7 (1): 109--19.

\leavevmode\hypertarget{ref-KeyfitzMath}{}%
Keyfitz, Nathan, and Hal Caswell. 2005. \emph{Applied Mathematical
Demography}. Vol. 47. Springer.

\leavevmode\hypertarget{ref-lee_death_2014}{}%
Lee, Chioun, Dana A. Glei, Maxine Weinstein, and Noreen Goldman. 2014.
``Death of a Child and Parental Wellbeing in Old Age: Evidence from
Taiwan.'' \emph{Social Science \& Medicine} 101 (January): 166--73.
\url{https://doi.org/10.1016/j.socscimed.2013.08.007}.

\leavevmode\hypertarget{ref-livi_bacci_concise_1997}{}%
Livi Bacci, Massimo. 1997. \emph{A Concise History of World Population}.
2nd ed. Cambridge, MA: Blackwell.

\leavevmode\hypertarget{ref-lotka1931orphanhood}{}%
Lotka, Alfred J. 1931. ``Orphanhood in Relation to Demographic
Factors.'' \emph{Metron} 9 (2): 37--109.

\leavevmode\hypertarget{ref-Michalski2018}{}%
Michalski, Grigoriev, Anatoli I, and Vasily P and Gorlishchev. 2018. ``R
Programs for Splitting Abridged Fertility Data into a Fine Grid of Ages
Using the Quadratic Optimization Method.'' Max Planck Institute for
Demographic Research, Rostock, Germany.

\leavevmode\hypertarget{ref-WPP19}{}%
Nations, United. 2019. \emph{World Population Prospects 2019: Data
Booklet}.
\url{https://doi.org/https://doi.org/https://doi.org/10.18356/3e9d869f-en}.

\leavevmode\hypertarget{ref-preston1978effects}{}%
Preston, Samuel H, and others. 1978. \emph{The Effects of Infant and
Child Mortality on Fertility.} Academic Press, Inc., 111 Fifth Avenue,
New York/New York 10003, USA.

\leavevmode\hypertarget{ref-smith-greenaway_maternal_2020}{}%
Smith-Greenaway, Emily, and Jenny Trinitapoli. 2020. ``Maternal
Cumulative Prevalence Measures of Child Mortality Show Heavy Burden in
Sub-Saharan Africa.'' \emph{Proceedings of the National Academy of
Sciences}, February, 201907343.
\url{https://doi.org/10.1073/pnas.1907343117}.

\leavevmode\hypertarget{ref-Vaupel2014}{}%
Vaupel, James W., and Trifon I. Missov. 2014. ``Unobserved population
heterogeneity: A review of formal relationships.'' \emph{Demographic
Research} 31 (September): 659--86.
\url{https://www.demographic-research.org/Volumes/Vol31/22}.

\leavevmode\hypertarget{ref-Vaupel1986}{}%
Vaupel, J. W. 1986. ``How Change in Age-Specific Mortality Affects Life
Expectancy.'' \emph{Population Studies} 40 (1): 147--57.
\url{https://doi.org/10.2307/2174285}.

\leavevmode\hypertarget{ref-volk_infant_2013}{}%
Volk, Anthony A., and Jeremy A. Atkinson. 2013. ``Infant and Child Death
in the Human Environment of Evolutionary Adaptation.'' \emph{Evolution
and Human Behavior} 34 (3): 182--92.
\url{https://doi.org/10.1016/j.evolhumbehav.2012.11.007}.

\leavevmode\hypertarget{ref-wrycza2012}{}%
Wrycza, Tomasz, and Annette Baudisch. 2012. ``How Life Expectancy Varies
with Perturbations in Age-Specific Mortality.'' \emph{Demographic
Research} 27: 365--76.
\end{cslreferences}

\end{document}

### 2020-05-14 R/funs.R

@@ -1,0 +1,244 @@
+library(quadprog)
+library(wpp2019)
+library(tidyverse)
+library(gridExtra)
+source("R/QOSplit.R")
+
+# main_funtion
+
+# age survivor spline
+lx_fun <- function(data, age){
+  splinefun(data$x, data$lx, method = "monoH.FC")(age)
+}
+
+# function to calculate all the measures
+CS_results <- function(LA, country, period, a){   # LA is the data base
+  
+  # example
+  # country = "Brazil"; period = "1970-1975"; a = 70
+  
+  # filter country
+  country = LA %>% filter(name == country, Period==period)%>% 
+                    mutate(dx = c(-diff(lx),lx[101]),
+                           Lx = ifelse(x==0, Lx_n, # assume uniform deaths within interval
+                                       ifelse(x==100, ex, 
+                                              lx-dx/2)),
+                           Lx = as.numeric(Lx), # in case
+                           ex = rev(cumsum(rev(Lx)))/lx)
+  
+  # calculate: 
+    # assume that birth is in exact age (not in the middle). That allow us to work with lx, instead of Lx.
+    # consider that i=a is age a-1.
+  CS = sum(unlist(sapply(1:a, function(i) country$asfr[i]*country$lx[a+2-i]/100000)), na.rm = T)
+  Fa = sum(country$asfr[1:a], na.rm = T)
+  CS_prob = CS/Fa
+  k = sum(country$asfr[1:a]*0:(a-1), na.rm = T)/Fa # must be kappa but...
+  CS_aprox = Fa * country$lx[a+1-k]/100000
+  error_aprox = (CS-CS_aprox)/CS*100 # relative error approx
+  MTSL = 0
+  for(i in 1:a){
+    for(j in 1:(a+1-i)){
+      MTSL = sum(MTSL,
+                 country$asfr[i] * country$dx[j]/100000 * sum(country$Lx[j:a])/country$lx[j],
+                 na.rm=T)
+    }
+  }
+  MTSA = 0
+  for(i in 1:a){
+    for(j in 1:(a+1-i)){
+      MTSA = sum(MTSA,
+                 country$asfr[i] * sum(country$Lx[j:a])/country$lx[j],
+                 na.rm=T)
+    }
+  }
+  ITL = MTSL/MTSA * 100
+  m_MAD = 0
+  for(i in 1:a){
[…]

### 2020-05-14 R/tests.R

@@ -1,0 +1,275 @@
+# age survivor
+lx_fun <- function(data, age){
+  splinefun(data$x, data$lx, method = "monoH.FC")(age)
+}
+# p-y
+Lx_fun <- function(data, agei, agen){
+  integrate(lx_fun, agei, agen, data=data)
+  }
+
+# split asfr
+install.packages("quadprog")
+library(quadprog)
+source("split/QOSplit.R")
+int = QOSplit (Fx= c(0.00008, 0.02655, 0.10538, 0.09983, 0.05105, 0.01904, 0.00359,
+               0.00033, 0.00000), L=seq(10,50,5), AgeInt=rep (5,9))
+
+plot(L,Fx,t="s")
+lines(int,col=2)
+
+# data
+library(tidyverse)
+library(quadprog)
+source("split/QOSplit.R")
+library(wpp2019)
+
+# Get wpp data
+data(UNlocations)
+data(tfr)
+data(percentASFR)
+paises_LA <- UNlocations %>% filter(area_name == "Latin America and the Caribbean") %>% 
+              select(country_code, name) %>% distinct()
+paises_LA_seleccion <- c(192,214,332,388,630,188,222,320,340,484,558,591,32,68,76,152,170,218,600,604,858,862)
+lt_paises_LA <- readxl::read_xlsx("WPP2019_MORT_F17_3_ABRIDGED_LIFE_TABLE_FEMALE.xlsx", sheet = 1, 
+                range = "A17:T77381") %>% rename(country_code=5,
+                                      x = "Age (x)", 
+                                      mx = "Central death rate m(x,n)",
+                                      qx = "Probability of dying q(x,n)",
+                                      dx = "Number of deaths d(x,n)",
+                                      lx = "Number of survivors l(x)",
+                                      Lx_n = "Number of person-years lived L(x,n)",
+                                      ex = "Expectation of life e(x)")  %>% 
+  select(country_code,Period,x, mx, qx, dx, lx, Lx_n, ex) %>% 
+  filter(country_code %in% paises_LA_seleccion)
+tfr_paises_LA <- tfr %>% select(-last.observed) %>% gather(Period,tfr,-country_code,-name) %>% 
+  filter(country_code %in% paises_LA_seleccion)
[…]

### 2020-05-14 covid/covid_proposal.Rmd

@@
 ---
-title: 'Research proposal: Mortality exposure in times of Covid-19'
+title: 'Estimating excess bereavement due to Covid-19'
+subtitle: 'A research proposal'
 author: "Diego Alburez-Gutierrez and Ivan Williams"
-date: "12 May 2019"
+date: "15 May 2019"
 output:
@@
                       message=F, warning=F, echo = F)
-
-# library(tidyverse)
-# library(gridExtra)
-# library(quadprog)
-# library(wpp2019)
-# options(digits=2)
-# source("funs.R")
 ```
@@
 
-The subject of death has, once again, come to the forefront of public debate as the world grapples with a global pandemic of the Covid-19 disease. 
-Much attention has been given to estimating age-specific case fatality rates and mortality rates in general [ref].
-The degree to which the pandemic will increase the exposure to mortality in the population as a whole has recieved little or no attention in the demographic literature. 
-Studies in sociology and public health have consistently shown the negative and long-term consequences of bereavement on mental, physical, and emotional health, especially for women [ref].
-Yet, no study to date has considered the degree to which covid-19 excess mortality will increase expossure to death for people around the world.
+Death and bereavement has come to the forefront of public debate as the world grapples with the global pandemic of the Covid-19 disease, which has already brought about the death of thousands of mainly elderly people.
+Each death represents the loss of somebody's parent, grandparent, great-grandparent, aunt, uncle, cousin, or child.
+Studies in sociology and public health have consistently shown the negative and long-term consequences of bereavement on mental, physical, and emotional health, especially for women [@umberson_death_2017].
+Bereavement also matters because relatives are crucial providers of social and financial support and their absence will be direly felt by bereaved relatives [@Hendrickson2009].
+For many people living in the Global North this will be their first close encouter with death, as historically low mortality rates have implied a reduction in the exposure to mortality at all ages in high-income countries [@murphy_long-term_2011; @alburez_gutierrez_womens_2019].
+For people in the Global South, it will add to an already high burden of bereavement [@smith-greenaway_maternal_2020; @roberton_early_2020].
 
-Given the age-gradient of the Covid-19 case fatality rates [ref], we expect that the pandemic will result in the death of thousands of elderly relatives, such as parents, grandparents, great-grandparents, aunts and uncles.
-More worringly, it might result in the death of adult children for even older parents who depend on the crucial support given by their offspring or grandchildren.
-For many in the Global North this will be their first close encouter with death, as historically low mortality rates have implied a reduction in the exposure to mortality at all ages in high-income countries [@murphy_long-term_2011, @alburez_gutierrez_womens_2019].
+Much attention has been given to Covid-19 excess mortality rates, but little is known about how the pandemic will increase the expossure to death for people around the world.
+This project will combine existing and future data on Covid-19 excess mortality with powerful methods from formal demography [@goodman1974] to estimate the burden of Covid-19 bereavement.
+The project is not interested in the total number of people who will die from Covid-19, but in the number of people who will be affected by the death of a relative because of the disease.
+We are interested in `excess bereavement'.
+We ask:
+
+   1. How many people will lose a relative to Covid-19, either in the nuclear or in the extended family? 
+   2. What is the age distribution of the bereaved population?
+   3. How will the experience of bereavement affect the availability of kin resources for those that survive the disease?
 
@@
 
-This study will combine existing and future data on Covid-19 excess mortality [ref] with powerful methods from formal demography [@goodman1974] to estimate the potential burden of covid-19 bereavement.
-As such, we are not interested in the total number of people who will die from Covid-19, but in the number of people who will be affected by the death of a relative because of the disease.
-In other words, this is a study on `excess bereavement'.
-Given the range of covid-19 mortality rates in different countries, we ask: (1) How many people will lose a relative to Covid-19, either in the nuclear or in the extended family?
-(2) What is the age distribution of the bereaved population?
+### Methodological component
 
-The project comprises a methodological and an empirical component.
-In terms of methodology, we seek to generalize the Kin-Cohort Method, which is itself an extension of the Goodman, Keyfitz Pullum kinship equations [@goodman1974], to estimate the potential increase in the expossure to the death of relatives.
-this means deriving the mathematical expressions needed to model the effect of changes in mortality on the population-level prevalence of excess bereavement.
+We seek to generalize the Kin-Cohort Method [@alburez_gutierrez_womens_2019], itself an extension of the Goodman, Keyfitz Pullum kinship equations [@goodman1974], to estimate the potential increase in the expossure to the death of relatives.
+A first step will focus on deriving mathematical expressions to model the effect of changes in mortality on the population-level prevalence of bereavement.
 
 We exemplify this for the case of parental bereavement, which we understand as the probability of losing a child. 
-Let $CS_a$ be the expected number of surviving children to a mother alive at aged $a$ in a female stable population with fertility rates $m_{x}$, mortality hazard $\mu_{x}$ and survival function $l_x=e^{-\int_{0}^{x}{\mu_t}\,dt}$ (with unit radix $l_0=1$), as proposed by @goodman1974:
+Let $CS_a$ be the expected number of surviving children to a mother alive at age $a$ in a female stable population with fertility rates $m_{x}$, mortality hazard $\mu_{x}$ and survival function $l_x=e^{-\int_{0}^{x}{\mu_t}\,dt}$ (with unit radix $l_0=1$), as proposed by @goodman1974:
 
@@
 
+<!-- Si sacamos nuestro paper como pre-print, lo podemos citar aqui!! -->
+
 $$\frac{\Delta CS_{a}}{CS_{a}} \approx -(a - k_a) \Delta\delta.$$
 
-This is a first step towards developing a formal understanding the consequences of an increase in age-specific mortality on child survival but further development is needed, especially for the case in which mortality does not increase by the same factor at all ages.
-This is, of course, the case of Covid-19, which is known to affect people in older ages [ref]. 
-Further work is also needed to generalize these relationships to other types of kin, such as mothers, grandmothers, cousins and aunts.
-A final strand of methodological component of the project will seek to develop the introduce heterogeneity in the Kin-Cohort models. 
-In their current state, these models ignore the role of heterogeneity in the experience of kin death, a particular limitation when considering that deaths from the infectious disease are likely to be clustered in the population.
+Deriving these formal relationships for other types of relative (mothers, grandmothers, siblings, cousins, and aunts) will provide key insight for the empirical analysis of excess bereavement. 
+Further development is needed to consider age-specific changes in mortality, as the Covid-19 disease affects mainly people in older ages.
+Finally, the existing models ignore heterogeneity in the experience of kin death and within-group correlations.
+This is a limitation considering that deaths from infectious disease are likely to be clustered in the population.
+The final strand of the methodological component of the project will adress these issues.
 
-The empirical componente of the project will apply the mathematical models to month-specific mortality and fertility rates (interpolated, if needed, from the available data taking the seasonsality of mortality and fertility into account) to estimate the excess burden of kin death derived from Covid-19.
-Combining this with data on the age and sex structure of the population will allow us to estimate excess bereavement, which we will compare to a baseline counter-factual scenario using rates that exclude the excess mortality from Covid-19.
-The study will produce a range of estimates, given the uncertainty inherent to the Covid-19 mortality statistics that are required as input for our mathematical models.
+### Empirical component
 
-As an initial step, the analysis will focus on Germany and Italy, for which quality data on covid-19 death rates [ref] is already available at the sub-national level and where more quality data is expected in the future [ref].
-Our methodology is country-agnostic, meaning that it can be applied to other settings once quality data on Covid-19 excess mortality is becomes available.
+We will estimate the excess burden of kin death from the Covid-19 disease using our novel mathematical models with mortality and fertility rates as inputs.
+Combining this with data on the age and sex structure of the population will allow us to estimate the magnitude of the excess bereavement and the age distribution of the bereaved population\footnote{`Excess' bereavement can be compared to a baseline counter-factual scenario using rates that exclude the excess mortality from Covid-19.}.
+
+As an initial step, the analysis will focus on Germany and Italy, for which quality data on Covid-19 death rates is already available.
+Further development may apply our innovative methodology to other setting once data on Covid-19 excess mortality becomes available. 
+The project will produce a range of estimates, given the uncertainty inherent to the Covid-19 mortality statistics that are required as input. 
+However, we expect the accuracy of our models to improve as more data quality becomes available.
 
@@
 
-- A set of academic publications describing the methodology and the main results of the study
 - A flexible and scalable methodology for estimating the excess burden of covid-19 bereavement.
-The code to produce the estimates will be made freely available on the code-sharing platform [Github](Github.com).
+The open-source code to produce the estimates will be made freely available on the code-sharing platform [Github](www.github.com).
 - The first set of estimates of how the Covid-19 disease will affect the expossure to mortality at a population-level
+- A set of academic publications describing the methodology and the main results of the project
 - A online platform for scientists and policy makers to access and download the data easily (see [this interactive app](https://research-app.shinyapps.io/child_death_paa/))
 
-## Work plan
+<!-- DIAGRAMA DE PARENTEZCO, DONDE EL EFECTO DE MORTALITYDAD ES LA FLECHA Y EL NODO EL NUMERO ESPERADO DE FAMILIARES -->
 

### 2020-05-14 paper.Rmd
clean docs

@@
 title: 'Child survival for mothers: mortality change and related measures'
-author: "Ivan Williams and Diego Alburez-Gutierrez"
+author: "Iván Williams and Diego Alburez-Gutierrez"
 date: "May 2020"
@@
 Comentarios:
+- hay dos comentarios más abajo: uno sobre algo que agregaría a la introducción y otra cosa sobre una idea interesante que podríamos agregar como mini corolario (line 217 aprox).
 - Puede sonar que la instroducción es un poco larga para el resto del cuerpo. Lo dejo a tu criterio porque me cuesta sacar algo de allí, aprece todo bastante relevante.
-- hay dos comentarios más abajo: uno sobre algo que agregaría a la introducción y otra cosa sobre una idea interesante que podríamos agregar como mini corolario (line 217 aprox).
 - Perdón por mi spanglish.
@@
 *What can we look for here?*:
+
 * _a formalization of covariance_
@@
 * _estimate joint distribution using historical data. Idea: ¿maybe taking all the period/country observations as realized values of an hypothetical stable population, and using bayesian procedures?_
+
+...
 
@@
                   geom_smooth(se = F, color = "grey", method = "lm",
-                              formula = y ~ log(x)) +
+                              formula = y ~ log(x), linetype = "dashed") +
                   geom_line(data = LA %>% filter(x==0, name=="Latin America and the Caribbean"),
@@
              scale_y_continuous("Approximated change in CS") +
-             scale_color_continuous("Absolute change\nin delta") +
+             scale_color_continuous("Delta") +
              theme_classic()
@@
              scale_y_continuous("Approximated change in CS") +
-             scale_color_continuous("Relative change\nin delta") +
+             scale_color_continuous("Delta") +
              theme_classic()

### 2020-05-15 covid/alburez_covid_proposal.Rmd

---
title: 'Estimating excess bereavement due to Covid-19'
subtitle: 'Proof of concept for Volkswagen "Corona Crisis and Beyond – Perspectives for Science, Scholarship and Society" grant'
author: "Diego Alburez-Gutierrez, Ivan Williams, and Emilio Zagheni (?)"
date: "15 May 2019"
output:
  pdf_document:
    fig_caption: yes
header-includes: \usepackage{todonotes}
bibliography: bib.bibtex
urlcolor: blue
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      message=F, warning=F, echo = F)
```

## In a nutshell

Much attention has been given to Covid-19 excess mortality rates, but little is known about how the pandemic will increase the exposure to the death of relatives for people around the world.
This project seeks to estimate the global burden of bereavement associated with the Covid-19 disease.
It is the first attempt to quantify this phenomenon and its wide-ranging implications for society using a set of innovate methods from mathematical demography and demographic micro-simulation.

**Keywords** demography, bereavement, family support, excess mortality

## Project description

This project will combine existing and future data on Covid-19 excess mortality with powerful methods from formal demography to estimate the global burden of Covid-19 bereavement.
The project is not interested in the total number of people who will die with Covid-19, but in the number of people who will be affected by the death of a relative because of the disease, a phenomenon we call `excess bereavement'.

### Objectives

1. Operationalize a set of equations from mathematical demography to estimate levels of Covid-19 bereavement
1. Quantify the number of people expected to lose a relative to Covid-19 in the nuclear and in the extended family in countries around the world
1. Determine which demographic groups will be at a higher risk of losing a relative to Covid-19 and how this will vary by type of relative
1. Quantify the degree to which Covid-19 will affect the availability of family resources for grieving relatives over the life-course

   <!-- 1. How many people will lose a relative to Covid-19, either in the nuclear or in the extended family?  -->
   <!-- 1. Which demographic groups will be at a higher risk of losing a relative to Covid-19 and how will this vary by type of relative lost? -->
   <!-- 1. How will excess bereavement affect the availability of family resources for those that survive the disease? -->

### Expected main results

As a first step, we seek to generalize the Kin-Cohort Method [@alburez_gutierrez_womens_2019], itself an extension of the Goodman, Keyfitz Pullum Kinship equations [@goodman1974], to estimate the potential increase in the exposure to the death of relatives given a set of age-specific Covid-19 mortality rates. 
We will focus on deriving mathematical expressions to model the effect of changes in mortality on the population-level prevalence of excess bereavement.

As a second step, we will estimate the excess burden of kin death from the Covid-19 disease using our novel mathematical models with demographic micro-simulations [@zagheni_impact_2011].
Combining this with data on the age and sex structure of the population will allow us to estimate the magnitude of the excess bereavement and the age distribution of the bereaved population\footnote{`Excess' bereavement can be compared to a baseline counter-factual scenario using rates that exclude the excess mortality from Covid-19.}.

Our models will be calibrated using data from Germany and Italy, for which quality data on Covid-19 death rates is already available.
We will then extend our innovative methodology to other parts of the world once reliable data on Covid-19 excess mortality becomes available. 
Demographic micro-simulation will be used to obtain estimates for countries without high-quality data.
The project will produce a range of estimates to reflect the uncertainty inherent to the Covid-19 mortality statistics that are required as input. 
However, we expect the accuracy of our models to improve as more data quality becomes available.

The expected results are:

1. A flexible and scalable methodology for estimating the excess burden of Covid-19 bereavement. The open-source code to produce the estimates will be freely available on the code-sharing platform [Github](www.github.com).
1. Country-level estimates of how the Covid-19 disease will affect the exposure to mortality at a population-level
1. A set of academic publications in high-impact journals describing the methodology and the main results of the project
1. A online platform for scientists and policy makers to access and download the data easily (see [this interactive app](https://research-app.shinyapps.io/child_death_paa/) from one of the authors)

### Relevance of the topic for society

Death and bereavement have come to the forefront of public debate as the world grapples with the global pandemic of the Covid-19 disease.
The pandemic has already brought about the death of thousands of mainly elderly people. 
Each death is meaningful in itself, but it also represents the loss of somebody's parent, grandparent, great-grandparent, aunt, uncle, cousin, or child.
Studies in sociology and public health have consistently shown the negative and long-term consequences of bereavement on mental, physical, and emotional health, especially for women [@umberson_death_2017].
Bereavement also matters because relatives are crucial providers of social and financial support, the loss of which affects the individuals left behind [@Hendrickson2009].
However, there are currently no estimates of the number of people who will lose a relative to the Covid-19 disease.

Information on the expected number of bereaved relatives can help policy makers develop appropriate plans for supporting the grieving relatives. 
Data on the age of the grieving relatives is key for targeting this programs more effectively - a different approach may be needed for dealing with young orphans that for supporting elderly parents who lose a middle-aged child.
The effects of bereavement are also bound to vary by levels of socio-economic development
For many people living in the Global North this will be their first close encounter with death, as historically low mortality rates have implied a reduction in the exposure to mortality at all ages in high-income countries [@murphy_long-term_2011; @alburez_gutierrez_womens_2019].
For people in the Global South, it will add to an already high burden of bereavement [@smith-greenaway_maternal_2020; @roberton_early_2020].

### Relevance of the topic for science

Our project will operationalize, for the first time, a set of demographic equations that can be used to estimate the expected number of surviving kin and the number of kin expected to die in the context of dramatic changes in demographic rates.
We show how complex estimations, previously thought unfeasible [@KeyfitzMath], are now possible given the rapid advances in computational power and parallel processing.
Our interdisciplinary approach, combining demographic theory and data science, has the potential to make methodological and substantial contirbutions to our understanding of branching processes which have a wide application in other fields.

Substantially, kin count estimation has a long pedigree in mathematical demography, starting with the work of @lotka1931orphanhood on modelling orphanhood in theoretical populations across demographic regimes.
The question of kin survival sits at the very center of demographic theory used for studying human and non-human populations. 
Historical demographers draw liberally on assumptions about kin availability and individual's exposure to the death of relatives to explain human behavior, especially in the context of rapid societal change such as epidemics, but these assumptions are often untested given the data sparsity [@livi_bacci_concise_1997;@volk_infant_2013]. 
This project will produce reliable quantitative estimates about the exposure to mortality in the context of a current pandemic.
The same approach can well be applied to understand the prevalence of bereavement in past and future global mortality crises.

## Details on collaboration

## References

### 2020-05-18 covid proposal/archive/20200518/alburez_covid_proposal.Rmd
first word draft

--- 2020-05-15 covid/alburez_covid_proposal.Rmd
@@
 ---
-title: 'Estimating excess bereavement due to Covid-19'
+title: 'Estimating excess bereavement due to Covid-19 around the world'
 subtitle: 'Proof of concept for Volkswagen "Corona Crisis and Beyond – Perspectives for Science, Scholarship and Society" grant'

### 2020-05-25 covid/archive/20200514.Rmd
changed name

---
title: 'Estimating excess bereavement due to Covid-19'
subtitle: 'Grant proposal draft'
author: "Diego Alburez-Gutierrez and Ivan Williams"
date: "15 May 2019"
output:
  pdf_document:
    fig_caption: yes
header-includes: \usepackage{todonotes}
bibliography: bib.bibtex
urlcolor: blue
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      message=F, warning=F, echo = F)
```

**Outline of grant proposal for Volkswagen's "Corona Crisis and Beyond – Perspectives for Science, Scholarship and Society" grants**

- [Website](https://www.volkswagenstiftung.de/en/funding/our-funding-portfolio-at-a-glance/corona-crisis-and-beyond-perspectives-for-science-scholarship-and-society)
- Amount: up to 120,000 Euro (flexible use)
- Deadline: June 4 2020
- Proposal: 4-page project description
- Team (up to three members): Diego Alburez, Ivan Williams, Emilio Zagheni (?)

## Introduction

Death and bereavement have come to the forefront of public debate as the world grapples with the global pandemic of the Covid-19 disease, which has already brought about the death of thousands of mainly elderly people. 
Each death is meaningful in itself, but it also represents the loss of somebody's parent, grandparent, great-grandparent, aunt, uncle, cousin, or child.
Studies in sociology and public health have consistently shown the negative and long-term consequences of bereavement on mental, physical, and emotional health, especially for women [@umberson_death_2017].
Bereavement also matters because relatives are crucial providers of social and financial support, the loss of which affects the individuals left behind [@Hendrickson2009].

Much attention has been given to Covid-19 excess mortality rates, but little is known about how the pandemic will increase the exposure to death for people around the world.
For many people living in the Global North this will be their first close encounter with death, as historically low mortality rates have implied a reduction in the exposure to mortality at all ages in high-income countries [@murphy_long-term_2011; @alburez_gutierrez_womens_2019].
For people in the Global South, it will add to an already high burden of bereavement [@smith-greenaway_maternal_2020; @roberton_early_2020].

This project will combine existing and future data on Covid-19 excess mortality with powerful methods from formal demography [@goodman1974] to estimate the burden of Covid-19 bereavement.
The project is not interested in the total number of people who will die with Covid-19, but in the number of people who will be affected by the death of a relative because of the disease.
We call this phenomenon `excess bereavement'.
We ask:

   1. How many people will lose a relative to Covid-19, either in the nuclear or in the extended family? 
   1. Which demographic groups will be at a higher risk of losing a relative to Covid-19 and how will this vary by type of relative lost?
   1. How will excess bereavement affect the availability of family resources for those that survive the disease?

<!-- 2. What is the expected age distribution of the bereaved population? -->

## Project description

### Methodological component

We seek to generalize the Kin-Cohort Method [@alburez_gutierrez_womens_2019], itself an extension of the Goodman, Keyfitz Pullum kinship equations [@goodman1974], to estimate the potential increase in the exposure to the death of relatives given a set of age-specific Covid-19 mortality rates. 
A first step will focus on deriving mathematical expressions to model the effect of changes in mortality on the population-level prevalence of excess bereavement.

\todo[inline]{This comes from our working paper with Ivan}
We exemplify this for the case of parental bereavement - the loss of a child from the perspective of a parent. 
Let $CS_a$ be the expected number of surviving children to a mother alive at age $a$ in a female stable population with fertility rates $m_{x}$, mortality hazard $\mu_{x}$ and survival function $l_x=e^{-\int_{0}^{x}{\mu_t}\,dt}$ (with unit radix $l_0=1$), as proposed by @goodman1974:

$$CS_a = \int_{0}^a{m_{x}l_{a-x}dx}.$$

Now consider an absolute increase in mortality $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk.
We can think of this hypothetical increase in mortality as the all-age excess mortality derived from the Covid-19 disease so that $m_{x,\delta}=m_{x}+\delta$ (@wrycza2012) and $l_{a-x}^\delta = e^{-\int_{0}^{a-x}{(\mu_t+\delta})dt}$:

$$CS_{a}^\delta = \int_{\alpha}^{a}m_{x} l_{a-x} e^{-\delta (a-x)} dx.$$

As we have shown elsewhere [ref to our working paper], we can conceptualize the effects of this absolute change $\delta$ in mortality as the difference between maternal age $a$ and the mean age of the mother at the birth of her surviving daughters $k_a$:

<!-- Si sacamos nuestro paper como pre-print, lo podemos citar aqui!! -->

$$\frac{\Delta CS_{a}}{CS_{a}} \approx -(a - k_a) \Delta\delta.$$

We seek to derive these formal relationships for other types of relative (mothers, grandmothers, siblings, cousins, and aunts) to enable the empirical analysis of Covid-19 excess bereavement. 
Further development is needed to consider age-specific changes in mortality, as the Covid-19 disease affects mainly people in older ages and the model presented above assumes a uniform mortality change at all ages.
Finally, the existing models ignore heterogeneity in the experience of kin death and within-group correlations.
This is a limitation considering that deaths from infectious disease are likely to be clustered in the population.
The final strand of the methodological component of the project will dress these issues.

### Empirical component

We will estimate the excess burden of kin death from the Covid-19 disease using our novel mathematical models with mortality and fertility rates as inputs.
Combining this with data on the age and sex structure of the population will allow us to estimate the magnitude of the excess bereavement and the age distribution of the bereaved population\footnote{`Excess' bereavement can be compared to a baseline counter-factual scenario using rates that exclude the excess mortality from Covid-19.}.

As an initial step, the analysis will focus on Germany and Italy, for which quality data on Covid-19 death rates is already available.
Further development will extend our innovative methodology to other setting once reliable data on Covid-19 excess mortality becomes available. 
The project will produce a range of estimates to reflect the uncertainty inherent to the Covid-19 mortality statistics that are required as input. 
However, we expect the accuracy of our models to improve as more data quality becomes available.

## Expected main results

- A flexible and scalable methodology for estimating the excess burden of Covid-19 bereavement. The open-source code to produce the estimates will be freely available on the code-sharing platform [Github](www.github.com).
- The first set of estimates of how the Covid-19 disease will affect the exposure to mortality at a population-level
- A set of academic publications describing the methodology and the main results of the project
- A online platform for scientists and policy makers to access and download the data easily (see [this interactive app](https://research-app.shinyapps.io/child_death_paa/) from one of the authors)

<!-- DIAGRAMA DE PARENTEZCO, DONDE EL EFECTO DE MORTALITYDAD ES LA FLECHA Y EL NODO EL NUMERO ESPERADO DE FAMILIARES -->

## References

### 2020-06-02 covid/latest/attachments/covid-19 VW grant/CV_Diego_Alburez_VW.tex
added cvs

% LaTeX file for resume 
% This file uses the resume document class (res.cls)

\documentclass{res} 
%\usepackage{helvetica} % uses helvetica postscript font (download helvetica.sty)
%\usepackage{newcent}   % uses new century schoolbook postscript font 
\newsectionwidth{0pt}  % So the text is not indented under section headings
\usepackage{fancyhdr}  % use this package to get a 2 line header
\usepackage{booktabs}
\usepackage{hyperref}
%\usepackage{tabularx}
\usepackage{lipsum}

\hypersetup{
    colorlinks=true,
    urlcolor=blue,
}

\renewcommand{\headrulewidth}{0pt} % suppress line drawn by default by fancyhdr
\setlength{\headheight}{24pt} % allow room for 2-line header
\setlength{\headsep}{24pt}  % space between header and text
\setlength{\headheight}{24pt} % allow room for 2-line header
\pagestyle{fancy}     % set pagestyle for document
\rhead{ {\it D. Alburez}\\{\it p. \thepage} } % put text in header (right side)
\cfoot{}                                     % the foot is empty
\topmargin=-0.5in % start text higher on the page

\begin{document}
\thispagestyle{empty} % this page has no header

\today

\name{Diego Alburez-Gutierrez\\[12pt]}% the \\[12pt] adds a blank line after name

\address{Konrad-Zuse-Stra{\ss}e 1, 18057 Rostock, Germany \\  \href{mailto:alburezgutierrez@demogr.mpg.de}{alburezgutierrez@demogr.mpg.de}; \href{www.alburez.me}{www.alburez.me} }

% \href{https://twitter.com/d_alburez}{@d\textunderscore alburez}

\begin{resume}

I am a social scientist specialising in inter-generational demography. 
My work has used mathematical modeling, micro-simulation techniques, and empirical analysis to explore changes in the experience of life events such as kin availability, family bereavement, and caring responsibilities across demographic regimes.

%\vspace{0.2in}
\section{\centerline{EDUCATION}}

\textbf{The London School of Economics and Political Science, UK} \\
{\sl PhD}, Demography and Population Studies \hfill 2014 - 2018

{\sl MSc},  Social Research Methods (Distinction) \hfill 
 2013-2014

\textbf{Del Valle University, Guatemala} \\
{\sl BA},  Anthropology (First class) \hfill 
2008-2012

%\vspace{0.2in}

\section{\centerline{AWARDS AND SCHOLARSHIPS}}

   \begin{itemize}
    \item 2019 \href{https://www.mpg.de/prizes/otto-hahn-medal}{Otto Hahn Medal} (2020) awarded by the Max Planck Society
   	\item LSE PhD Academy Prize (2018) awarded by the London School of Economics for research on state-sponsored massacres in Guatemala
 	\item LSE PhD Studentship (2014-2018) awarded by the London School of Economics
%    \item \href{https://www.mpg.de/prizes/otto-hahn-medal}{Otto Hahn Medal} (2019; EUR 7,000 $\approx$ USD 7,500) awarded by the Max Planck Society for `outstanding achievement from the initial creative period of young scientists', in particular `for methodological innovations for studying the long-term consequences of mass killings on local populations' -
%   	\item LSE PhD Academy Prize (2018) for research on state-sponsored massacres in Guatemala
% 	\item LSE PhD Studentship (2014-2018; GBP 72,000 $\approx$ USD 121,000) 
 \end{itemize}

\section{\centerline{RESEARCH EXPERIENCE}}

\textbf{Max Planck Institute for Demographic Research (MPIDR), Germany} \\
{\sl Research Scientist (post-doc) and Research Area Chair (Aging and Generational Processes)}, Lab of Digital and Computational Demography \hfill        Since 2018

\textbf{United Nations High Commissioner for Refugees (UNHCR), Denmark} \\
{\sl Statistics Team Intern}, Field Information and Coordination Support Section (FICSS) \hfill        2018

\textbf{The London School of Economics and Political Science, UK} \\
{\sl Research Assistant}, positions in the Sociology, Social Policy, Methodology, and Law departments  \hfill       2013-2017

\textbf{The Population Council}, Guatemala \\
{\sl Research Sub Coordinator} \hfill    2012

\section{\centerline{PUBLICATIONS IN PEER-REVIEWED JOURNALS}}

\begin{flushleft}
\hangindent=2em
Nepomuceno, M., Acosta, E., \textbf{Alburez-Gutierrez, D.}, Aburto, J.M., Gagnon, A., and Turra, C. (Forthcoming). ``Besides population age-structure, health and other demographic factors can contribute to understanding the COVID-19 burden across the globe." Proceedings of the National Academy of Sciences (Accepted for publication). \\
\textbf{Alburez-Gutierrez, D.}, Chandrasekharan, E., Chunara, R., Gil-Clavel, S., Hannak, A., Interdonato, R., Joseph, K., Kalimeri, K., Malik, M., Mayer, K., Mejova, Y., Paolotti, D., and Zagheni, E. (2019). ``Reports of the Workshops Held at the 2019 International AAAI Conference on Web and Social Media." AI Magazine, 40(4), 78-82. doi:\href{https://doi.org/10.1609/aimag.v40i4.5287}{10.1609/aimag.v40i4.5287}. \\
Leone, T., \textbf{Alburez-Gutierrez, D.}, Gandour, R., Coast, E., and Giacaman, R. (2019). ``Maternal and child access to care and intensity of conflict in the occupied Palestinian territory: a pseudo-longitudinal analysis (2000–2014)." Conflict and Health 13(1). doi:\href{https://doi.org/10.1186/s13031-019-0220-2}{10.1186/s13031-019-0220-2}. \\
\textbf{Alburez-Gutierrez, D}. (2019). Blood is thicker than bloodshed: A genealogical approach to reconstruct populations after armed conflicts. Demographic Research, 40(23): 627-656. doi:\href{https://www.demographic-research.org/volumes/vol40/23/}{10.4054/DemRes.2019.40.23}. \\
McGovern, P., and \textbf{D. Alburez-Gutierrez}. (2017) ``Who takes workplace case-study methods seriously? The influence of gender, academic rank and PhD training." Industrial Relations Journal 48(2):98-114, doi:\href{http://onlinelibrary.wiley.com/doi/10.1111/irj.12171/full}{10.1111/irj.12171}.
\end{flushleft}

\section{\centerline{BOOK CHAPTERS AND TECHNICAL REPORTS}}

\begin{flushleft}

\hangindent=2em
\textbf{Alburez-Gutierrez, D.}, S. Aref, S. Gil-Clavel, A. Grow, D.V. Negraia, E. Zagheni. (2019). ``Demography in the Digital Era: New Data Sources for Population Research.'' Proceedings of the 2019 Conference of the Italian Statistical Society. doi: \href{https://doi.org/10.31235/osf.io/24jp7}{10.31235/osf.io/24jp7}. \\
\hangindent=2em
\textbf{Alburez-Gutierrez, D.} and Segura Garcia, C. (2018). ``The UNHCR Demographic Projection Tool: estimating the future size and composition of forcibly displaced populations." UNHCR Statistics Technical Series, \href{http://www.unhcr.org/en-au/5ae9ee747.pdf}{2018/1}.

\end{flushleft}

\section{\centerline{INVITED PRESENTATIONS (THREE SELECTED)}}

\begin{flushleft}

\hangindent=2em
\textbf{Alburez-Gutierrez, D.} (27 May 2020, postponed due to Covid-19 lockdown) ``Die demografischen Auswirkungen von Massengewalt in Guatemala [The demographic consequences of Genocide in Guatemala]." Berlin, Germany (Ibero-American Institute). \\
\hangindent=2em
\textbf{Alburez-Gutierrez, D.} (27 Nov 2019) ``Women's experience of child death: a global demographic perspective." \href{https://www.iser.essex.ac.uk/events/seminars/external/2019-11-27}{ISER External Seminar}, Colchester, UK (University of Essex). \\
\hangindent=2em
\textbf{Alburez-Gutierrez, D.} (23 May 2019) ``The Demographic Foundations of the Lived Experience of Kin Death: A Genealogical Reconstruction." \href{https://www.suda.su.se/about-us/events/suda-demographic-colloquium-series/suda-demographic-colloquium-diego-alburez-gutierrez-max-planck-institute-for-demographic-research-1.423293}{SUDA Demographic Colloquium Series}, Stockholm University.

\end{flushleft}

\section{\centerline{OUTREACH AND SCIENCE COMMUNICATION}}

\begin{itemize}
  \item Academic video: `How Can We Quantify the Immediate and Lingering Impact of Genocide on Local Populations?' (\href{https://lt.org/publication/how-can-we-quantify-immediate-and-lingering-impact-genocide-local-populations}{Latest Thinking, Apr 2020})
  \item Opinion piece: `Coronavirus: a warning to Latin America and the Caribbean to dramatically increase COVID-19 testing' (\href{https://theconversation.com/coronavirus-a-warning-to-latin-america-and-the-caribbean-to-dramatically-increase-covid-19-testing-135759}{The Conversation, UK, Apr 2020})
  \item Blog post: `Genocide is the tip of the iceberg: Reviewing the Guatemalan case' (\href{https://blogs.lse.ac.uk/humanrights/2016/02/18/genocide-is-the-tip-of-the-iceberg-reviewing-the-guatemalan-case/?fbclid=IwAR3mPVASOpmBfAu19BMXk99_TGgo6-0-axr6ZT0i5msU7atkppm4rgant_Q}{LSE Human Rights Blog, Feb 2016})
\end{itemize}

\section{\centerline{ORGANISER OF INTERNATIONAL ACADEMIC EVENTS}}

\begin{itemize}
    \item \textit{Organiser}: \href{https://www.demogr.mpg.de/en/news_events_6123/news_press_releases_4630/news/open_access_to_demographic_research_6699}{Rostock Open Science Workshop}, Max Planck Institute for Demographic Research (MPIDR), Rostock\hfill 2019
    \item \textit{Organiser}: \href{https://iussp.org/en/workshop-demographic-research-web-and-social-media-data}{Workshop on Demographic Research with Web and Social Media Data}, $13^{th}$ International AAAI Conference on Web and Social Media (ICWSM), Munich \hfill 2019
\end{itemize}

\section{\centerline{SERVICE TO THE DISCIPLINE}}

\begin{itemize}
    \item \textit{Reviewer} for academic journals: Demography; Journal of Marriage and Family; Demographic Research; Revue Quetelet/Quetelet Journal; $11^{th}$ International Conference on Social Informatics (SocInfo `19)
	\item \textit{Organiser} of the `Social Policy Qualitative Research Group', London School of Economics  \hfill 2017-2018 
	\item \textit{President} of the `Methodology Student Union', London School of Economics \hfill 2013-2014
%	\item \textit{Organiser} of the II Colloquium of Anthropology and Sociology, Universidad Del Valle\hfill 2011 
\end{itemize}

\end{resume} 
\end{document}

### 2020-06-02 paper.Rmd
added cvs

@@
 Recent development have underscored the need to understand how sudden changes in mortality affect the availability of kin. 
-In the context of the global pandemic caused by the Covid-19 disease, elderly people depend on support provided by their relatives, especially during periods of lockdown and in the absence of governmental support mechanisms.<!-- Maybe here are repeated --> Elderly parents may be at a higher risk of losing the key support provided by their adult children given the known age-gradient in the Covid-19 case fatality rate.
+In the context of the global pandemic caused by the Covid-19 disease, elderly people depend on support provided by their relatives, especially during periods of lockdown and in the absence of governmental support mechanisms.
+<!-- Maybe here are repeated --> Elderly parents may be at a higher risk of losing the key support provided by their adult children given the known age-gradient in the Covid-19 case fatality rate. 
 How do changes in mortality affect the availability of offspring over age from the point of view of a prospective mother? 
@@
 How this would work in an heterogeneity population?
-In the remainder of the paper, we aim to formalize the relationship population-level changes in mortality rates and perceived change in the lived experience of child death from the perspective of a mother.
+
+In this paper we aim to formalize the relationship population-level changes in mortality rates and perceived change in the lived experience of child death from the perspective of a mother.
 Our approach is key for understanding the impact of age-specific excess mortality on the resilience or otherwise of kinship networks for the vulnerable elderly population.

### 2020-08-11 ALAP2020_old/ALAP_abs.Rmd
Aclaraciones,  mejora redaccion

---
title: "Sobrevivencia de descendientes: aspectos formales y medidas para Latinoamérica"
author:  |
  Iván Williams [^iw] y Diego Alburez-Gutierrez [^da]

    [^iw]: Universidad Nacional de Luján - Max Planck Institute for Demographic Research
    [^da]: Max Planck Institute for Demographic Research
institute: "IX Congreso de la Asociación Latinoamericana de Población - Valparaíso"
date: "Marzo 2020"
output: pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Resumen

### Antecedentes y Objetivo

El número de hijos sobrevivientes que una madre espera tener a distintas edades es una relación importante en la teoría demográfica del parentezco. Si bien esto afecta a las madres de varias maneras, no está claro cómo los cambios en los patrones de mortalidad, dado un esquema de fecundidad dado, afectan el número de niños sobrevivientes desde el punto de vista de las madres. Los objetivos de este trabajo son dos:

- Encontrar expresiones formales para los cambios absolutos y relativos en la mortalidad, y su impacto en la sobrevivencia de hijos.

- Analizar la evolución de este fenómeno en Latinoamérica durante la segunda parte del SXX.

### Materiales y Métodos

Se utilizarán herramientas demográficas formales para caracterizar la sobrevivencia de descendientes desde el punto de vista de una madre, considerando cambios tanto absolutos como relativos en las tasas de mortalidad. Se estimarán numpericamente el impacto compararando poblaciones latinoamericanas en distintos estadíos demográficos a partir de tablas de mortalidad disponibles en LAMBdA^[https://www.ssc.wisc.edu/cdha/latinmortality/].

### Primeros resultados

El cambio esperado en la sobrevivencia de hijos dado un cambio absoluto en la mortalidad depende de la diferencia entre la edad de una madre y la estructura por edad de sus hijos sobrevivientes. El efecto de un cambio relativo será mayor cuando la mortalidad sea dispersa en edades donde la fecundiad sea concentrada. Tambien se ofrecen expresiones para computar el tiempo medio y relativo que transitan las mujeres con hijos difuntos (en función de $e^\dagger$) y la edad media a la pérdida (en función de la edad media a la fecundidad y la paridad).

### 2020-08-11 paper.Rmd
Aclaraciones,  mejora redaccion

--- 2020-05-14 paper.Rmd
@@
 @bongaarts_projection_1987 used a similar approach to estimate descendants in his 'Family Status Model'.
+Recently @Caswell2019 reformulated these relationships in a matrix framework, extending the potential results to variance estimation, multistate models, and others.
 
@@
 In the context of the global pandemic caused by the Covid-19 disease, elderly people depend on support provided by their relatives, especially during periods of lockdown and in the absence of governmental support mechanisms.<!-- Maybe here are repeated --> Elderly parents may be at a higher risk of losing the key support provided by their adult children given the known age-gradient in the Covid-19 case fatality rate.
+
 How do changes in mortality affect the availability of offspring over age from the point of view of a prospective mother?
@@
 In the remainder of the paper, we aim to formalize the relationship population-level changes in mortality rates and perceived change in the lived experience of child death from the perspective of a mother.
-Our approach is key for understanding the impact of age-specific excess mortality on the resilience or otherwise of kinship networks for the vulnerable elderly population.
+Our approach is relevant for understanding the impact of overall and age-specific excess mortality on the resilience or otherwise of kinship networks for the vulnerable elderly population.
 We consider the case when these changes are additive and mutliplicative.
@@
 
-As an initial step, we seek an intuitive understanding of the effect of changing mortality on child survival. 
+As an initial step, we seek an intuitive understanding of the previous relation.
 Building on work by @KeyfitzMath for the probability of a living mother, we can show which features of the child survival function explain the expected number of children.
@@
 
-We find that, seen from the perspective of a mother, child survival mainly depends on the cummulative fertility function^[Different from the findings of @KeyfitzMath of having living ancestors.] and the survival from birth to $a-\kappa$.
-Also that the approximation is affected (negativally) by how disperse is the fertility by age (variance $\sigma^2$) and the suvival curvature around (with negative sign), in an age range where $l_x$ is tipycally very flat (range 20-40 years old) in transitioned populations^[This approximation is useful also for get a dimension changes in the fertility average age \kappa: taking logs in $F_a \, l_{a-\kappa}$ and deriving we get $\frac{\Delta CS_a}{CS_a} \approx - \mu_{a-\kappa} \Delta\kappa$].
+We find that, seen from the perspective of a mother, child survival mainly depends on the cummulative fertility function and the survival of daughters from birth to $a-\kappa$.
+Also that the approximation is affected (negativally) by how disperse is the fertility by age (variance $\sigma^2$) and the survival curvature around (with negative sign), in an age range where $l_x$ is tipycally very flat (range 20-40 years old) in transitioned populations^[This approximation is useful also for get a dimension changes in the fertility average age $\kappa$, keeping the level: taking logs in $F_a \, l_{a-\kappa}$ and deriving we get $\frac{\Delta CS_a}{CS_a} \approx - \mu_{a-\kappa} \Delta\kappa$].
 
@@
 
-$$\frac{dCS^{\delta}}{d\delta} = -a\int_{\alpha}^{a}m_{x}{l_{a-x} dx} + \int_{\alpha}^{a} m_{x} {l_{a-x}x dx}\\
+$$\frac{dCS^{\delta}}{d\delta} = -a\int_{\alpha}^{a}m_{x}{l_{a-x} dx} + \int_{\alpha}^{a} x m_{x} {l_{a-x} dx}\\
 = -a \, CS_{a}  + \int_{\alpha}^{a} {x  \, m_{x} l_{a-x} dx}$$
@@
 The expected change in descendants survival is inversely proportional to the difference between maternal age $a$ and the mean age of the mother at the birth of her surviving daughters $k_a$.
-The magnitude of the change depends negatively on the age distribution of the surviving offspring.
-This is intuitive considering that younger descendants experiences longer periods of exposure to risk.
+The magnitude of the change depends negatively on the age distribution of the surviving offspring (the younger the fertility bigger the impact), which is intuitive considering that older descendants experiences longer periods of exposure to risk.
 
 That an equal change in mortality could happens at all ages is unlikelihood.
-This measure could be much related a change in first ages (child and youth) than adult or ones.
-As a corollary, if the change only happens in the age range $\left[0;1\right)$, infant mortality, we can inspect the effect starting with splitting the integral (considering those born last year and not, and of course for $a>1$):
+This measure could be much related a change in first ages (child and youth) or adolescent ones.
+If the change only happens in the age range $\left[0;1\right)$, infant mortality, we can inspect the effect in child survival $CS_{a}^{\delta_0}$ starting with splitting the integral between those daughters who were exposed all the range and those who do not (now generalizing for $\alpha$ positive and $a>1$ for simplicity):
 
@@
 
-$$CS_{a}^{\delta_0} =-\int_{0}^{a-1} m_{x} l_{a-x} dx - a \int_{a-1}^{a} {m_{x} l_{a-x} dx} + \int_{a-1}^{a} {x \, m_{x} l_{a-x} dx}$$
-If we name $CS_{a-1,a}=\int_{a-1}^{a} {m_{x} l_{a-x} dx}$, we can express the first term in the right as $CS_a - CS_{a-1,a}$, and finally divide by $CS_a$: 
-$$ = - CS_a + CS_{a-1,a} (1-a+\kappa_{a-1,a})\\
-\frac{\Delta CS_{a}}{CS_{a}} \approx - \left[1-\frac{CS_{a-1,a}}{CS_a} (1-(a-\kappa_{a-1,a}))\right] \Delta\delta$$
+$$\frac{dCS^{\delta_0}}{d\delta_0} =-\int_{0}^{a-1} m_{x} l_{a-x} dx - a \int_{a-1}^{a} {m_{x} l_{a-x} dx} + \int_{a-1}^{a} {x \, m_{x} l_{a-x} dx}$$
+If we name $CS_{a-1,a}=\int_{a-1}^{a} {m_{x} l_{a-x} dx}$, we can express the first term in the right as $CS_a - CS_{a-1,a}$, and recognizing that last term could be expressed as  $\kappa_{a-1,a} CS_{a-1,a}$, we get:
 
-This means that an absolute change in infant hazard rate affects proportionally for all the age range extension (because of all the cohorts aged more than 1), except for the mean time exposure for those who born and survive in the last year $1-(a-k_{a-1,a})$, times $\frac{CS_{a-1,a}}{CS_a}$ the portion of cummulative alive daughters for the same period, weighting how important was this in total successful experience (in terms of alive descendants).
+$$ \frac{dCS^{\delta_0}}{d\delta_0} = - CS_a + CS_{a-1,a} (1-a+\kappa_{a-1,a})$$
+Where the factor $1-(a-\kappa_{a-1,a})$, which lies between 0 and 1, is equal to the average portion of time spent with living daughters durings age *a*.
+Assuming uniform distribution of those births, discretizing the change and dividing by $CS_a$ we get the final expression for change in infant mortality:
+
+$$\frac{\Delta CS_{a}^{\delta_0}}{CS_{a}} \approx - \left[1-\frac{CS_{a-1,a}}{CS_a}\frac{1}{2}\right] \Delta\delta$$
+
+This means that an absolute change in infant hazard rate affects proportionally for all the age range extension (cohorts aged more than 1), except for "half" the portion $\frac{CS_{a-1,a}}{CS_a}$ on how important was this las age in total "successful" experience (in terms of alive descendants).
+If *a* is near *\beta* then $\frac{\Delta CS_{a}^{\delta_0}}{CS_{a}} \approx -\Delta\delta$.
 
@@
 
+Following this, in populations with high infant mortality $MAD_{a-x}<1$ for all *x*, so $MAL_a\approx\kappa_a+f_0$, being $f_0$ the average time spent for those child that dies in their first year of life.
 A numerical approximation of both measures are done in [Applications](#Applications) section.
 
-### Heterogeneity (first ideas, some guidance is needed)
+### Heterogeneity (first ideas)
+
+<!--quiero ver la correlacion entre fecundidad y mortalidad infantil. 
+la proporción de hijas muertas es la proporción de hermanas muertas-->
 
@@
 Isolating for an age *x* we can inspect the part related to z and k: $(l_{a-x})^z \, k\, f_{kz}$. 
-Following @Gupta2018, one way to create a correlated bivariate distribution is assume for example that $Z = Y_0 + Y_1$ and $K = e^{(Y_0 + Y_2)}$, given the fact that the historical relation is not linear between both components (see figure \@ref(fig:plot_tfr_q0), a funtion *f* like $F_\beta=f(ln(q_{0,5}))$.
+Following @Gupta2018, one way to create a correlated bivariate distribution is assume for example that $Z = Y_0 + Y_1$ and $K = e^{(Y_0 + Y_2)}$, given the fact that the historical relation is not linear between both components (see figure \@ref(fig:plot_tfr_q0), a funtion *f* like $F_\beta=f(ln(q_{0,5}))$.
 The auxiliar variables $Y_0$, $Y_1$ and $Y_2$ are independet and Gamma distribuited.
@@
 
-...
+To continue...
+
+<!--
 
@@
 
-...
+-->
 
@@
 
-The approximation proposed for CS_a is reaalyprecise, especially for recent periods.
-For exaple, Guatemala improved the approximation with years due to rectangularization process in $l_x$ (figure \@ref(fig:plot_CS_aprox)).
+The approximation proposed for CS_a is realy precise, especially for recent periods.
+For example, Guatemala improved the approximation with years due to rectangularization process in $l_x$ (figure \@ref(fig:plot_CS_aprox)).
 
@@
 
-```{r CS_abs_app, fig.cap= "Effect of an absolute absolute change in mortality on child survival by maternal age of women in Latin America  for the 1950-2015 period.
+```{r CS_abs_app, fig.cap= "Effect of absolute and relative change in mortality on child survival by maternal age of women in Latin America for the 1950-2015 period.
 Goodness of fit of approximation compared to direct estimation"}
@@
              geom_abline(slope=1, intercept = 0) +
-             scale_x_continuous("Empirical change in CS") +
-             scale_y_continuous("Approximated change in CS") +
+             scale_x_continuous("Empirical") +
+             scale_y_continuous("Approximated") +
              scale_color_continuous("Delta") +
+             ggtitle("Absolute change")+
              theme_classic()
@@
              geom_abline(slope=1, intercept = 0) +
-             scale_x_continuous("Empirical change in CS") +
-             scale_y_continuous("Approximated change in CS") +
+             scale_x_continuous("Empirical") +
+             scale_y_continuous("Approximated") +
              scale_color_continuous("Delta") +
+             ggtitle("Relative change") +
              theme_classic()
@@
 grid.arrange(ch_abs, ch_rel, ncol=2)
-
 ```

### 2020-08-15 paper.Rmd
mods for alap

@@
 output:
+  pdf_document: default
   html_document:
@@
 
-$$CS_a \approx l_{a-\kappa} \int_{0}^a{m_{x}dx}+(l_{a-\kappa})^{'}\int_{0}^a{(x-\kappa)m_{x}dx}+
+$$
+\begin{aligned}
+CS_a \approx & l_{a-\kappa} \int_{0}^a{m_{x}dx}+(l_{a-\kappa})^{'}\int_{0}^a{(x-\kappa)m_{x}dx}+
 (l_{a-\kappa})^{''}\int_{0}^a{\frac{(x-\kappa)^2}{2} m_{x}dx}\\
-  \approx F_a l_{a-\kappa} + \frac{\sigma^2}{2}F_a (l_{a-\kappa})^{''}\\
-  \approx F_a \, l_{a-\kappa}[1+\frac{\sigma^2}{2}\frac{(l_{a-\kappa})^{''}}{l_{a-\kappa}}]$$
+  & \approx F_a l_{a-\kappa} + \frac{\sigma^2}{2}F_a (l_{a-\kappa})^{''}\\
+  & \approx F_a \, l_{a-\kappa}[1+\frac{\sigma^2}{2}\frac{(l_{a-\kappa})^{''}}{l_{a-\kappa}}]
+\end{aligned}
+$$
 
@@
 We find that, seen from the perspective of a mother, child survival mainly depends on the cummulative fertility function and the survival of daughters from birth to $a-\kappa$.
-Also that the approximation is affected (negativally) by how disperse is the fertility by age (variance $\sigma^2$) and the survival curvature around (with negative sign), in an age range where $l_x$ is tipycally very flat (range 20-40 years old) in transitioned populations^[This approximation is useful also for get a dimension changes in the fertility average age $\kappa$, keeping the level: taking logs in $F_a \, l_{a-\kappa}$ and deriving we get $\frac{\Delta CS_a}{CS_a} \approx - \mu_{a-\kappa} \Delta\kappa$].
+The approximation is affected negativally by how disperse is the fertility by age (variance $\sigma^2$) and the survival curvature around (with negative sign), in an age range where $l_x$ is tipycally very flat in transitioned populations (range 20-40 years old)^[This approximation is useful also for get an idea of impact because of changes in the fertility average age $\kappa$, keeping constant the level: taking logs in $F_a \, l_{a-\kappa}$ and deriving we get $\frac{\Delta CS_a}{CS_a} \approx \mu_{a-\kappa} \Delta\kappa$, being daughters exposed to less time beacuse of the delay in pattern].
 
@@
 
-$$\frac{dCS^{\delta}}{d\delta} = -a\int_{\alpha}^{a}m_{x}{l_{a-x} dx} + \int_{\alpha}^{a} x m_{x} {l_{a-x} dx}\\
-= -a \, CS_{a}  + \int_{\alpha}^{a} {x  \, m_{x} l_{a-x} dx}$$
+$$
+\begin{aligned}
+\frac{dCS^{\delta}}{d\delta} = -a\int_{\alpha}^{a}m_{x}{l_{a-x} dx} + \int_{\alpha}^{a} x m_{x} {l_{a-x} dx}\\
+= -a \, CS_{a}  + \int_{\alpha}^{a} {x  \, m_{x} l_{a-x} dx}
+\end{aligned}
+$$
 
@@
 
-$$CS_{a}^{\delta_0} = \int_{0}^{a-1} m_{x} {e{^{-\int_{1}^{a-x}\mu_t \,dt-\int_{0}^{1}(\mu_t + \delta)\,dt}}} dx + \int_{a-1}^{a} m_{x} e^{-\int_{0}^{a-x}(\mu_t+ \delta)\,dt}dx\\
-=\int_{0}^{a-1} m_{x} l_{a-x}e^{-\delta} dx + \int_{a-1}^{a} m_{x} l_{a-x} e^{-\delta(a-x)}\,dx$$
-
+$$
+\begin{aligned}
+CS_{a}^{\delta_0} = \int_{0}^{a-1} m_{x} {e{^{-\int_{1}^{a-x}\mu_t \,dt-\int_{0}^{1}(\mu_t + \delta)\,dt}}} dx + \int_{a-1}^{a} m_{x} e^{-\int_{0}^{a-x}(\mu_t+ \delta)\,dt}dx \\
+=\int_{0}^{a-1} m_{x} l_{a-x}e^{-\delta} dx + \int_{a-1}^{a} m_{x} l_{a-x} e^{-\delta(a-x)}\,dx
+\end{aligned}
+$$
 Deriving by $\delta$ and valuating near 0, we get:
@@
 This means that an absolute change in infant hazard rate affects proportionally for all the age range extension (cohorts aged more than 1), except for "half" the portion $\frac{CS_{a-1,a}}{CS_a}$ on how important was this las age in total "successful" experience (in terms of alive descendants).
-If *a* is near *\beta* then $\frac{\Delta CS_{a}^{\delta_0}}{CS_{a}} \approx -\Delta\delta$.
+If $a$ is near $\beta$ then $\frac{\Delta CS_{a}^{\delta_0}}{CS_{a}} \approx -\Delta\delta$.
 
@@
 
-$$\frac{d CS_{a}^\delta}{d \delta} = \int_{0}^{a} {m_{x} l_{a-x} \log(l_{a-x}) dx}\\
+$$
+\begin{aligned}
+\frac{d CS_{a}^\delta}{d \delta} = \int_{0}^{a} {m_{x} l_{a-x} \log(l_{a-x}) dx}\\
 = - \int_{0}^{a} {m_{x} l_{a-x} \int_{0}^{a-x}{\mu_t \, dt}\, dx}\\
-= - \int_{0}^{a}{\mu_t} \int_{0}^{a-t} {m_{x} l_{a-t-x} \, dx}\, dt$$
+= - \int_{0}^{a}{\mu_t} \int_{0}^{a-t} {m_{x} l_{a-t-x} \, dx}\, dt
+\end{aligned}
+$$
 
@@
 Dividing by $CS_a$ and multiplying by $\delta$ get one useful expression.
-This is the negative of cumulative hazard $H_a$ but considering a positive factor $\frac{CS_{a-t}}{CS_a}$ that takes in account the relative amount of surviving descendants that would be lost at each child age at risk.
+This is the negative of cumulative hazard $H_a$ but considering a positive factor $\frac{CS_{a-t}}{CS_a}$ that takes in account the relative amount of surviving descendants that would be lost at each child age at risk, that gives more weight to first ages because all the parity was exposed in that age.
 
@@
 Grasping this relationship intuitively is more difficult given the interaction of birth and mortality rates (@KeyfitzMath). 
-This factor has a S-shape because $l_x$ curvature and fertility accumulation, that gives more weight to first ages, generally less than 1 except for mother ages close to age valuation *a* with complete reproductive life and high infant mortality (no additional children and high risk for one age additional of survive), typically pretransitional countries with a near 50.
-A numerical evaluation of both analytical expressions (absolute an relative) are done in [Applications](#Applications) section.
+The factor $\frac{CS_{a-t}}{CS_a}$  has a S-shape because $l_x$ curvature and fertility accumulation.
+Taking the difference between consecutive ages $a$ and $a+1$, we have $CS_{a-1}=\int_{0}^{a-1} {m_{x} l_{a-1} \, dx}$, $CS_{a}=\int_{0}^{a-1} {m_{x} l_{a} \, dx}+CS_{a-1,a}$.
+In the last one, a second term is added because of additional born daughters after $a-1$, and in the first term the existing one are survived at on $\frac{l_{a-x}}{l_{a-1-x}}$.
+In advanced ages when fertility is complete, the difference is only explained by the surviving part, that is why this factor is not a strictly increasing function but for mother ages close to $\beta$ (last theoretical fecundity age).
+A numerical evaluation of both analytical expressions (absolute an relative) and its factor $\frac{CS_{a-t}}{CS_a}$ are done in [Applications](#Applications) section.
 
@@
 
-$$MAL_a = \frac{\int_0^a{m_x {\frac{\int_0^{a-x} l_t \, \mu_t (x+t) dt}{\int_0^{a-x} l_t \, \mu_t dt}}dx}}{\int_0^a{m_x}dx} \\
-  MAL_a = \frac{\int_0^a{m_x \left[{x+\frac{\int_0^{a-x} l_t \, \mu_t \, t \, dt}{\int_0^{a-x} l_t \, \mu_t dt}}\right]dx}}{\int_0^a{m_x}} \\
-  MAL_a = \frac{\int_0^a{m_x \,x \,dx}}{F_a} + \frac{\int_0^a{m_x MAD_{a-x}dx}}{F_a} \\
-  MAL_a = \kappa + \frac{\int_0^a{m_x MAD_{a-x}dx}}{F_a}$$
+$$
+\begin{aligned}
+MAL_a = & \frac{\int_0^a{m_x {\frac{\int_0^{a-x} l_t \, \mu_t (x+t) dt}{\int_0^{a-x} l_t \, \mu_t dt}}dx}}{\int_0^a{m_x}dx} \\
+  MAL_a = & \frac{\int_0^a{m_x \left[{x+\frac{\int_0^{a-x} l_t \, \mu_t \, t \, dt}{\int_0^{a-x} l_t \, \mu_t dt}}\right]dx}}{\int_0^a{m_x}} \\
+  MAL_a = & \frac{\int_0^a{m_x \,x \,dx}}{F_a} + \frac{\int_0^a{m_x MAD_{a-x}dx}}{F_a} \\
+  MAL_a = & \kappa + \frac{\int_0^a{m_x MAD_{a-x}dx}}{F_a}
+\end{aligned}
+$$
 
@@
 
-$$\overline{CS}_a = \int_{0}^{\infty}\int_{0}^{\infty}\left[\int_{0}^{a}{ {m(k)}_x l(z)_{a-x}\,dx}\right]f_{kz}\,dz \,dk\\
-= F_\beta \int_{0}^{a-x} r_x \int_{0}^{\infty} \int_{0}^{\infty} {(l_{a-x})^z} \, k\, f_{kz} \,dk\,dz\,dx$$
+$$
+\begin{aligned}
+\overline{CS}_a = \int_{0}^{\infty}\int_{0}^{\infty}\left[\int_{0}^{a}{ {m(k)}_x l(z)_{a-x}\,dx}\right]f_{kz}\,dz \,dk\\
+= F_\beta \int_{0}^{a-x} r_x \int_{0}^{\infty} \int_{0}^{\infty} {(l_{a-x})^z} \, k\, f_{kz} \,dk\,dz\,dx
+\end{aligned}
+$$
 
@@
 
-$$\int_{0}^{\infty} \int_{0}^{\infty} {(l_{a-x})^z \, k\, f_{kz} \,dk\,dz}\\
+$$
+\begin{aligned}
+\int_{0}^{\infty} \int_{0}^{\infty} {(l_{a-x})^z \, k\, f_{kz} \,dk\,dz}\\
 \int_{0}^{\infty} \int_{0}^{\infty}{e^{-H_x(Y_0 + Y_1)}e^{Y_0 + Y_2}}\\
-L_{Y_0}[1-H_x]L_{Y_1}[-H_x]L_{Y_2}\,dY_0\,dY_1\,dY_2$$
+L_{Y_0}[1-H_x]L_{Y_1}[-H_x]L_{Y_2}\,dY_0\,dY_1\,dY_2
+\end{aligned}
+$$
 
@@
 The figure \@ref(fig:CS_abs_app) shows that goodness of fit is decreasing with the change size, given that $\delta$ is assumed near zero.
-To ilustrate that considering $\delta=0.01$ means at age 50, it means for the absolute change case that $\frac{l_{50}}{e^{\delta\,50}} = l_{50}/$`r round(exp((50*0.01)),1)`, a big change).
+To ilustrate that considering $\delta=0.01$ means at age 50, it means for the absolute change case that $\frac{l_{50}}{e^{\delta\,50}} = l_{50}/$ `r round(exp((50*0.01)),1)`, a big change).
 For the relative case was assumed a constant within each interval $\mu_{x+t}=\mu_x$ for $t$ between 0 and 1, $\int_{x}^{x+1}{\mu_t dt} = \log(l_{x+1})-\log(l_{x})$, and was used this approximation on the empirical data $\frac{\Delta CS_{a}}{CS_a} \approx - \left[\sum_{0}^{a-1}{[\log(l_{x+1})-\log(l_{x})]\, \frac{CS_{a-1-x}}{CS_a}\,}\right] \Delta \delta$.
@@
 grid.arrange(ch_abs, ch_rel, ncol=2)
+```
+
+*More develop here showing the factor and ggplot type...*
+
+```{r chek_CS_ratio, include = F, eval=F}
+# test factor
+countries = c("Guatemala", "Argentina", "Haiti")
+period = "1950-1955"
+Age = 50
+plot(NULL, xlim=c(15,Age), ylim=c(0,1.1),xlab="Age", ylab="Ratio")
+abline(h=1, lty=2, col="grey")
+for (country in countries){
+  points(15:Age,
+         sapply(15:Age, function(age)
+         CS_results(country = country, 
+             period = period, 
+             a = age)$CS)/
+          CS_results(country = country, 
+             period = period, 
+             a = Age)$CS,
+          col = which(country==countries), t="l")
+}
 ```

### 2020-09-03 paper.tex
not sure

@@
   pdftitle={Child survival for mothers: mortality change and related measures},
-  pdfauthor={Ivan Williams and Diego Alburez-Gutierrez},
+  pdfauthor={Iván Williams and Diego Alburez-Gutierrez},
   hidelinks,
@@
 \usepackage[margin=1in]{geometry}
-\usepackage{graphicx}
+\usepackage{graphicx,grffile}
 \makeatletter
@@
 \setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newenvironment{cslreferences}%
-  {\setlength{\parindent}{0pt}%
-  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
-  {\par}
 
-\title{Child survival for mothers: mortality change and related
-measures}
-\author{Ivan Williams and Diego Alburez-Gutierrez}
-\date{07 August 2019}
+\title{Child survival for mothers: mortality change and related measures}
+\author{Iván Williams and Diego Alburez-Gutierrez}
+\date{May 2020}
 
@@
 
-Kin count estimation has a long and distinguished pedigree in
-mathematical demography, starting with the work of Lotka (1931) on
-modelling orphanhood in theoretical populations across demographic
-regimes. To the best of our knowledge, Brass (1953) first proposed a
-formula to represent child survival over age. The notion of estimating
-the expected number of living daughters in a stable population was
-generalized by Goodman, Keyfitz, and Pullum (1974) for other kin
-relations (granddaughters, cousins, etc.). The ``counting method''
-approach was further popularized in Keyfitz and Caswell (2005).
-Bongaarts (1987) used a similar approach to estimate descendants in his
-`Family Status Model'.
-
-The question of child survival, particularly the survival of daughters
-for mothers, sits at the very center of demographic theory. This is
-evidenced by the omnipresence of the `net reproductive rate' in accounts
-of human and non-human populations. Historical demographers draw
-liberally on assumptions about kin availability and individual's
-exposure to offspring mortality to explain fertility decisions,
-especially around the demographic transition. These assumptions are
-often untested given the data sparsity (Livi Bacci 1997; Volk and
-Atkinson 2013). Economic theories of fertility that consider children as
-`investment' for old-age support also rely on notions of child survival
-(Preston and others 1978).
-
-Child survival - and it's counterpart, child death - gain new
-significance in the wake of the fertility transition. Depressed
-fertility means that fewer children are tasked with providing key
-emotional, social, and financial transfers to aging parents for ever
-increasing periods of time. With increasing periods of generational
-overlap, individuals find themselves `sandwiched' between aging parents
-and young children requiring their simultaneous attention and care
-(Daatland, Veenstra, and Lima 2010). In the context of global population
-aging, elderly parents without access to formal social security and
-pension systems are particularly reliant on these transfers to make ends
-meet (Smith-Greenaway and Trinitapoli 2020).\\
-Losing an only child may become a more common experience in the context
-of global fertility decline, something particularly worrying for parents
-suddenly rendered childless. Given the known psychological, health, and
-social consequences of child loss for bereaved parents and families
+The question of child survival for mothers sits at the very center of
+demographic theory. Offspring survival is usually studied alongside
+fertility decline in the context of the demographic transition, but it
+also matters for post-transitional societies. Lower fertility and higer
+life expectancy means that fewer children are expected to provide key
+emotional, social, and financial transfers to aging parents for longer
+periods of time. With increasing periods of generational overlap,
+individuals find themselves `sandwiched' between aging parents and young
+children requiring their simultaneous attention and care (Daatland,
+Veenstra, and Lima 2010). In the context of global population aging,
+elderly parents without access to formal social security and pension
+systems are particularly reliant on these transfers (Smith-Greenaway and
+Trinitapoli 2020). Given the known psychological, health, and social
+consequences of child loss for bereaved parents and families
 (Hendrickson 2009; Lee et al. 2014), it is surprising that the
 demographic processes that shape parental bereavement remain very poorly
-understood. Recent development have underscored the need to understand
-how sudden changes in mortality affect the availability of kin. In the
-context of the global pandemic caused by the Covid-19 disease, elderly
-people depend on support provided by their relatives, especially during
-periods of lockdown and in the absence of governmental support
-mechanisms. Elderly parents may be at a higher risk of losing the key
-support provided by their adult children given the known age-gradient in
-the Covid-19 case fatality rate. How do changes in mortality affect the
-availability of offspring over age from the point of view of a
-prospective mother? How can we characterize the timing of offspring
-mortality as experienced by a mother from a formal demographic
-perspective? How this would work in an heterogeneity population? In the
-remainder of the paper, we aim to formalize the relationship
-population-level changes in mortality rates and perceived change in the
-lived experience of child death from the perspective of a mother. Our
-approach is key for understanding the impact of age-specific excess
-mortality on the resilience or otherwise of kinship networks for the
-vulnerable elderly population. We consider the case when these changes
-are additive, mutliplicative, and when they are not evenly distributed
-across age. In this paper, we uncover formal relationships to answer
-these questions and provide numerical applications using data from
-countries in Latin America during the second half of the twentieth
-century. See the \protect\hyperlink{Data}{data section} in the
-\protect\hyperlink{Applications}{Applications} section for more details
-on the source, calculation procedures and main results. We used period
-measures, in an stable context, so the objetive is give moment measures.
-A more sophisticated model using the appropriate age-specific cohort
-rates in the subscripts has been proposed by Alburez-Gutierrez, Kolk,
-and Zagheni (2019).
+understood.
 
-\hypertarget{relationship-proof}{%
-\subsection{Relationship \& Proof}\label{relationship-proof}}
+The Covid-19 pandemic has underscored the need to understand how sudden
+changes in mortality affect the availability of kin. The elderly depend
+on kin support, especially during periods of lockdown and in the absence
+of governmental support mechanisms. Elderly mothers may be at a higher
+risk of losing the key support provided by their adult children given
+the known age-gradient in the Covid-19 case fatality rate. How do
+changes in mortality affect the availability of offspring over age from
+the point of view of a prospective mother? How do these processes play
+out in a heterogeneous population?
+
+In this paper, we aim to formalize the relationship between
+population-level changes in mortality rates (additive or multiplicative)
+and changes in the lived experience of kin death. This is key for
+understanding the impact of age-specific excess mortality on the
+resilience or otherwise of kinship networks for the vulnerable elderly
+population. This paper uncovers formal relationships and provides
+numerical applications using data from Latin America in the second half
+of the twentieth century using period measures in a stable
+context.\footnote{See the \protect\hyperlink{Applications}{Applications}
+  section for more details on the source, calculation procedures and
+  main numerical results.} We continue a long tradition in mathematical
+demography, starting with the work of Lotka (1931) on modelling
+orphanhood in theoretical populations, Brass (1953) on child survival,
+Goodman, Keyfitz, and Pullum (1974) on kin survival in stable
+populations, and recently, Caswell (2019) kinship matrices.
+
+\hypertarget{relationship-and-proofs}{%
+\subsection{Relationship (and proofs)}\label{relationship-and-proofs}}
 
@@
 
-\[
-\begin{aligned}
-CS_a = \int_{0}^a{m_{x}l_{a-x}dx}
-\end{aligned}
-\]
+\[CS_a = \int_{0}^a{m_{x}l_{a-x}dx}\]
 
@@
 
-As an initial step, we seek an intuitive understanding of the effect of
-changing mortality on child survival. Building on work by Keyfitz and
-Caswell (2005) for the probability of a living mother, we show which
-features of the child survival function explain the expected number of
-children. For this we use an approximate of \(l_x\) using Taylor´s
-theorem until second order around the mean age of childbearing
-\(\kappa\) for all \(a>=\alpha\):
+As an initial step, we seek an intuitive understanding of the previous
+relation. Building on work by Keyfitz and Caswell (2005) for the
+probability of a living mother, we can show which features of the child
+survival function explain the expected number of children. For this we
+use an approximate of \(l_x\) using Taylor´s theorem until second order
+around the mean age of childbearing \(\kappa\):
 
-\[CS_a \approx l_{a-\kappa} \int_{\alpha}^a{m_{x}dx}+(l_{a-\kappa})^{'}\int_{0}^a{(x-\kappa)m_{x}dx}+
+\[
+\begin{aligned}
+CS_a \approx & l_{a-\kappa} \int_{0}^a{m_{x}dx}+(l_{a-\kappa})^{'}\int_{0}^a{(x-\kappa)m_{x}dx}+
 (l_{a-\kappa})^{''}\int_{0}^a{\frac{(x-\kappa)^2}{2} m_{x}dx}\\
-  \approx F_a l_{a-\kappa} + \frac{\sigma^2}{2}F_a (l_{a-\kappa})^{''}\\
-  \approx F_a \, l_{a-\kappa}[1+\frac{\sigma^2}{2}\frac{(l_{a-\kappa})^{''}}{l_{a-\kappa}}]\]
+  & \approx F_a l_{a-\kappa} + \frac{\sigma^2}{2}F_a (l_{a-\kappa})^{''}\\
+  & \approx F_a \, l_{a-\kappa}[1+\frac{\sigma^2}{2}\frac{(l_{a-\kappa})^{''}}{l_{a-\kappa}}]
+\end{aligned}
+\]
 
 where the fertility pattern by age is concentrated around \(\kappa\) and
-the accumulated fertility (or gross reproduction rate in our same-sex
-model) is \(F_a = \int_{\alpha}^a{m_{x}dx}\). The second Taylor´s term
-is null because \(\int_{\alpha}^a{x \, m_{x}dx} = \kappa \, F_a\).
+the accumulated fertility (or gross reproduction rate in our
+female-dominant scenario) is \(F_a = \int_{\alpha}^a{m_{x}dx}\). The
+second Taylor´s term is null because
+\(\int_{\alpha}^a{x \, m_{x}dx} = \kappa \, F_a\).
 
 We find that, seen from the perspective of a mother, child survival
-mainly depends on the cummulative fertility function\footnote{Note that
-  this runs contrary to the findings of Keyfitz and Caswell (2005), who
-  found that levels of fertility did not affect the probability of
-  having living ancestors.} and the mean age at childbearing \(\kappa\)
-acts through the factor \(l_{a-\kappa}\). The dispersion of of fertility
-\(\sigma^2\) is negatively proportional to child survival. All else
-constant, a higher dispersion of fertility \(\sigma^2\) is associated
-with an increase in \(CS_a\) during reproductive age. The negative
-relationship stems from the fact that the second derivative of
-\(l_{a-\kappa}\) is negative in the 15-49 age range, given a concave
-downwards survival function.
-
-The second Taylor´s term is null because
-\(\int_{\alpha}^a{x \, m_{x}dx} = \kappa \, F_a\). The formula tells us
-that the survival experience for a woman aged \emph{a} depends mainly of
-cumulated fertility (possible cases, different from the ancestor count
-case where this doesn´t affect) and child survival probability (success
-cases). Also that the approximation is affected by how disperse is the
-fertility by age (variance \(\sigma^2\)) and the suvival curvature
-around (with negative sign), in a \(l_x\) range where is tipycally very
-flat (range 20-40 years old) in transitioned populations.
-
-And change in \(\kappa\), taking logs and deriving:
-\(\frac{dCS_a}{d\kappa}\frac{1}{CS_a} \approx \mu_{a-\kappa}F_a \Delta\kappa\)
+mainly depends on the cummulative fertility function and the survival of
+daughters from birth to \(a-\kappa\). The approximation is affected
+negativally by how disperse is the fertility by age (variance
+\(\sigma^2\)) and the survival curvature around (with negative sign), in
+an age range where \(l_x\) is tipycally very flat in transitioned
+populations (range 20-40 years old)\footnote{This approximation is
+  useful also for get an idea of impact because of changes in the
+  fertility average age \(\kappa\), keeping constant the level: taking
+  logs in \(F_a \, l_{a-\kappa}\) and deriving we get
+  \(\frac{\Delta CS_a}{CS_a} \approx \mu_{a-\kappa} \Delta\kappa\),
+  being daughters exposed to less time beacuse of the delay in pattern}.
 
@@
 mortality in the range \([0,a-\alpha]\), where \(\alpha\) is the start
-age of fertility risk. As a \protect\hyperlink{Corollary}{corollary} we
-consider the case where mortality only changes for infants, considering
-that it is unlikelily for a potential change in mortality to affect all
-ages equally in real populations. For now, let us consider
+age of fertility risk. For now, let us consider
 \(m_{x,\delta}=m_{x}+\delta\) (Wrycza and Baudisch (2012)) and
@@
 
-\[\frac{dCS^{\delta}}{d\delta} = -a\int_{\alpha}^{a}m_{x}{l_{a-x} dx} + \int_{\alpha}^{a} m_{x} {l_{a-x}x dx}\\
-= -a \, CS_{a}  + \int_{\alpha}^{a} {x  \, m_{x} l_{a-x} dx}\]
+\[
+\begin{aligned}
+\frac{dCS^{\delta}}{d\delta} = -a\int_{\alpha}^{a}m_{x}{l_{a-x} dx} + \int_{\alpha}^{a} x m_{x} {l_{a-x} dx}\\
+= -a \, CS_{a}  + \int_{\alpha}^{a} {x  \, m_{x} l_{a-x} dx}
+\end{aligned}
+\]
 
-Dividing both sides by \(CS_{a}\) in an discrete approximation we get
+Dividing both sides by \(CS_{a}\) in an discrete approximation we get:
 
@@
 the difference between maternal age \(a\) and the mean age of the mother
-at the birth of her surviving daughters \(k_a\) (always smaller than
-\(\kappa_a\)). The magnitude of the change depends negatively on the age
-distribution of the surviving offspring. This is intuitive considering
-that younger descendants experiences longer periods of exposure to risk.
-The figure @ref(fig:CS\_abs\_app) shows that goodness of fit is
-decreasing with the change size, given that \(\delta\) is assumed near
-zero (as an example considering \(\delta=0.01\) means at age 50 that
-\(\frac{l_50}{e^{\delta\,50}} = \frac{l_50}{1.6}\), a big change).
+at the birth of her surviving daughters \(k_a\). The magnitude of the
+change depends negatively on the age distribution of the surviving
+offspring (the younger the fertility bigger the impact), which is
+intuitive considering that older descendants experiences longer periods
+of exposure to risk.
 
-That an equal change in mortality could happens at all ages is
-unlikelihood. This measure is much more related to first ages (child and
-youth) than adult or ones. As a corollary, if the change only happens in
-the age range \(\left[0;1\right)\), infant mortality, we can inspect the
-effect starting with splitting the integral (of course, for \(a>1\)){]}:
+However, it is unlikely for mortality to change at the same rate at all
+ages. In the course of the demographic transition, for example, we would
+expect larger changes in young-age mortality. Consider a change in
+infant mortality in the age range \(\left[0;1\right)\). We can inspect
+the effect in child survival \(CS_{a}^{\delta_0}\) by splitting the
+integral between those daughters who were exposed to the mortality
+change and those who were not (now generalizing for \(\alpha\) positive
+and \(a>1\) for simplicity):
 
-\[CS_{a}^{\delta_0} = \int_{0}^{a-1} m_{x} {e{^{-\int_{1}^{a-x}\mu_t \,dt-\int_{0}^{1}(\mu_t + \delta)\,dt}}} dx + \int_{a-1}^{a} m_{x} e^{-\int_{0}^{a-x}(\mu_t+ \delta)\,dt}dx\\
-=\int_{0}^{a-1} m_{x} l_{a-x}e^{-\delta} dx + \int_{a-1}^{a} m_{x} l_{a-x} e^{-\delta(a-x)}\,dx\]
+\[
+\begin{aligned}
+CS_{a}^{\delta_0} = \int_{0}^{a-1} m_{x} {e{^{-\int_{1}^{a-x}\mu_t \,dt-\int_{0}^{1}(\mu_t + \delta)\,dt}}} dx + \int_{a-1}^{a} m_{x} e^{-\int_{0}^{a-x}(\mu_t+ \delta)\,dt}dx \\
+=\int_{0}^{a-1} m_{x} l_{a-x}e^{-\delta} dx + \int_{a-1}^{a} m_{x} l_{a-x} e^{-\delta(a-x)}\,dx
+\end{aligned}
+\] Deriving by \(\delta\) and valuating near 0, we get:
 
-Deriving by \(\delta\) and valuating near 0, we get:
+\[\frac{dCS^{\delta_0}}{d\delta_0} =-\int_{0}^{a-1} m_{x} l_{a-x} dx - a \int_{a-1}^{a} {m_{x} l_{a-x} dx} + \int_{a-1}^{a} {x \, m_{x} l_{a-x} dx}\]
+Given that \(CS_{a-1,a}=\int_{a-1}^{a} {m_{x} l_{a-x} dx}\), we can
+express the first term in the right as \(CS_a - CS_{a-1,a}\). The
+righ-most term could also be expressed as \(\kappa_{a-1,a} CS_{a-1,a}\),
+so that:
 
-\[CS_{a}^{\delta_0} =-\int_{0}^{a-1} m_{x} l_{a-x} dx - a \int_{a-1}^{a} {m_{x} l_{a-x} dx} + \int_{a-1}^{a} {x \, m_{x} l_{a-x} dx}\]
-If we name \(CS_{a-1,a}=\int_{a-1}^{a} {m_{x} l_{a-x} dx}\), we can
-express the first term in the right as \(CS_a - CS_{a-1,a}\), and
-finally divide by \(CS_a\):
-\[ = - CS_a + CS_{a-1,a} (1-a+\kappa_{a-1,a})\\
-\frac{\Delta CS_{a}}{CS_{a}} \approx - \left[1-\frac{CS_{a-1,a}}{CS_a} (1-(a-\kappa_{a-1,a}))\right] \Delta\delta\]
+\[ \frac{dCS^{\delta_0}}{d\delta_0} = - CS_a + CS_{a-1,a} (1-a+\kappa_{a-1,a})\]
+The factor \(1-(a-\kappa_{a-1,a})\), which lies between 0 and 1, is
+equal to the average portion of time spent with living daughters durings
+age \emph{a}. Assuming a uniform distribution of those births,
+discretizing the change and dividing by \(CS_a\) we get the final
+expression for change in infant mortality:
+
+\[\frac{\Delta CS_{a}^{\delta_0}}{CS_{a}} \approx - \left[1-\frac{CS_{a-1,a}}{CS_a}\frac{1}{2}\right] \Delta\delta\]
 
 This means that an absolute change in infant hazard rate affects
-proportionally for all the age range extension (because of all the
-cohorts aged more than 1), except for the mean time exposure for those
-who born and survive in the last year \(1-(a-k_{a-1,a})\), times
-\(\frac{CS_{a-1,a}}{CS_a}\) the portion of cummulative alive daughters
-for the same period, weighting how important was this in total
-successful experience (in terms of alive descendants).
+proportionally for all the age range extension (cohorts aged more than
+1), except for ``half'' the portion \(\frac{CS_{a-1,a}}{CS_a}\) on how
+important was this las age in total ``successful'' experience (in terms
+of alive descendants). If \(a\) is near \(\beta\) then
+\(\frac{\Delta CS_{a}^{\delta_0}}{CS_{a}} \approx -\Delta\delta\).
 
@@
 
-\[\frac{d CS_{a}^\delta}{d \delta} = \int_{0}^{a} {m_{x} l_{a-x} \log(l_{a-x}) dx}\\
+\[
+\begin{aligned}
+\frac{d CS_{a}^\delta}{d \delta} = \int_{0}^{a} {m_{x} l_{a-x} \log(l_{a-x}) dx}\\
 = - \int_{0}^{a} {m_{x} l_{a-x} \int_{0}^{a-x}{\mu_t \, dt}\, dx}\\
-= - \int_{0}^{a}{\mu_t} \int_{0}^{a-t} {m_{x} l_{a-t-x} \, dx}\, dt\]
+= - \int_{0}^{a}{\mu_t} \int_{0}^{a-t} {m_{x} l_{a-t-x} \, dx}\, dt
+\end{aligned}
+\]
 
-The last integral is \(CS_{a-t}\). Dividing by \(CS_a\) and multiplying
-by \(\delta\) get one useful expression. This is the negative of
-cumulative hazard \(H_a\) but considering a positive factor
-\(\frac{CS_{a-t}}{CS_a}\) that takes in account the relative amount of
-surviving descendants that would be lost at each child age at risk.
+Considering that the last integral is equal to \(CS_{a-t}\), we divide
+by \(CS_a\) and multiply by \(\delta\) to obtain:
 
@@
 
+This is the negative of cumulative hazard \(H_a\) but considering a
+positive factor \(\frac{CS_{a-t}}{CS_a}\) that takes in account the
+relative amount of surviving descendants that would be lost at each
+child age at risk, that gives more weight to first ages because all the
+parity was exposed in that age.
+
 Grasping this relationship intuitively is more difficult given the
 interaction of birth and mortality rates (Keyfitz and Caswell (2005)).
-This factor has a S-shape because \(l_x\) curvature and fertility
-accumulation, that gives more weight to first ages, generally less than
-1 except for mother ages close to age valuation \emph{a} with complete
-reproductive life and high infant mortality (no additional children and
-high risk for one age additional of survive), typically some countries
-with \(a=50\) and first periods.\\
+The factor \(\frac{CS_{a-t}}{CS_a}\) has a S-shape because \(l_x\)
+curvature and fertility accumulation. Taking the difference between
+consecutive ages \(a\) and \(a+1\), we have
+\(CS_{a-1}=\int_{0}^{a-1} {m_{x} l_{a-1} \, dx}\),
+\(CS_{a}=\int_{0}^{a-1} {m_{x} l_{a} \, dx}+CS_{a-1,a}\). In the last
+one, a second term is added because of additional born daughters after
+\(a-1\), and in the first term the existing one are survived at on
+\(\frac{l_{a-x}}{l_{a-1-x}}\). In advanced ages when fertility is
+complete, the difference is only explained by the surviving part, that
+is why this factor is not a strictly increasing function but for mother
+ages close to \(\beta\) (last theoretical fecundity age). We provide a
+numerical evaluation of both analytical expressions (absolute an
+relative) in the \protect\hyperlink{Applications}{Applications} section.
 
@@
 mother aged \emph{a}, which can be expressed in terms of a temporary
-expected lost years index in line with \(e^\dagger\) (Vaupel (1986)):
+expected lost years index, in line with \(e^\dagger\) (Vaupel (1986)):
 
@@
 would expect to live with their daughters. We call this the
-\emph{Intensity Time in Bereavement} (\(ITB\))): a ratio between
-expected time with a ``lost'' and expected time with a ``life'', that
-allows to make comparisons between population regimes.
+\emph{Intensity Time in Bereavement} (\(ITB\)): a ratio between expected
+time with a ``lost'' and expected time with a ``life'', that allows to
+make comparisons between population regimes.
 
@@
 Looks similar to the transcendental entropy measure \(H\) (Keyfitz and
-Caswell (2005)) but considering all the cohorts born during the the
-mother´s life, weighted by their relative size \(m_x\). In the figure
+Caswell (2005)) but considering all the cohorts born during the mother´s
+life, weighted by their relative size \(m_x\). In the figure
 @ref(fig:plot\_ITL\_MAL) is shown that \emph{ITB} is bigger for young
@@
 \(MAD_{a-x}\) refers to the mean age at death for newborns that die
-before \(a-x\), \(F_a\) is the accumulated fertility for a women aged
+before \(a-x\) and \(F_a\) is the accumulated fertility for a women aged
 \(a\):
 
-\[MAL_a = \frac{\int_0^a{m_x {\frac{\int_0^{a-x} l_t \, \mu_t (x+t) dt}{\int_0^{a-x} l_t \, \mu_t dt}}dx}}{\int_0^a{m_x}dx} \\
-  MAL_a = \frac{\int_0^a{m_x \left[{x+\frac{\int_0^{a-x} l_t \, \mu_t \, t \, dt}{\int_0^{a-x} l_t \, \mu_t dt}}\right]dx}}{\int_0^a{m_x}} \\
-  MAL_a = \frac{\int_0^a{m_x \,x \,dx}}{F_a} + \frac{\int_0^a{m_x MAD_{a-x}dx}}{F_a} \\
-  MAL_a = \kappa + \frac{\int_0^a{m_x MAD_{a-x}dx}}{F_a}\]
+\[
+\begin{aligned}
+MAL_a = & \frac{\int_0^a{m_x {\frac{\int_0^{a-x} l_t \, \mu_t (x+t) dt}{\int_0^{a-x} l_t \, \mu_t dt}}dx}}{\int_0^a{m_x}dx} \\
+  MAL_a = & \frac{\int_0^a{m_x \left[{x+\frac{\int_0^{a-x} l_t \, \mu_t \, t \, dt}{\int_0^{a-x} l_t \, \mu_t dt}}\right]dx}}{\int_0^a{m_x}} \\
+  MAL_a = & \frac{\int_0^a{m_x \,x \,dx}}{F_a} + \frac{\int_0^a{m_x MAD_{a-x}dx}}{F_a} \\
+  MAL_a = & \kappa + \frac{\int_0^a{m_x MAD_{a-x}dx}}{F_a}
+\end{aligned}
+\]
 
-\hypertarget{heterogeneity}{%
-\subsubsection{Heterogeneity}\label{heterogeneity}}
+Following this, in populations with high infant mortality
+\(MAD_{a-x}<1\) for all \emph{x}, so \(MAL_a\approx\kappa_a+f_0\), being
+\(f_0\) the average time spent for those child that dies in their first
+year of life. A numerical approximation of both measures are done in
+\protect\hyperlink{Applications}{Applications} section.
 
-In a model with heterogeneity, determined at birth with a multiplicative
+\hypertarget{heterogeneity-first-ideas}{%
+\subsubsection{Heterogeneity (first
+ideas)}\label{heterogeneity-first-ideas}}
+
+In a model with heterogeneity, determined at birth by a multiplicative
 effect, \(CS_a\) could be interpreted as a conditional expectation with
@@
 as \({m(k)}_x = F_\beta\,k\,r_x\), where \(F_\beta\) is the baseline
-cumulated fertility until upper limit in fecundity age range \(\beta\),
-\(K\) is a random variable with mean 1 that allows variability between
+cumulated fertility until upper limit in all age range until \(\beta\),
+\(k\) is a random variable with mean 1 that allows variability between
 groups and \(r_x\) is the fertility structure by age
-(\(\int_{0}^{\beta}{r_x}=1\)). Frailty part can be thought in a cohort
-effect way, also in the multiplicative assumption as Vaupel and Missov
-(2014), with \(l(z)_x=e^{-H_{x}Z}=l_x^Z\), with baseline hazard \(\mu\).
-Considering the joint distribution \(f_{kz}\), the intuition could be
-expressed as \(f_{zk}=f_{k|z}f_z\): groups with higher descendant´s
-mortality would adjust their fertility level, with positive correlation.
-Replacing both variables, the unconditional mean \(\overline{CS}_a\)
-would be\footnote{Note that given K and Z values fertility and child
-  survival are independent.}:
+(\(\int_{0}^{\beta}{r_x}=1\)). In the other side, frailty part can be
+thought in a cohort effect way, also in the multiplicative assumption as
+Vaupel and Missov (2014), with \(l(z)_x=e^{-H_{x}z}=l_x^z\), with
+baseline hazard \(\mu\). Considering the joint distribution \(f_{kz}\),
+the intuition could be expressed as \(f_{zk}=f_{k|z}f_z\): groups with
+higher descendant´s mortality would adjust their fertility level, with
+positive correlation. Replacing both variables, the unconditional mean
+\(\overline{CS}_a\) would be\footnote{Note that given k and z values,
+  fertility and child survival are independent.}:
 
-\[\overline{CS}_a = \int_{0}^{\infty}\int_{0}^{\infty}\left[\int_{0}^{a}{ {m(k)}_x l(z)_{a-x}\,dx}\right]f_{kz}\,dz \,dk\\
-= F_\beta \int_{0}^{a-x} r_x \int_{0}^{\infty} \int_{0}^{\infty} {l_{a-x}}^z \, k\, f_{kz} \,dk\,dz\,dx\]
+\[
+\begin{aligned}
+\overline{CS}_a = \int_{0}^{\infty}\int_{0}^{\infty}\left[\int_{0}^{a}{ {m(k)}_x l(z)_{a-x}\,dx}\right]f_{kz}\,dz \,dk\\
+= F_\beta \int_{0}^{a-x} r_x \int_{0}^{\infty} \int_{0}^{\infty} {(l_{a-x})^z} \, k\, f_{kz} \,dk\,dz\,dx
+\end{aligned}
+\]
 
-Isolating for an age \emph{x} we can inspect the part related to Z and
-K. Assume \(Z = Y_0 + Y_1\) and \(K = e^{(Y_0 + Y_2)}\), given the fact
-that the the historical relation is not linear (see figure
-@ref(fig:heterogeneity)), like \(F_\beta=ln(q_{0,5})\). \emph{Y\_0},
-\emph{Y\_1} and \emph{Y\_2} are independet and gamma distribuited. Then
-we can express:
+Isolating for an age \emph{x} we can inspect the part related to z and
+k: \((l_{a-x})^z \, k\, f_{kz}\). Following Gupta (2018), one way to
+create a correlated bivariate distribution is assume for example that
+\(Z = Y_0 + Y_1\) and \(K = e^{(Y_0 + Y_2)}\), given the fact that the
+historical relation is not linear between both components (see figure
+@ref(fig:plot\_tfr\_q0), a funtion \emph{f} like
+\(F_\beta=f(ln(q_{0,5}))\). The auxiliar variables \(Y_0\), \(Y_1\) and
+\(Y_2\) are independet and Gamma distribuited. Then we can express,
+using Laplace transform properties:
 
-\[{l_{a-x}}^z \, k\, f_{kz} \,dk\,dz\\
-e^{-H_x(Y_0 + Y_1)}e^{Y_0 + Y_2}\\
-L_{Y_0}[1-H_x]L_{Y_1}[-H_x]L_{Y_2}\]
+\[
+\begin{aligned}
+\int_{0}^{\infty} \int_{0}^{\infty} {(l_{a-x})^z \, k\, f_{kz} \,dk\,dz}\\
+\int_{0}^{\infty} \int_{0}^{\infty}{e^{-H_x(Y_0 + Y_1)}e^{Y_0 + Y_2}}\\
+L_{Y_0}[1-H_x]L_{Y_1}[-H_x]L_{Y_2}\,dY_0\,dY_1\,dY_2
+\end{aligned}
+\]
+
+To continue\ldots{}
 
@@
 female percentage of newborns of 0.49 for all period-country cases.
-Calculations was done in a discrete way assuming that the \emph{m\_x}
-live borths are born at exact mother´s age \emph{x}. A negative relation
-between mortality until 10 and TFR is shown in
-@ref(fig:plot\_tfr\_q010). Specific paths like Argentina and Guatemala
-shows very different demographic profiles during second part of XX
-century.
+Calculations were done in a discrete way assuming that the \(m_x\) live
+births are born at exact mother´s age \emph{x}. An already known log
+relation between infant mortality and TFR is shown in Latinoamerican
+experience (@ref(fig:plot\_tfr\_q0)).
+
+\begin{figure}
+\centering
+\includegraphics{paper_files/figure-latex/plot_tfr_q0-1.pdf}
+\caption{Female probability of death at birth by gross reproduction
+rate. Latin America countries in period 1950-2015}
+\end{figure}
 
@@
 Consider a Latin American woman standing before us. If this were
-1950-1955, she could reasonable expect to have 1.87 surviving children
+1950-1955, she could reasonable expect to have 2.23 surviving daughters
 on her 50\(^{th}\) birthday. In 2010-2015 a woman the same age would
-only have 1.87 living children. The difference of 0 children is
+only have 1.1 living children. The difference of 1.13 children is
 explained by reduced fertility and improved mortality in the region
@@
 as a proportion of the daughters ever born to a woman that age,
-\(\frac{CS_a}{F_a}\). Given Eq. \ref{eq:CS}, the increase from 0.82 in
-1950-1955 to 0.82 in 2010-2015 must be explained by a change in
+\(\frac{CS_a}{F_a}\). Given Eq. \ref{eq:CS}, the increase from 0.78 in
+1950-1955 to 0.97 in 2010-2015 must be explained by a change in
 mortality.
-
-An interesting corollary of these results is the question: at wich age,
-a mother would have the same \emph{CS} than \emph{a=20}, like a mirror
-effect, \(\int_{0}^{a}{m_xl_{a-x}}dx=\int_{0}^{y}{m_xl_{y-x}}dx\) with
-\(a<\beta<y\). Could be solved numerically or with our approximation
-taking \(F_al_{a-\kappa_a}=F_yl_{y-\kappa_y}\), solving for \emph{y}
-which \(\frac{F_a}{F_y} = \frac{l_{a-\kappa_a}}{l_{y-\kappa_y}}\).
 
@@
 
-Guatemala improved the approximation with years due to
-rectangularization process in \(l_x\) (figure
+The approximation proposed for CS\_a is realy precise, especially for
+recent periods. For example, Guatemala improved the approximation with
+years due to rectangularization process in \(l_x\) (figure
 @ref(fig:plot\_CS\_aprox)).
@@
 
-With \(\mu_{x+t}=\mu_x\) for \(t\) between 0 and 1,
-\(\int_{x}^{x+1}{\mu_t dt} = M_x\), so discretizing
-\(\frac{\Delta CS_{a}}{CS_a} \approx - \left[\sum_{0}^{a-1}{M_x \, \frac{CS_{a-1-x}}{CS_a}\,}\right] \Delta \delta\).
+The figure @ref(fig:CS\_abs\_app) shows that goodness of fit is
+decreasing with the change size, given that \(\delta\) is assumed near
+zero. To ilustrate that considering \(\delta=0.01\) means at age 50, it
+means for the absolute change case that
+\(\frac{l_{50}}{e^{\delta\,50}} = l_{50}/\) 1.6, a big change). For the
+relative case was assumed a constant within each interval
+\(\mu_{x+t}=\mu_x\) for \(t\) between 0 and 1,
+\(\int_{x}^{x+1}{\mu_t dt} = \log(l_{x+1})-\log(l_{x})\), and was used
+this approximation on the empirical data
+\(\frac{\Delta CS_{a}}{CS_a} \approx - \left[\sum_{0}^{a-1}{[\log(l_{x+1})-\log(l_{x})]\, \frac{CS_{a-1-x}}{CS_a}\,}\right] \Delta \delta\).
+
+\begin{figure}
+\centering
+\includegraphics{paper_files/figure-latex/CS_abs_app-1.pdf}
+\caption{Effect of absolute and relative change in mortality on child
+survival by maternal age of women in Latin America for the 1950-2015
+period. Goodness of fit of approximation compared to direct estimation}
+\end{figure}
+
+\emph{More develop here showing the factor and ggplot type\ldots{}}
 
@@
 1950-1955 Haiti women aged 30 would have experienced an intensity of
-0.5\%, and in 2010-2015 0.5\%, while the women of Costa Rica 0.5\%, and
-in 2010-2015 0.5\%. For those Latin American mothers aged 50 with 3
+4.8\%, and in 2010-2015 1.4\%, while the women of Costa Rica 2.3\%, and
+in 2010-2015 0.2\%. For those Latin American mothers aged 50 with 3
 daughters born who suffered a lost, they experienced that at age 30 in
-average (figure @ref(fig:plot\_ITL\_MAL)) (More can be analyzed here).
+average (figure @ref(fig:plot\_ITL\_MAL)).
 
@@
 \includegraphics{paper_files/figure-latex/plot_ITL_MAL-1.pdf}
-\caption{Intensity Tome Lost of womans aged 30, 40 and 50. Years 1950
-(darker) to 2015 (lighter). Latinamerican countries in period 1950-2015}
+\caption{a) Intensity Tome Lost of women aged 30, 40 and 50. b) Mean age
+at Lost for women aged 40. Years 1950 (darker) to 2015 (lighter).
+Latinamerican countries in period 1950-2015}
 \end{figure}
@@
 \hypertarget{refs}{}
-\begin{cslreferences}
-\leavevmode\hypertarget{ref-alburez_gutierrez_womens_2019}{}%
-Alburez-Gutierrez, Diego, Martin Kolk, and Emilio Zagheni. 2019.
-``Women's Experience of Child Death over the Life Course: A Global
-Demographic Perspective.'' Preprint. SocArXiv.
-\url{https://doi.org/10.31235/osf.io/s69fz}.
-
-\leavevmode\hypertarget{ref-bongaarts_projection_1987}{}%
-Bongaarts, John. 1987. ``The Projection of Family Composition over the
-Life Course with Family Status Life Tables.'' In \emph{Family
-Demography: Methods and Their Application}, edited by John Bongaarts and
-Thomas Burch. Oxford: IUSSP Series, Oxford Univer- sity Pres.
-
 \leavevmode\hypertarget{ref-brass_derivation_1953}{}%
@@
 Studies} 7 (2): 137. \url{https://doi.org/10.2307/2172029}.
+
+\leavevmode\hypertarget{ref-caswell_formal_2019}{}%
+Caswell, Hal. 2019. ``The Formal Demography of Kinship: A Matrix
+Formulation.'' \emph{Demographic Research} 41 (September): 679--712.
+\url{https://doi.org/10.4054/DemRes.2019.41.24}.
 
@@
 
+\leavevmode\hypertarget{ref-Gupta2018}{}%
+Gupta, R. C. 2018. ``Association measures in the bivariate correlated
+frailty model.'' \emph{ResearchGate} 16 (2): 257--78.
+
 \leavevmode\hypertarget{ref-Hendrickson2009}{}%
@@
 
-\leavevmode\hypertarget{ref-livi_bacci_concise_1997}{}%
-Livi Bacci, Massimo. 1997. \emph{A Concise History of World Population}.
-2nd ed. Cambridge, MA: Blackwell.
-
 \leavevmode\hypertarget{ref-lotka1931orphanhood}{}%
@@
 \url{https://doi.org/https://doi.org/https://doi.org/10.18356/3e9d869f-en}.
-
-\leavevmode\hypertarget{ref-preston1978effects}{}%
-Preston, Samuel H, and others. 1978. \emph{The Effects of Infant and
-Child Mortality on Fertility.} Academic Press, Inc., 111 Fifth Avenue,
-New York/New York 10003, USA.
 
@@
 
-\leavevmode\hypertarget{ref-volk_infant_2013}{}%
-Volk, Anthony A., and Jeremy A. Atkinson. 2013. ``Infant and Child Death
-in the Human Environment of Evolutionary Adaptation.'' \emph{Evolution
-and Human Behavior} 34 (3): 182--92.
-\url{https://doi.org/10.1016/j.evolhumbehav.2012.11.007}.
-
 \leavevmode\hypertarget{ref-wrycza2012}{}%
@@
 Research} 27: 365--76.
-\end{cslreferences}
 

### 2020-09-03 paper_diego_old.Rmd
not sure

--- 2020-06-02 paper.Rmd
@@
 author: "Iván Williams and Diego Alburez-Gutierrez"
-date: "May 2020"
+date: "07 July 2020"
 output:
@@
 
-The question of child survival, particularly the survival of daughters for mothers, sits at the very center of demographic theory.
-This is evidenced by the omnipresence of the 'net reproductive rate' in accounts of human and non-human populations. 
-Historical demographers draw liberally on assumptions about kin availability and individual's exposure to offspring mortality to explain fertility decisions, especially around the demographic transition.
-These assumptions are often untested given the data sparsity [@livi_bacci_concise_1997;@volk_infant_2013]. 
-Economic theories of fertility that consider children as 'investment' for old-age support also rely on notions of child survival [@preston1978effects].
+The question of child survival for mothers sits at the very center of demographic theory.
+<!-- This is evidenced by the omnipresence of the 'net reproductive rate' in accounts of human and non-human populations. -->
+<!-- Historical demographers draw liberally on assumptions about kin availability and individual's exposure to offspring mortality to explain fertility decisions, especially around the demographic transition [@gortfelder_survival_2020].
+-->
+<!-- These assumptions are often untested given the data sparsity [@livi_bacci_concise_1997;@volk_infant_2013].  -->
+<!-- Economic theories of fertility that consider children as 'investment' for old-age support also rely on notions of child survival [@preston1978effects]. -->
+Offspring survival is usually studied alongside fertility decline in the context of the demographic transition [@gortfelder_survival_2020], but it also matters for post-transitional societies. 
+Lower fertility and higer life expectancy means that fewer children are expected to provide key emotional, social, and financial transfers to aging parents for longer periods of time.
+With increasing periods of generational overlap, individuals find themselves 'sandwiched' between aging parents and young children requiring their simultaneous attention and care [@daatland_norwegian_2010].
+<!-- Losing an only child at old age is expected to become a more common experience in the future [@alburez_gutierrez_womens_2019]. -->
+In the context of global population aging, elderly parents without access to formal social security and pension systems are particularly reliant on these transfers [@smith-greenaway_maternal_2020].
+Given the known psychological, health, and social consequences of child loss for bereaved parents and families [@Hendrickson2009;@lee_death_2014], it is surprising that the demographic processes that shape parental bereavement remain very poorly understood.
 
-Kin count estimation has a long and distinguished pedigree in mathematical demography, starting with the work of @lotka1931orphanhood on modelling orphanhood in theoretical populations across demographic regimes.
-To the best of our knowledge, @brass_derivation_1953 first proposed a formula to represent child survival over age.
-The notion of estimating the expected number of living daughters in a stable population was generalized by @goodman1974 for other kin relations (granddaughters, cousins, etc.).
-The "counting method" approach was further popularized in @KeyfitzMath.
-@bongaarts_projection_1987 used a similar approach to estimate descendants in his 'Family Status Model'.
+The Covid-19 pandemic has underscored the need to understand how sudden changes in mortality affect the availability of kin. 
+The elderly depend on kin support, especially during periods of lockdown and in the absence of governmental support mechanisms.
+Elderly mothers may be at a higher risk of losing the key support provided by their adult children given the known age-gradient in the Covid-19 case fatality rate.
+How do changes in mortality affect the availability of offspring over age from the point of view of a prospective mother? 
+<!-- How can we characterize the timing of offspring mortality from a mother's perspective in a formal demographic framework?  -->
+How do these processes play out in a heterogeneous population?
 
-Child survival - and it's counterpart, child death - gain new significance in the wake of the fertility transition.
-Depressed fertility means that fewer children are tasked with providing key emotional, social, and financial transfers to aging parents for ever increasing periods of time.
-With increasing periods of generational overlap, individuals find themselves 'sandwiched' between aging parents and young children requiring their simultaneous attention and care [@daatland_norwegian_2010].
-In the context of global population aging, elderly parents without access to formal social security and pension systems are particularly reliant on these transfers to make ends meet [@smith-greenaway_maternal_2020].
-Losing an only child may become a more common experience in the context of global fertility decline, something particularly worrying for parents suddenly rendered childless.
-Given the known psychological, health, and social consequences of child loss for bereaved parents and families [@Hendrickson2009;@lee_death_2014], it is surprising that the demographic processes that shape parental bereavement remain very poorly understood.
-Recent development have underscored the need to understand how sudden changes in mortality affect the availability of kin. 
-In the context of the global pandemic caused by the Covid-19 disease, elderly people depend on support provided by their relatives, especially during periods of lockdown and in the absence of governmental support mechanisms.
-<!-- Maybe here are repeated --> Elderly parents may be at a higher risk of losing the key support provided by their adult children given the known age-gradient in the Covid-19 case fatality rate. 
-How do changes in mortality affect the availability of offspring over age from the point of view of a prospective mother? 
-How can we characterize the timing of offspring mortality as experienced by a mother from a formal demographic perspective? 
-How this would work in an heterogeneity population?
+In this paper, we aim to formalize the relationship between population-level changes in mortality rates (additive or multiplicative) and changes in the lived experience of kin death.
+<!-- Ivan: si decimos esto, quiza deberia ser el enfoque principale del paper tener age-specific changes?  -->
+This is key for understanding the impact of age-specific excess mortality on the resilience or otherwise of kinship networks for the vulnerable elderly population.
+This paper uncovers formal relationships and provides numerical applications using data from Latin America in the second half of the twentieth century using period measures in a stable context.^[See the [Applications](#Applications) section for more details on the source, calculation procedures and main numerical results.]
+<!-- A more sophisticated model using changing age-specific cohort rates in the subscripts has been proposed by @alburez_gutierrez_womens_2019. -->
+We continue a long tradition in mathematical demography, starting with the work of @lotka1931orphanhood on modelling orphanhood in theoretical populations, @brass_derivation_1953 on child survival, @goodman1974 on kin survival in stable populations, and recently, @caswell_formal_2019 kinship matrices.
+<!-- The 'counting method' approach was further popularized in @KeyfitzMath. @bongaarts_projection_1987 used a similar approach to estimate descendants in his 'Family Status Model'. -->
 
-In this paper we aim to formalize the relationship population-level changes in mortality rates and perceived change in the lived experience of child death from the perspective of a mother.
-Our approach is key for understanding the impact of age-specific excess mortality on the resilience or otherwise of kinship networks for the vulnerable elderly population.
-We consider the case when these changes are additive and mutliplicative.
-In this paper, we uncover formal relationships to answer these questions and provide numerical applications using data from countries in Latin America during the second half of the twentieth century.
-See the [Applications](#Applications) section for more details on the source, calculation procedures and main numerical results.
-We used period measures, in an stable context, so the objetive is give moment measures.
-A more sophisticated model using the appropriate age-specific cohort rates in the subscripts has been proposed by @alburez_gutierrez_womens_2019.
+<!-- starting with the work of @lotka1931orphanhood on modelling orphanhood in theoretical populations across demographic regimes.  -->
+<!-- To the best of our knowledge, @brass_derivation_1953 first proposed a formula to represent child survival over age.  -->
+<!-- The notion of estimating the expected number of living daughters in a stable population was generalized by @goodman1974 for other kin relations (granddaughters, cousins, etc.).  -->
+<!-- The 'counting method' approach was further popularized in @KeyfitzMath. @bongaarts_projection_1987 used a similar approach to estimate descendants in his 'Family Status Model'. -->
 
@@
 
-As an initial step, we seek an intuitive understanding of the effect of changing mortality on child survival. 
+As an initial step, we provide an intuitive understanding of the effect of changing mortality on child survival. 
 Building on work by @KeyfitzMath for the probability of a living mother, we can show which features of the child survival function explain the expected number of children. 
@@
 
-where the fertility pattern by age is concentrated around $\kappa$ and the accumulated fertility (or gross reproduction rate in our female-dominant scenario) is $F_a =  \int_{\alpha}^a{m_{x}dx}$.
+where the fertility pattern by age is concentrated around $\kappa$ and the accumulated fertility (or gross reproduction rate in our one-sex population) is $F_a =  \int_{\alpha}^a{m_{x}dx}$.
 The second Taylor´s term is null because $\int_{\alpha}^a{x \, m_{x}dx} = \kappa \, F_a$.
@@
 We find that, seen from the perspective of a mother, child survival mainly depends on the cummulative fertility function^[Different from the findings of @KeyfitzMath of having living ancestors.] and the survival from birth to $a-\kappa$.
-Also that the approximation is affected (negativally) by how disperse is the fertility by age (variance $\sigma^2$) and the suvival curvature around (with negative sign), in an age range where $l_x$ is tipycally very flat (range 20-40 years old) in transitioned populations^[This approximation is useful also for get a dimension changes in the fertility average age \kappa: taking logs in $F_a \, l_{a-\kappa}$ and deriving we get $\frac{\Delta CS_a}{CS_a} \approx - \mu_{a-\kappa} \Delta\kappa$].
+The approximation is affected (negativally) by the disperssion of fertility by age (variance $\sigma^2$).
+The same is true for the curvature of the survival surve, in an age range where $l_x$ is tipycally very flat (range 20-40 years old) in transitioned populations^[This approximation is useful also for get a dimension changes in the fertility average age \kappa: taking logs in $F_a \, l_{a-\kappa}$ and deriving we get $\frac{\Delta CS_a}{CS_a} \approx - \mu_{a-\kappa} \Delta\kappa$].
 
-### Mortality Changes
+### Absolute Mortality Change
 
-We now consider the consequences of an absolute change $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk.
+We now consider the consequences of an absolute, or additive, change $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk.
 For now, let us consider $m_{x,\delta}=m_{x}+\delta$ (@wrycza2012) and $l_{a-x}^\delta = e^{-\int_{0}^{a-x}{(\mu_t+\delta})dt}$:
 
-$$CS_{a}^\delta = \int_{\alpha}^{a}m_{x} l_{a-x} e^{-\delta (a-x)} dx$$
+$$CS_{a}^\delta = \int_{\alpha}^{a}m_{x} l_{a-x} e^{-\delta (a-x)} dx.$$
 
-We get the derivative of $dCS_{(a)}^\delta / d\delta$ evaluated near zero (@KeyfitzMath) to find the effects of adding $\delta$ to hazard rates:
+We get the derivative of $dCS_{(a)}^\delta / d\delta$ evaluated near zero (@KeyfitzMath) to find the effects of adding $\delta$ to the hazard rates:
 
@@
 
-Dividing both sides by $CS_{a}$ in an discrete approximation we get:
+Dividing both sides by $CS_{a}$ we obtain a discrete approximation:
 
-$$\frac{\Delta CS_{a}}{CS_{a}} \approx -(a - k_a) \Delta\delta$$
+$$\frac{\Delta CS_{a}}{CS_{a}} \approx -(a - k_a) \Delta\delta.$$
 
@@
 
-That an equal change in mortality could happens at all ages is unlikelihood.
-This measure could be much related a change in first ages (child and youth) than adult or ones.
-As a corollary, if the change only happens in the age range $\left[0;1\right)$, infant mortality, we can inspect the effect starting with splitting the integral (considering those born last year and not, and of course for $a>1$):
+However, it is unlikely for mortality to change at the same rate at all ages. 
+In the course of the demographic transition, for example, we owuld expect larger changes in young-age mortality.
+As a corollary, we consider an additive change in infant mortality in the age range $\left[0;1\right)$. 
+For this, we split the integral to considering those born in the previous year for $a>1$:
 
@@
 $$CS_{a}^{\delta_0} =-\int_{0}^{a-1} m_{x} l_{a-x} dx - a \int_{a-1}^{a} {m_{x} l_{a-x} dx} + \int_{a-1}^{a} {x \, m_{x} l_{a-x} dx}$$
-If we name $CS_{a-1,a}=\int_{a-1}^{a} {m_{x} l_{a-x} dx}$, we can express the first term in the right as $CS_a - CS_{a-1,a}$, and finally divide by $CS_a$: 
+Since $CS_{a-1,a}=\int_{a-1}^{a} {m_{x} l_{a-x} dx}$, we can express the first term in the right as $CS_a - CS_{a-1,a}$, and finally divide by $CS_a$: 
 $$ = - CS_a + CS_{a-1,a} (1-a+\kappa_{a-1,a})\\
@@
 
-This means that an absolute change in infant hazard rate affects proportionally for all the age range extension (because of all the cohorts aged more than 1), except for the mean time exposure for those who born and survive in the last year $1-(a-k_{a-1,a})$, times $\frac{CS_{a-1,a}}{CS_a}$ the portion of cummulative alive daughters for the same period, weighting how important was this in total successful experience (in terms of alive descendants).
+<!-- Este parrafo necesita mas claridad -->
+An absolute change in infant hazard rate affects proportionally for all the age range extension (because of all the cohorts aged more than 1), except for the mean time exposure for those who born and survive in the last year $1-(a-k_{a-1,a})$, times $\frac{CS_{a-1,a}}{CS_a}$ the portion of cummulative alive daughters for the same period, weighting how important was this in total successful experience (in terms of alive descendants).
+
+### Proportional Mortality Change
 
 We now consider the consequences of a proportional change in mortality on child survival.
-Given a proportional change in mortality $\mu_{x,\delta}=\mu_{x}(1+\delta)$ so that $l_{a-x}^\delta = e^{-\int_{0}^{a-x}{\mu_t(1+\delta})dt}=(l_{a-x})^{(1+\delta)}$, it follows that:
+Given a proportional change in mortality $\mu_{x,\delta}=\mu_{x}(1+\delta)$ so that $l_{a-x}^\delta = e^{-\int_{0}^{a-x}{\mu_t(1+\delta})dt}=(l_{a-x})^{(1+\delta)}$:
 
@@
 
-The last integral is $CS_{a-t}$.
-Dividing by $CS_a$ and multiplying by $\delta$ get one useful expression.
-This is the negative of cumulative hazard $H_a$ but considering a positive factor $\frac{CS_{a-t}}{CS_a}$ that takes in account the relative amount of surviving descendants that would be lost at each child age at risk.
+Considering that the last integral is equal to $CS_{a-t}$, we divide by $CS_a$ and multiplying by $\delta$ we obtain:
 
@@
 
+<!-- This needs more clarity -->
+This is the negative of cumulative hazard $H_a$ but considering a positive factor $\frac{CS_{a-t}}{CS_a}$ that takes in account the relative number of surviving descendants that would be lost at each child age at risk.
 Grasping this relationship intuitively is more difficult given the interaction of birth and mortality rates (@KeyfitzMath). 
 This factor has a S-shape because $l_x$ curvature and fertility accumulation, that gives more weight to first ages, generally less than 1 except for mother ages close to age valuation *a* with complete reproductive life and high infant mortality (no additional children and high risk for one age additional of survive), typically pretransitional countries with a near 50.
-A numerical evaluation of both analytical expressions (absolute an relative) are done in [Applications](#Applications) section.
+We provide a numerical evaluation of both analytical expressions (absolute an relative) in the [Applications](#Applications) section.
 
@@
 
-We motivate this paper with an empirical example using fertility and mortality rates from the 2019 Revision of the UN World Population Prospects (UN WPP) for the Latin American Region (@WPP19).
+We provide empirical estimates using fertility and mortality rates from the 2019 Revision of the UN World Population Prospects (UN WPP) for the Latin American Region (@WPP19).
 We smoothed female $l_x$ in quinquennial ages, using cubic-splines constrained to monotonic decrease, taking $L_0$ and $T_{100}$ from raw life tables as inputs for year-person calculations.
@@
 
-```{r plot_ITL_MAL, fig.cap = "a) Intensity Tome Lost of  women aged 30, 40 and 50.
+```{r plot_ITL_MAL, fig.cap = "a) Intensity Time Lost of  women aged 30, 40 and 50.
 b) Mean age at Lost for women aged 40.

### 2020-09-21 R/funs.R
borradores

@@ -4,9 +4,9 @@ library(tidyverse)
 library(gridExtra)
 source("R/QOSplit.R")
 
-# main_funtion
+# main functions ----------------------------------------------------------
 
-# age survivor spline
+# age survivor split spline
 lx_fun <- function(data, age){
   splinefun(data$x, data$lx, method = "monoH.FC")(age)
 }
@@ -23,7 +23,7 @@ CS_results <- function(LA, country, period, a){   # LA is the data base
                            Lx = ifelse(x==0, Lx_n, # assume uniform deaths within interval
                                        ifelse(x==100, ex, 
                                               lx-dx/2)),
-                           Lx = as.numeric(Lx), # in case
+                           Lx = as.numeric(Lx), # just in case
                            ex = rev(cumsum(rev(Lx)))/lx)
   
   # calculate: 
@@ -32,7 +32,7 @@ CS_results <- function(LA, country, period, a){   # LA is the data base
   CS = sum(unlist(sapply(1:a, function(i) country$asfr[i]*country$lx[a+2-i]/100000)), na.rm = T)
   Fa = sum(country$asfr[1:a], na.rm = T)
   CS_prob = CS/Fa
-  k = sum(country$asfr[1:a]*0:(a-1), na.rm = T)/Fa # must be kappa but...
+  k = sum(country$asfr[1:a]*0:(a-1), na.rm = T)/Fa # should be noted as kappa but...
   CS_aprox = Fa * country$lx[a+1-k]/100000
   error_aprox = (CS-CS_aprox)/CS*100 # relative error approx
   MTSL = 0
@@ -68,6 +68,9 @@ CS_results <- function(LA, country, period, a){   # LA is the data base
   return(data.frame(CS,Fa,CS_prob,k,CS_aprox,error_aprox,MTSL,ITL,MAL))
 }
 
+
+# get data and smooth -----------------------------------------------------
+
 # Get wpp locations
 data(UNlocations)
 
@@ -90,7 +93,7 @@ asfr_paises_LA <- percentASFR %>% gather(Period, asfr, -name, -country_code, -ag
                   select(-age) %>% 
                   filter(country_code %in% paises_LA_seleccion)
 
-# lt must downloaded manually (maybe too heavy for CRAN)
+# lt data must be downloaded manually (maybe too heavy for CRAN)
 lt_paises_LA <- readxl::read_xlsx("Data/WPP2019_MORT_F17_3_ABRIDGED_LIFE_TABLE_FEMALE.xlsx", 
[…]

### 2020-09-21 try/Demographic Research 4495 Matlab files/kin_program.m

@@ -1,0 +1,131 @@
+%program to calculate kinship structures
+% this program:
+% reads in qx and fx files
+% constructs U and F matrices
+% calls kinship function
+
+%load qx and fx files from csv files
+%qx = probability of death in age class x
+%fx = age-specific fertility in age class x
+
+qxyears=dlmread('qx_years.csv');
+fxyears=dlmread('fx_years.csv');
+
+%pick a year; here, choose year 1, which is 1947
+qx=qxyears(:,1);
+fx=fxyears(:,1);
+
+%divide fertility by 2 to get female offspring
+fx=fx/2;
+
+%construct U matrix and F matrix
+px=1-qx(:,1);
+
+%remove last age class (qx=1, px=0)
+
+U=diag(px(1:end-1),-1);
+F=zeros(size(U));
+F(1,:)=(px.*fx(:,1))'; %post-breeding birth pulse model
+
+%call kinship program
+kin_out=kinship_function_DR(U,F);
+
+%names of kin
+kinname={'daughter',...
+    'granddaughter', ...
+    'great-granddaughter', ...
+    'mother', ...
+    'grandmother', ...
+    'great-grandmother', ...
+    'older sisters', ...
+    'younger sisters', ...
+    'nieces (older)', ...
+    'nieces (younger)', ...
+    'aunts (older)', ...
+    'aunts (younger)', ...
+    'cousins (older)', ...
+    'cousins (younger)'};
+
+
+%names for combined kin (combining age; e.g., older and younger sisters
+%become sisters)
+kinname2={'daughters',...
+    'granddaughters', ...
+    'great-granddaughters', ...
+    'mothers', ...
+    'grandmothers', ...
+    'great-grandmothers', ...
+    'sisters', ...
+    'nieces', ...
+    'aunts', ...
+    'cousins'};
+
+%following code gives examples of calculation of various aspects of kin,
+%from the output structure arrays
+
+%for purposes of illustration, figures are created to plot the results for
+%sisters (combining older and younger sisters), which are kin type 7.
+
+om=kin_out.om;  %maximum age class
+
+%age distributions of kin
+% choose an age for Focal
+ageFocal = 30;
+
+for ikin=1:10 
+    y_age(:,ikin) = (kin_out.allkin2(1:om,ageFocal,ikin))';
+end %for ikin
+
[…]

### 2020-09-21 try/Demographic Research 4495 Matlab files/kinship_function_DR.m

@@ -1,0 +1,208 @@
+function out=kinship_function_DR(U,F)
+%inputs
+% U=subdiagonal survival matrix
+% F = fertility matrix
+% U and F refer to living ages
+% Utilde and Ftilde, including ages at death, are constructed here,
+%inside the function
+
+%size of U
+[om,om]=size(U);
+%identity matrix
+Iom=eye(om);
+
+%create model with absorbing states for age at death
+
+Mort=diag(1-sum(U));
+
+%construct survival matrix
+
+%must specify how deaths should be counted:
+%   accumulated deaths up to age x of Focal or ...
+%   experienced deaths at age x of Focal
+
+% accumulated deaths 
+type_death='cum';
+Utilde=[U zeros(om,om);
+    Mort Iom];
+
+% %experienced deaths 
+%    type_death='exp'
+%     Utilde=[U zeros(om,om);
+%         Mort zeros(om,om)];
+
+%construct fertility matrix Ftilde
+Ftilde=[F zeros(om,om);
+    zeros(om,2*om)];
+
+%identity matrix for Utilde
+I2om=eye(2*om);
+
+%find distribution of age of mothers in stable pop
+
+Amat=U+F;
+[w,d]=eig(Amat);
+d=diag(d);
+pick=find(d==max(d));
+w=w(:,pick);
+w=w/sum(w);
+
+lambda=d(pick);
+
+clear d
+
+%distribution of ages of mothers of children
+pi=F(1,:)'.*w;
+
+%append zeros for the dead
+pi=[pi;zeros(om,1)];
+%normalize the vector
+pi=pi/sum(pi);
+
+%frequently used zero vector for initial condition
+zvec=zeros(2*om,1);
+
+%frequently used om-1 limit for iterations
+omz=om-1;
+
+%following code calculates age distribution, including the dead
+%for each type of kin, and stores these as columns of an array
+% e.g., a(x) = daughters at age x; A(:,x) contains a(x)
+
+% a: daughters of focal
[…]

### 2020-09-21 try/kins.R

@@ -1,0 +1,78 @@
+library(tidyverse)
+
+get_kins <- function(U_list, F_list, w=100, pop_mother){
+  # x ego´s age at time t + n
+  # U_list survive matrix
+  # F_list fertility matrix
+  
+  asfrt = LA %>% filter(name=="Argentina", Period=="1950-1955") %>%
+              mutate(asfr = ifelse(is.na(asfr),0,asfr)) %>% pull(asfr)
+  Pt = LA %>% filter(name=="Argentina", Period=="1950-1955") %>% 
+              mutate(px = 1-qx) %>% pull(px)
+  ages = length(Pt)
+  Ut = Ft = matrix(0, nrow=ages, ncol=ages)
+  Ut[row(Ut)-1 == col(Ut)] <- Pt[-101]
+  Ut[ages,ages]=Pt[101]
+  Ft[1,] = asfrt
+  
+  kins = data.frame(x=0, a=a, b=b, c=c, n=n)
+  
+  a = b = c = n = rep(0,ages)
+  # vectors of kins by age
+  d = runif(ages)
+  d[c(1:15,50:ages)] = 0
+  d = d/sum(d)
+  g = d
+  m = d
+  
+  e = rep(0,ages)
+  
+  age_ego = 70
+  for(x in 2:age_ego){
+    
+    # daughters
+    e[x] = 1 
+    a = Ut %*% a + Ft %*% e
+    
+    # granddaughters
+    b = Ut %*% b + Ft %*% a
+    
+    # mother
+    d = Ut %*% d
+    
+    # younger sisters
+    n = Ut %*% n + Ft %*% d
+    
+    # grandmother
+    g = Ut %*% g
+    
+    # older sisters
+    m = Ut %*% m
+    
+    e = rep(0,ages)
+    }
+    
+  plot(a,ylim=c(0,.1))
+  lines(b)  
+  plot(n)
+  
+    # get results
+    kins = rbind(kins, data.frame(x=x, a=a, b=b, c=c, d=d, g=g, m=m, n=n))
+  }
+  
+  
+  
+
+
+
+
+
+
+F_m = matrix(c(.3,.3,.3,
+               0,0,0,
+               0,0,0),nrow = 3,byrow = T)
+
+
+ex = c(0,1,0)
+
+F_m%*%t(ex)

### 2020-12-28 Old/paper.Rmd
code para reproducible

--- 2020-08-15 paper.Rmd
@@
 output:
-  pdf_document: default
+  pdf_document: 
+    keep_tex: yes
   html_document:
@@
 
-The question of child survival, particularly the survival of daughters for mothers, sits at the very center of demographic theory.
-This is evidenced by the omnipresence of the 'net reproductive rate' in accounts of human and non-human populations. 
-Historical demographers draw liberally on assumptions about kin availability and individual's exposure to offspring mortality to explain fertility decisions, especially around the demographic transition.
-These assumptions are often untested given the data sparsity [@livi_bacci_concise_1997;@volk_infant_2013]. 
-Economic theories of fertility that consider children as 'investment' for old-age support also rely on notions of child survival [@preston1978effects].
+The question of child survival for mothers sits at the very center of demographic theory.
+Offspring survival is usually studied alongside fertility decline in the context of the demographic transition, but it also matters for post-transitional societies. 
+Lower fertility and higer life expectancy means that fewer children are expected to provide key emotional, social, and financial transfers to aging parents for longer periods of time.
+With increasing periods of generational overlap, individuals find themselves 'sandwiched' between aging parents and young children requiring their simultaneous attention and care [@daatland_norwegian_2010].
+In the context of global population aging, elderly parents without access to formal social security and pension systems are particularly reliant on these transfers [@smith-greenaway_maternal_2020].
+Given the known psychological, health, and social consequences of child loss for bereaved parents and families [@Hendrickson2009;@lee_death_2014], it is surprising that the demographic processes that shape parental bereavement remain very poorly understood.
 
-Kin count estimation has a long and distinguished pedigree in mathematical demography, starting with the work of @lotka1931orphanhood on modelling orphanhood in theoretical populations across demographic regimes.
+The Covid-19 pandemic has underscored the need to understand how sudden changes in mortality affect the availability of kin. 
+The elderly depend on kin support, especially during periods of lockdown and in the absence of governmental support mechanisms.
+Elderly mothers may be at a higher risk of losing the key support provided by their adult children given the known age-gradient in the Covid-19 case fatality rate.
+How do changes in mortality affect the availability of offspring over age from the point of view of a prospective mother? 
+How do these processes play out in a heterogeneous population?
+
+In this paper, we aim to formalize the relationship between population-level changes in mortality rates (additive or multiplicative) and changes in the lived experience of kin death.
+<!-- Ivan: si decimos esto, quiza deberia ser el enfoque principale del paper tener age-specific changes?  -->
+This is key for understanding the impact of age-specific excess mortality on the resilience or otherwise of kinship networks for the vulnerable elderly population.
+This paper uncovers formal relationships and provides numerical applications using data from Latin America in the second half of the twentieth century using period measures in a stable context.^[See the [Applications](#Applications) section for more details on the source, calculation procedures and main numerical results.]
+We continue a long tradition in mathematical demography, starting with the work of @lotka1931orphanhood on modelling orphanhood in theoretical populations, @brass_derivation_1953 on child survival, @goodman1974 on kin survival in stable populations, and recently, @caswell_formal_2019 kinship matrices.
+
+<!-- The question of child survival, particularly the survival of daughters for mothers, sits at the very center of demographic theory. -->
+<!-- This is evidenced by the omnipresence of the 'net reproductive rate' in accounts of human and non-human populations.  -->
+<!-- Historical demographers draw liberally on assumptions about kin availability and individual's exposure to offspring mortality to explain fertility decisions, especially around the demographic transition.
+-->
+<!-- These assumptions are often untested given the data sparsity [@livi_bacci_concise_1997;@volk_infant_2013].  -->
+<!-- Economic theories of fertility that consider children as 'investment' for old-age support also rely on notions of child survival [@preston1978effects]. -->
+
+<!-- Kin count estimation has a long and distinguished pedigree in mathematical demography, starting with the work of @lotka1931orphanhood on modelling orphanhood in theoretical populations across demographic regimes.
 To the best of our knowledge, @brass_derivation_1953 first proposed a formula to represent child survival over age.
@@
 Recently @Caswell2019 reformulated these relationships in a matrix framework, extending the potential results to variance estimation, multistate models, and others.
+-->
 
-Child survival - and it's counterpart, child death - gain new significance in the wake of the fertility transition.
-Depressed fertility means that fewer children are tasked with providing key emotional, social, and financial transfers to aging parents for ever increasing periods of time.
-With increasing periods of generational overlap, individuals find themselves 'sandwiched' between aging parents and young children requiring their simultaneous attention and care [@daatland_norwegian_2010].
-In the context of global population aging, elderly parents without access to formal social security and pension systems are particularly reliant on these transfers to make ends meet [@smith-greenaway_maternal_2020].
-Losing an only child may become a more common experience in the context of global fertility decline, something particularly worrying for parents suddenly rendered childless.
-Given the known psychological, health, and social consequences of child loss for bereaved parents and families [@Hendrickson2009;@lee_death_2014], it is surprising that the demographic processes that shape parental bereavement remain very poorly understood.
-Recent development have underscored the need to understand how sudden changes in mortality affect the availability of kin. 
-In the context of the global pandemic caused by the Covid-19 disease, elderly people depend on support provided by their relatives, especially during periods of lockdown and in the absence of governmental support mechanisms.<!-- Maybe here are repeated --> Elderly parents may be at a higher risk of losing the key support provided by their adult children given the known age-gradient in the Covid-19 case fatality rate.
+<!-- Child survival - and it's counterpart, child death - gain new significance in the wake of the fertility transition. -->
+<!-- Depressed fertility means that fewer children are tasked with providing key emotional, social, and financial transfers to aging parents for ever increasing periods of time. -->
+<!-- With increasing periods of generational overlap, individuals find themselves 'sandwiched' between aging parents and young children requiring their simultaneous attention and care [@daatland_norwegian_2010].
+-->
+<!-- In the context of global population aging, elderly parents without access to formal social security and pension systems are particularly reliant on these transfers to make ends meet [@smith-greenaway_maternal_2020].
+-->
+<!-- Losing an only child may become a more common experience in the context of global fertility decline, something particularly worrying for parents suddenly rendered childless. -->
+<!-- Given the known psychological, health, and social consequences of child loss for bereaved parents and families [@Hendrickson2009;@lee_death_2014], it is surprising that the demographic processes that shape parental bereavement remain very poorly understood.
+-->
+<!-- Recent development have underscored the need to understand how sudden changes in mortality affect the availability of kin.  -->
+<!-- In the context of the global pandemic caused by the Covid-19 disease, elderly people depend on support provided by their relatives, especially during periods of lockdown and in the absence of governmental support mechanisms.
+-->
+<!-- Elderly parents may be at a higher risk of losing the key support provided by their adult children given the known age-gradient in the Covid-19 case fatality rate.  -->
 
-How do changes in mortality affect the availability of offspring over age from the point of view of a prospective mother?
+<!-- How do changes in mortality affect the availability of offspring over age from the point of view of a prospective mother?
 How can we characterize the timing of offspring mortality as experienced by a mother from a formal demographic perspective?
 How this would work in an heterogeneity population?
-In the remainder of the paper, we aim to formalize the relationship population-level changes in mortality rates and perceived change in the lived experience of child death from the perspective of a mother.
+-->
+<!-- In the remainder of the paper, we aim to formalize the relationship population-level changes in mortality rates and perceived change in the lived experience of child death from the perspective of a mother.
 Our approach is relevant for understanding the impact of overall and age-specific excess mortality on the resilience or otherwise of kinship networks for the vulnerable elderly population.
-We consider the case when these changes are additive and mutliplicative.
+-->
+<!-- We consider the case when these changes are additive and mutliplicative.
 In this paper, we uncover formal relationships to answer these questions and provide numerical applications using data from countries in Latin America during the second half of the twentieth century.
@@
 A more sophisticated model using the appropriate age-specific cohort rates in the subscripts has been proposed by @alburez_gutierrez_womens_2019.
+-->
 
@@
 
-That an equal change in mortality could happens at all ages is unlikelihood.
-This measure could be much related a change in first ages (child and youth) or adolescent ones.
-If the change only happens in the age range $\left[0;1\right)$, infant mortality, we can inspect the effect in child survival $CS_{a}^{\delta_0}$ starting with splitting the integral between those daughters who were exposed all the range and those who do not (now generalizing for $\alpha$ positive and $a>1$ for simplicity):
+However, it is unlikely for mortality to change at the same rate at all ages. 
+In the course of the demographic transition, for example, we would expect larger changes in young-age mortality.
+<!-- This measure could be much related a change in first ages (child and youth) or adolescent ones.  -->
+Consider a change in infant mortality in the age range $\left[0;1\right)$.
+We can inspect the effect in child survival $CS_{a}^{\delta_0}$ by splitting the integral between those daughters who were exposed to the mortality change and those who were not (now generalizing for $\alpha$ positive and $a>1$ for simplicity):
 
@@
 $$\frac{dCS^{\delta_0}}{d\delta_0} =-\int_{0}^{a-1} m_{x} l_{a-x} dx - a \int_{a-1}^{a} {m_{x} l_{a-x} dx} + \int_{a-1}^{a} {x \, m_{x} l_{a-x} dx}$$
-If we name $CS_{a-1,a}=\int_{a-1}^{a} {m_{x} l_{a-x} dx}$, we can express the first term in the right as $CS_a - CS_{a-1,a}$, and recognizing that last term could be expressed as  $\kappa_{a-1,a} CS_{a-1,a}$, we get:
+Given that $CS_{a-1,a}=\int_{a-1}^{a} {m_{x} l_{a-x} dx}$, we can express the first term in the right as $CS_a - CS_{a-1,a}$. 
+The righ-most term could also be expressed as  $\kappa_{a-1,a} CS_{a-1,a}$, so that:
 
 $$ \frac{dCS^{\delta_0}}{d\delta_0} = - CS_a + CS_{a-1,a} (1-a+\kappa_{a-1,a})$$
-Where the factor $1-(a-\kappa_{a-1,a})$, which lies between 0 and 1, is equal to the average portion of time spent with living daughters durings age *a*.
-Assuming uniform distribution of those births, discretizing the change and dividing by $CS_a$ we get the final expression for change in infant mortality:
+The factor $1-(a-\kappa_{a-1,a})$, which lies between 0 and 1, is equal to the average portion of time spent with living daughters durings age *a*.
+Assuming a uniform distribution of those births, discretizing the change and dividing by $CS_a$ we get the final expression for change in infant mortality:
 
@@
 
+<!-- Este parrafo no esta muy claro... -->
 This means that an absolute change in infant hazard rate affects proportionally for all the age range extension (cohorts aged more than 1), except for "half" the portion $\frac{CS_{a-1,a}}{CS_a}$ on how important was this las age in total "successful" experience (in terms of alive descendants).
@@
 
-The last integral is $CS_{a-t}$.
-Dividing by $CS_a$ and multiplying by $\delta$ get one useful expression.
-This is the negative of cumulative hazard $H_a$ but considering a positive factor $\frac{CS_{a-t}}{CS_a}$ that takes in account the relative amount of surviving descendants that would be lost at each child age at risk, that gives more weight to first ages because all the parity was exposed in that age.
+Considering that the last integral is equal to $CS_{a-t}$, we divide by $CS_a$ and multiply by $\delta$ to obtain:
 
 $$\frac{\Delta CS_{a}}{CS_a} \approx - \left[\int_{0}^{a}{\mu_x  \, \frac{CS_{a-x}}{CS_a} \,dx \,}\right] \Delta \delta$$
+
+This is the negative of cumulative hazard $H_a$ but considering a positive factor $\frac{CS_{a-t}}{CS_a}$ that takes in account the relative amount of surviving descendants that would be lost at each child age at risk, that gives more weight to first ages because all the parity was exposed in that age.
 
@@
 In advanced ages when fertility is complete, the difference is only explained by the surviving part, that is why this factor is not a strictly increasing function but for mother ages close to $\beta$ (last theoretical fecundity age).
-A numerical evaluation of both analytical expressions (absolute an relative) and its factor $\frac{CS_{a-t}}{CS_a}$ are done in [Applications](#Applications) section.
+We provide a numerical evaluation of both analytical expressions (absolute an relative) in the [Applications](#Applications) section.
+
+$$
+
+CS_{a}^{\delta_0} = \int_{0}^{a} m_{x} {e{^{-\int_{1}^{a-x}\mu_t \,dt-\int_{0}^{1}\mu_t\,(1+\delta)\,dt}}} dx \\
+
+CS_{a}^{\delta_0} = \int_{0}^{a} m_{x} {e{^{-\int_{0}^{a-x}\mu_t \,dt - \delta \int_{0}^{1}\mu_t\,dt}}} dx\\
+
+CS_{a}^{\delta_0} = \int_{0}^{a} m_{x} {e{^{-\int_{0}^{a-x}\mu_t \,dt - \delta H_{0,1}}}} dx\\
+
+CS_{a}^{\delta_0} = \int_{0}^{a} m_{x} {e^{-\int_{0}^{a-x}\mu_t \,dt} e^{- \delta H_{0,1}}} dx\\
+
+CS_{a}^{\delta_0} = e^{- \delta H_{0,1}} CS_{a}\\
+
+CS_{a}^{\delta_0} = l_1^\delta CS_{a}
+
+$$
 

### 2020-12-28 R/QOSplit.R

@@ -1,61 +1,0 @@
-## Michalski AI, Grigoriev P, Gorlishchev VP  ##
-
-## MPIDR Technical Report ##
-
-## 22.11.2017 ##
-
-## Splitting Fx (age-specific fertility rates) 
-## using quadratic optimization (QO) method  (HFD data) ##
-
-
-## Arguments of 'QOSplit' function ## 
-## Fx - vector of aggregated ASFRs to be split into single ages ##
-## L - vector of ages  (lower limits. e.g 10-15 is 10, 20-24 is 20, and so on) ##
-## AgeInt - vector of age intervals ##
-QOSplit<-function(Fx,L,AgeInt){
-  
-  ## (0) Closing left (-99) and right (99) open intervals, if any ##
-  n<-length(AgeInt)
-  if (AgeInt[1]==-99) {L[1]<-L[1]-AgeInt[2]+1;AgeInt[1]<-AgeInt[2] } 
-  if (AgeInt[n]==99) {AgeInt[n]<-AgeInt[n-1]}
-  
-  h<-AgeInt			# number of 1-years age groups in the 5-years age groups
-  N<-sum(h)			# numberof columns  in G
-  
-  ## (1) Getting G matrix
-  G<-matrix(0,nrow=n,ncol=N)
-  G[1,1:h[1]]<-1/h[1]			
-  j1<-h[1]+1				
-   for (i in 2:n){ 
-    j2<-j1+h[i]-1	
-    G[i,j1:j2]<-1/h[i]			
-    j1<-j2+1  } 
-  
-  # (2) Quadratic minimization
-    
-  # Matrix for the 2-nd order derivatives
-  F<-matrix(0,nrow=N-2,ncol=N)
-  for (i in 1:(N-2)){  F[i,i]<-1; F[i,i+1]<--2; F[i,i+2]<-1;}
-  G<-G[,-N]; G<-G[,-1];
-  F<-F[,-N]; F<-F[,-1];
-  Dmat=t(F)%*%F			# matrix for Quadratic form	
-  dvec<-rep(0,N-2)
-  Amat<-t(rbind(G,diag(N-2)))	# constrains matrix
-  meq<-n					# number of equations
-  
-  bvec<-c(Fx,dvec)		# vector of birth
-  aux<-solve.QP(Dmat, dvec, Amat, bvec, meq=meq)
-  aux<-c(0,aux$solution,0)	#  the first and the last values are assumed to have no fertility!
[…]

### 2020-12-28 R/data.R

@@ -1,0 +1,67 @@
+get_data <- function(source = c("WPP", "HMD_HFD"),
+                     country = "Sweden",
+                     years = 1950:2020){
+  
+  
+  lx_fun <- function(data, age){
+    splinefun(data$x, data$lx, method = "monoH.FC")(age)
+  }
+  
+  ################# WPP
+  library(wpp2019)
+  # get fertility data
+  data(tfr)
+  data(tfrprojMed)
+  data(percentASFR)
+  
+  # tfr
+  tfr_country <- tfr %>% bind_cols(tfrprojMed %>% select(-1,-2)) %>% 
+                    select(-last.observed) %>% 
+                    gather(Period, tfr, -country_code, -name) %>% 
+                    filter(name %in% country)
+  # asfr join tfr
+  asfr_country <- percentASFR %>% gather(Period, asfr, -name, -country_code, -age) %>% 
+                    mutate(x = as.numeric(substr(age,1,2)), asfr=asfr/100) %>% 
+                    inner_join(tfr_country, by=c("country_code","name","Period")) %>% 
+                    mutate(asfr = asfr/5 * tfr) %>% 
+                    select(-age)
+  
+  # lt data must be downloaded manually (maybe too heavy for CRAN)
+  lt_paises_LA <- readxl::read_xlsx("C:/Proyectos/ChildSurv/Data/WPP2019_MORT_F17_3_ABRIDGED_LIFE_TABLE_FEMALE.xlsx", 
+                                    sheet = 1, range = "A17:T77381") %>% 
+                  rename(country_code=5,
+                         x = "Age (x)", 
+                         mx = "Central death rate m(x,n)",
+                         qx = "Probability of dying q(x,n)",
+                         dx = "Number of deaths d(x,n)",
+                         lx = "Number of survivors l(x)",
+                         Lx_n = "Number of person-years lived L(x,n)",
+                         ex = "Expectation of life e(x)")  %>% 
+                  select(country_code,Period,x, mx, qx, dx, lx, Lx_n, ex) %>% 
+                  filter(country_code %in% paises_LA_seleccion)
+  
+  # smooth both components
+  # initial df
+  asfr_paises_LA_s <- data.frame(country_code=0, Period = " ", x = 0, asfr = 0, stringsAsFactors = F)
[…]

### 2020-12-28 R/funs.R

@@ -1,464 +1,0 @@
-library(quadprog)
-library(wpp2019)
-library(tidyverse)
-library(gridExtra)
-source("R/QOSplit.R")
-
-# main functions ----------------------------------------------------------
-
-# age survivor split spline
-lx_fun <- function(data, age){
-  splinefun(data$x, data$lx, method = "monoH.FC")(age)
-}
-
-# function to calculate all the measures
-CS_results <- function(LA, country, period, a){   # LA is the data base
-  
-  # example
-  # country = "Brazil"; period = "1970-1975"; a = 70
-  
-  # filter country
-  country = LA %>% filter(name == country, Period==period)%>% 
-                    mutate(dx = c(-diff(lx),lx[101]),
-                           Lx = ifelse(x==0, Lx_n, # assume uniform deaths within interval
-                                       ifelse(x==100, ex, 
-                                              lx-dx/2)),
-                           Lx = as.numeric(Lx), # just in case
-                           ex = rev(cumsum(rev(Lx)))/lx)
-  
-  # calculate: 
-    # assume that birth is in exact age (not in the middle). That allow us to work with lx, instead of Lx.
-    # consider that i=a is age a-1.
-  CS = sum(unlist(sapply(1:a, function(i) country$asfr[i]*country$lx[a+2-i]/100000)), na.rm = T)
-  Fa = sum(country$asfr[1:a], na.rm = T)
-  CS_prob = CS/Fa
-  k = sum(country$asfr[1:a]*0:(a-1), na.rm = T)/Fa # should be noted as kappa but...
-  CS_aprox = Fa * country$lx[a+1-k]/100000
-  error_aprox = (CS-CS_aprox)/CS*100 # relative error approx
-  MTSL = 0
-  for(i in 1:a){
-    for(j in 1:(a+1-i)){
-      MTSL = sum(MTSL,
-                 country$asfr[i] * country$dx[j]/100000 * sum(country$Lx[j:a])/country$lx[j],
-                 na.rm=T)
-    }
-  }
-  MTSA = 0
-  for(i in 1:a){
-    for(j in 1:(a+1-i)){
-      MTSA = sum(MTSA,
-                 country$asfr[i] * sum(country$Lx[j:a])/country$lx[j],
-                 na.rm=T)
-    }
-  }
-  ITL = MTSL/MTSA * 100
-  m_MAD = 0
[…]

### 2020-12-28 code.R

@@ -1,0 +1,350 @@
+# function and results for paper "Child survival for mothers: mortality change and related measures"
+
+# libraries
+library(wpp2019)
+library(tidyverse)
+library(gridExtra)
+library(quadprog)
+library(ggpubr)
+source("QOSplit.R")
+
+# main functions ----------------------------------------------------------
+
+# age survivor split spline
+lx_fun <- function(data, age){
+  splinefun(data$x, data$lx, method = "monoH.FC")(age)
+}
+
+# function to calculate all the measures
+CS_results <- function(LA, country, period, a){   # LA is the data base
+  
+  # example
+    # country = "Brazil"; period = "1970-1975"; a = 70
+  
+  # filter country
+  country = LA %>% filter(name == country, 
+                          Period==period)%>% 
+                    mutate(dx = c(-diff(lx),lx[101]),
+                           Lx = ifelse(x==0, Lx_n, # assume uniform deaths within interval
+                                       ifelse(x==100, ex, 
+                                              lx-dx/2)),
+                           Lx = as.numeric(Lx), # just in case
+                           ex = rev(cumsum(rev(Lx)))/lx)
+  
+  # calculate: 
+    # assume that birth is in exact age (not in the middle). That allow us to work with lx, instead of Lx.
+    # consider that i=a is age a-1.
+  CS = sum(unlist(sapply(1:a, function(i) country$asfr[i]*country$lx[a+2-i]/100000)), na.rm = T)
+  Fa = sum(country$asfr[1:a], na.rm = T)
+  CS_prob = CS/Fa
+  k = sum(country$asfr[1:a]*0:(a-1), na.rm = T)/Fa # should be noted as kappa but...
+  CS_aprox = Fa * country$lx[a+1-k]/100000
+  error_aprox = (CS-CS_aprox)/CS*100 # relative error approx
+  MTSL = 0
+  for(i in 1:a){
+    for(j in 1:(a+1-i)){
+      MTSL = sum(MTSL,
+                 country$asfr[i] * country$dx[j]/100000 * sum(country$Lx[j:a])/country$lx[j],
+                 na.rm=T)
+    }
+  }
+  MTSA = 0
+  for(i in 1:a){
+    for(j in 1:(a+1-i)){
+      MTSA = sum(MTSA,
+                 country$asfr[i] * sum(country$Lx[j:a])/country$lx[j],
[…]
